### 2021-06-04 analysis/index.Rmd
chore: initial wflow site commit

---
title: "Home"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

Please find all project details in the [About](about.html) page.

### 2021-06-04 analysis/license.Rmd
chore: initial wflow site commit

---
title: "License"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

This work is licensed under CC BY 4.0

*You are free to:*

Share - copy and redistribute the material in any medium or format
Adapt - remix, transform, and build upon the material
for any purpose, even commercially.

*This license is acceptable for Free Cultural Works.*

The licensor cannot revoke these freedoms as long as you follow the license terms.

*Under the following terms:*

Attribution - You must give appropriate credit, provide a link to the license, and indicate if changes were made. You may do so in any reasonable manner, but not in any way that suggests the licensor endorses you or your use.

No additional restrictions - You may not apply legal terms or technological measures that legally restrict others from doing anything the license permits.

*Notices:*

You do not have to comply with the license for elements of the material in the public domain or where your use is permitted by an applicable exception or limitation.
No warranties are given. The license may not give you all of the permissions necessary for your intended use. For example, other rights such as publicity, privacy, or moral rights may limit how you use the material.

### 2021-06-11 scripts/R.sh
feat: add R run script

@@ -1,0 +1,32 @@
+#!/bin/bash
+
+# The cli package has an internal function cli:::win10_build
+# that uses utils::sessionInfo()$running, which returns osVersion.
+# osVersion is NULL in the container and `tar_make` will throw an error
+# because cli::win10_build fails because grepl("^Win...", osVersion) won't work.
+# This overwrites win10_build and just returns 0L
+# https://github.com/r-lib/cli/issues/273
+# We could also use
+# SINGULARITYENV_R_CLI_NUM_COLORS=1 \
+WORKAROUND='assignInNamespace("win10_build", function()0L, ns = "cli")'
+
+REXPR=${1:-"${WORKAROUND}; targets::tar_make()"}
+PREPEND_PATH=""
+BIND_PATH="/var/run"
+
+## on an HPC
+if ! [[ ${HOSTNAME} =~ (x1|t400) ]] ; then
+    module load singularity slurm
+    PREPEND_PATH=$(IFS=:; for i in ${PATH}; do if [[ $i =~ (slurm|munge|singularity) ]] ; then echo -n "${i}:"; fi; done;)
+    PREPEND_PATH=${PREPEND_PATH%:}
+    BIND_PATH="/etc/slurm/,/usr/lib64/libmunge.so.2,${BIND_PATH},$(echo ${PREPEND_PATH} | sed 's#:#\n#g; s#\(bin\|lib\)\/\?##g' | uniq | sed ':a;N;$!ba;s#\n#,#g')"
+fi
+
+# cli:::win10_build fails on guix generated containers, see
+#
+SINGULARITYENV_LOCPATH=/lib/locale/2.31/ \
+    SINGULARITYENV_TZ="UTC" \
+    SINGULARITYENV_PREPEND_PATH="${PREPEND_PATH}" \
+    SINGULARITY_BIND="${BIND_PATH}" \
+    singularity run container/ampel-leipzig-meld.sif \
+    --vanilla --silent -e "${REXPR}"

### 2021-06-11 analysis/histograms.Rmd
feat: add basic targets workflow

---
title: "Histograms"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r histograms_setup}
library("targets")
tar_load(raw_data)

.hist <- function(
    ..., xlab = deparse(substitute(x)), main = paste("Histogram of", xlab),
    col = palette.colors(2L)[2]
)hist(..., main = main, xlab = xlab, col = col)
```

# Age

```{r histograms_age}
.hist(raw_data$Age, xlab = "Age")
```

# Laboratory Measurements

```{r histograms_labs, results = "asis", message = FALSE, warning = FALSE}
labv <- sort(grep("_[SECQ]$", colnames(raw_data), value = TRUE))

for (v in labv) {
    cat("\n\n##", v , "\n\n")
    .hist(raw_data[[v]], main = paste0("Histogram of ", v), xlab = v)
}
```

### 2021-06-11 analysis/license.Rmd
feat: add basic targets workflow

@@
 
-This work is licensed under CC BY 4.0
+**Code** is released under the [GNU GPLv3 license][gpl].
 
-*You are free to:*
+**Text** is released under the
+[Creative Commons Attribution 4.0 International License][cc].
 
-Share - copy and redistribute the material in any medium or format
-Adapt - remix, transform, and build upon the material
-for any purpose, even commercially.
-
-*This license is acceptable for Free Cultural Works.*
-
-The licensor cannot revoke these freedoms as long as you follow the license terms.
-
-*Under the following terms:*
-
-Attribution - You must give appropriate credit, provide a link to the license, and indicate if changes were made.
-You may do so in any reasonable manner, but not in any way that suggests the licensor endorses you or your use.
-
-No additional restrictions - You may not apply legal terms or technological measures that legally restrict others from doing anything the license permits.
-
-*Notices:*
-
-You do not have to comply with the license for elements of the material in the public domain or where your use is permitted by an applicable exception or limitation.
-No warranties are given.
-The license may not give you all of the permissions necessary for your intended use.
-For example, other rights such as publicity, privacy, or moral rights may limit how you use the material.
+[gpl]: https://choosealicense.com/licenses/gpl-3.0/
+[cc]: https://creativecommons.org/licenses/by/4.0/

### 2021-06-11 _targets.R

@@ -1,0 +1,17 @@
+library("targets")
+library("tarchetypes")
+
+tar_option_set(packages = c(
+    "ameld",
+    "zlog"
+))
+
+source("code/01-data.R")
+source("code/03-reports.R")
+source("code/99-site.R")
+
+list(
+   tg_data,
+   tg_reports,
+   tg_site
+)

### 2021-06-11 code/01-data.R

@@ -1,0 +1,17 @@
+tg_data <- list(
+    tar_target(raw_data, {
+        data(eldd, package = "ameld")
+        eldd
+    }),
+    tar_target(ref_data, {
+        ## prepare reference data for zlog normalisation
+        data(eldr, package = "ameld")
+        r <- eldr[c("Code", "AgeDays", "Sex", "LowerLimit", "UpperLimit")]
+        names(r) <- c("param", "age", "sex", "lower", "upper")
+        r$age <- r$age / 365.25
+
+        ## set missing lower limits to 0.15 of the upper according to
+        ## Haeckel et al. 2015
+        set_missing_limits(r)
+    }, packages = "zlog")
+)

### 2021-06-11 code/03-reports.R

@@ -1,0 +1,16 @@
+tg_reports <- list(
+    tar_target(
+        report_histograms,
+        command = {
+            list(site_conf)
+            !!tar_knitr_deps_expr(file.path("analysis", "histograms.Rmd"))
+            workflowr::wflow_build(
+                file.path("analysis", "histograms.Rmd"),
+                message = "docs: rebuild histograms.html",
+                view = FALSE
+           )
+           file.path("analysis", "histograms.Rmd")
+        },
+        format = "file"
+    )
+)

### 2021-06-11 code/99-site.R

@@ -1,0 +1,50 @@
+## adopted from
+## https://github.com/jdblischak/workflowr/issues/238#issuecomment-782024069
+tg_site <- list(
+    tar_file(site_conf, file.path("analysis", "_site.yml")),
+    tar_target(
+        site_about,
+        command = {
+            list(site_conf)
+            !!tar_knitr_deps_expr(file.path("analysis", "about.Rmd"))
+            # Build
+            workflowr::wflow_build(
+                file.path("analysis", "about.Rmd"),
+                message = "docs: rebuild about.html",
+                view = FALSE
+            )
+            # Return file name
+            file.path("analysis", "about.Rmd")
+        },
+        format = "file"
+    ),
+    tar_target(
+        site_index,
+        command = {
+            list(site_conf)
+            !!tar_knitr_deps_expr(file.path("analysis", "index.Rmd"))
+            workflowr::wflow_build(
+                file.path("analysis", "index.Rmd"),
+                message = "docs: rebuild index.html",
+                view = FALSE
+            )
+            file.path("analysis", "index.Rmd")
+        },
+        format = "file"
+    ),
+    tar_target(
+        site_license,
+        command = {
+            list(site_conf)
+            !!tar_knitr_deps_expr(file.path("analysis", "license.Rmd"))
+            workflowr::wflow_build(
+                file.path("analysis", "license.Rmd"),
+                message = "docs: rebuild license.html",
+                view = FALSE
+            )
+
+            file.path("analysis", "license.Rmd")
+        },
+        format = "file"
+    )
+)

### 2021-06-11 code/03-reports.R
fix: remove message argument

@@ -6,7 +6,6 @@ tg_reports <- list(
             !!tar_knitr_deps_expr(file.path("analysis", "histograms.Rmd"))
             workflowr::wflow_build(
                 file.path("analysis", "histograms.Rmd"),
-                message = "docs: rebuild histograms.html",
                 view = FALSE
            )
            file.path("analysis", "histograms.Rmd")

### 2021-06-11 code/99-site.R

@@ -10,7 +10,6 @@ tg_site <- list(
             # Build
             workflowr::wflow_build(
                 file.path("analysis", "about.Rmd"),
-                message = "docs: rebuild about.html",
                 view = FALSE
             )
             # Return file name
@@ -25,7 +24,6 @@ tg_site <- list(
             !!tar_knitr_deps_expr(file.path("analysis", "index.Rmd"))
             workflowr::wflow_build(
                 file.path("analysis", "index.Rmd"),
-                message = "docs: rebuild index.html",
                 view = FALSE
             )
             file.path("analysis", "index.Rmd")
@@ -39,7 +37,6 @@ tg_site <- list(
             !!tar_knitr_deps_expr(file.path("analysis", "license.Rmd"))
             workflowr::wflow_build(
                 file.path("analysis", "license.Rmd"),
-                message = "docs: rebuild license.html",
                 view = FALSE
             )
 

### 2021-06-11 analysis/histograms.Rmd
feat: use imputed data

@@
 library("targets")
-tar_load(raw_data)
+tar_load(imp_data)
 
@@
     ..., xlab = deparse(substitute(x)), main = paste("Histogram of", xlab),
-    col = palette.colors(2L)[2]
+    col = palette.colors(2L)[2L]
 )hist(..., main = main, xlab = xlab, col = col)
@@
 ```{r histograms_age}
-.hist(raw_data$Age, xlab = "Age")
+.hist(imp_data$Age, xlab = "Age")
 ```
 
-# Laboratory Measurements
+# Raw Data (imputed)
+
+## Laboratory Measurements
 
 ```{r histograms_labs, results = "asis", message = FALSE, warning = FALSE}
-labv <- sort(grep("_[SECQ]$", colnames(raw_data), value = TRUE))
+labv <- sort(grep("_[SECQ]$", colnames(imp_data), value = TRUE))
 
@@
     cat("\n\n##", v , "\n\n")
-    .hist(raw_data[[v]], main = paste0("Histogram of ", v), xlab = v)
+    .hist(imp_data[[v]], main = paste0("Histogram of ", v), xlab = v)
 }

### 2021-06-11 code/01-data.R

@@ -13,5 +13,19 @@ tg_data <- list(
         ## set missing lower limits to 0.15 of the upper according to
         ## Haeckel et al. 2015
         set_missing_limits(r)
-    }, packages = "zlog")
+    }, packages = "zlog"),
+    tar_target(imp_data, {
+        withCallingHandlers(
+            impute_df(raw_data, ref_data, method = "logmean"),
+            warning = function(w) {
+                if(w$message != paste0(
+                    "No reference for column(s): DaysAtRisk, Deceased, LTx, ",
+                    "Cirrhosis, ALF, Ethyltoxic, HBV, HCV, AIH, PBC, PSC, ",
+                    "NASH, Cryptogenic, Dialysis, GIB, HCC, SBP"))
+                       stop(w$message)
+                else
+                    invokeRestart("muffleWarning")
+            }
+        )
+    })
 )

### 2021-06-11 analysis/missing.Rmd
feat: add missing data overview

---
title: "Missing Data"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r missing_setup}
library("targets")
tar_load(raw_data)

.dotplot <- function(x, xlim = c(0, length(x)),
                     col = palette.colors(2L)[2L], pch = 19L) {
    old.par <- par(no.readonly = TRUE)
    on.exit(par(old.par))

    n <- length(x)

    plot.new()
    w <- max(strwidth(names(x), "inch"), na.rm = TRUE) + 1/16
    mai <- par("mai")
    if (mai[2L] < w)
        mai[2L] <- mai[4L] + w # taken from dotchart
    par(mai = mai)

    plot.window(xlim = xlim, ylim = c(0L, n + 1L))
    title(main = "Missing Data", adj = 0L)
    title(xlab = "Frequency", adj = 1L)
    y <- seq_len(n)
    mtext(
        names(x), at = y, adj = 0L, side = 2L, las = 2L,
        line = (w + 0.1) / par("csi")
    )
    abline(h = y, col = "#808080", lty = "dotted", lwd = 1L)
    points(x, y, col = col, pch = pch)
    axis(1, lwd.ticks = 0L, col = "#808080")
}

.image <- function(x, col = c("white", palette.colors(2L)[2L])) {
    old.par <- par(no.readonly = TRUE)
    on.exit(par(old.par))

    nr <- nrow(x)
    nc <- ncol(x)
    nms <- colnames(x)

    plot.window(xlim = 0.5 + c(0, nc), ylim = 0.5 + c(0, nr))
    title(main = "Missing Data", adj = 0L)
    title(xlab = "Rows", adj = 1L)
    w <- max(strwidth(nms, "inch"), na.rm = TRUE) + 1/16
    mai <- par("mai")
    if (mai[2L] < w)
        mai[2L] <- mai[4L] + w # taken from dotchart
    par(mai = mai)

    y <- seq_len(nc)
    image(1L:nr, y, x[, rev(y)], axes = FALSE, xlab = "", ylab = "", col = col)
    mtext(
        rev(nms), at = y, adj = 0L, side = 2L, las = 2L,
        line = (w + 0.1) / par("csi")
    )
    abline(h = y, col = "#808080", lty = "dotted", lwd = 1L)
}
```

# Missing Data

```{r missing_table}
m <- raw_data
m$Sex <- as.numeric(m$Sex)
m <- as.matrix(m)
tbl <- colSums(is.na(m))
na <- sort(tbl[tbl > 0L], decreasing = TRUE)
knitr::kable(na, col.names = "# Missing Data")

.dotplot(sort(na), xlim = c(0L, nrow(m)))
```

```{r missing_image}
mna <- is.na(m)
mode(mna) <- "numeric"
.image(mna)
```

### 2021-06-11 analysis/missing.Rmd
feat: plot missing values

@@
     toc: true
+    fig.width: 10
+    fig.height: 10
 editor_options:
@@
 
-    plot.new()
+    mai <- par("mai")
     w <- max(strwidth(names(x), "inch"), na.rm = TRUE) + 1/16
-    mai <- par("mai")
     if (mai[2L] < w)
@@
         names(x), at = y, adj = 0L, side = 2L, las = 2L,
-        line = (w + 0.1) / par("csi")
+        line = (w + 0.1) / par("csi"), cex = 0.8
     )
@@
     points(x, y, col = col, pch = pch)
-    axis(1, lwd.ticks = 0L, col = "#808080")
+    axis(1L, lwd.ticks = 0L, col = "#808080")
 }
@@
 
-    plot.window(xlim = 0.5 + c(0, nc), ylim = 0.5 + c(0, nr))
-    title(main = "Missing Data", adj = 0L)
-    title(xlab = "Rows", adj = 1L)
+    mai <- par("mai")
     w <- max(strwidth(nms, "inch"), na.rm = TRUE) + 1/16
-    mai <- par("mai")
     if (mai[2L] < w)
@@
     y <- seq_len(nc)
-    image(1L:nr, y, x[, rev(y)], axes = FALSE, xlab = "", ylab = "", col = col)
+    image(
+        1L:nr, y, x[, rev(y)], col = col,
+        xlim = 0.5 + c(0L, nr), ylim = 0.5 + c(0L, nc),
+        axes = FALSE, xlab = "", ylab = ""
+    )
+    title(main = "Missing Data", adj = 0L)
+    title(xlab = "Rows", adj = 1L)
     mtext(
         rev(nms), at = y, adj = 0L, side = 2L, las = 2L,
-        line = (w + 0.1) / par("csi")
+        line = (w + 0.1) / par("csi"), cex = 0.8
     )
@@
 m <- as.matrix(m)
-tbl <- colSums(is.na(m))
+mna <- is.na(m)
+mode(mna) <- "numeric"
+tbl <- colSums(mna)
 na <- sort(tbl[tbl > 0L], decreasing = TRUE)
-knitr::kable(na, col.names = "# Missing Data")
+knitr::kable(
+    cbind(na, round(colMeans(mna[, names(na)]) * 100, 2)),
+    col.names = c("# Missing Data", "% of Missing Data")
+)
 
@@
 
-```{r missing_image}
-mna <- is.na(m)
-mode(mna) <- "numeric"
+```{r missing_image, fig.height = 10, fig.width = 12}
 .image(mna)

### 2021-06-11 analysis/corrplot.Rmd
feat: add correlation plot

---
title: "Corrplot"
output:
  workflowr::wflow_html:
    toc: true
    fig.width: 12
    fig.height: 12
editor_options:
  chunk_output_type: console
---

```{r corrplot_setup}
library("corrplot")
library("targets")
tar_load(raw_data)
```

# Correlation

```{r corrplot}
labv <- grep("_[SECQ]$", colnames(raw_data), value = TRUE)
cr <- cor(raw_data[labv], use = "pairwise.complete.obs")
corrplot.mixed(
    cr,
    tl.pos = "lt", tl.col = "black", tl.cex = 0.75,
    number.cex = 0.25, number.font = 1L,
    order = "AOE"
)
```

### 2021-06-11 analysis/corrplot.Rmd
fix: use imputed data for correlation

@@
 library("targets")
-tar_load(raw_data)
+tar_load(imp_data)
 ```
@@
 ```{r corrplot}
-labv <- grep("_[SECQ]$", colnames(raw_data), value = TRUE)
-cr <- cor(raw_data[labv], use = "pairwise.complete.obs")
+labv <- grep("_[SECQ]$", colnames(imp_data), value = TRUE)
+cr <- cor(imp_data[labv], use = "pairwise.complete.obs")
 corrplot.mixed(

### 2021-06-11 analysis/missing.Rmd
fix: wrong plot call

@@
 
-    plot.window(xlim = xlim, ylim = c(0L, n + 1L))
+    plot(NA, xlim = xlim, ylim = c(0L, n + 1L), axes = FALSE)
     title(main = "Missing Data", adj = 0L)
@@
         rev(nms), at = y, adj = 0L, side = 2L, las = 2L,
-        line = (w + 0.1) / par("csi"), cex = 0.8
+        line = (w + 0.1) / par("csi"), cex = 0.75
     )

### 2021-06-11 code/03-reports.R
fix: add targets for missing/corrplot

@@ -11,5 +11,31 @@ tg_reports <- list(
            file.path("analysis", "histograms.Rmd")
         },
         format = "file"
+    ),
+    tar_target(
+        report_missing,
+        command = {
+            list(site_conf)
+            !!tar_knitr_deps_expr(file.path("analysis", "missing.Rmd"))
+            workflowr::wflow_build(
+                file.path("analysis", "missing.Rmd"),
+                view = FALSE
+           )
+           file.path("analysis", "missing.Rmd")
+        },
+        format = "file"
+    ),
+    tar_target(
+        report_corrplot,
+        command = {
+            list(site_conf)
+            !!tar_knitr_deps_expr(file.path("analysis", "corrplot.Rmd"))
+            workflowr::wflow_build(
+                file.path("analysis", "corrplot.Rmd"),
+                view = FALSE
+           )
+           file.path("analysis", "corrplot.Rmd")
+        },
+        format = "file"
     )
 )

### 2021-06-11 analysis/histograms.Rmd
feat: add zlog_data

@@
 library("targets")
+library("viridisLite")
 tar_load(imp_data)
+tar_load(zlog_data)
 
@@
 )hist(..., main = main, xlab = xlab, col = col)
+
+.qqnorm <- function(y, ..., yname = deparse(substitute(y)),
+    ylab = paste0("Sample Quantiles (", yname, ")"),
+    col = palette.colors(2L)[2L]) {
+    qqnorm(y, ..., ylab = ylab, col = col, pch = 19L)
+    qqline(y, ..., col = col)
+}
+
+.bp <- function(x, formula, col = cividis(ncol(x)), main = NULL,
+                outline = FALSE) {
+    boxplot(
+        formula, data = x, col = gsub("..$", "40", col),
+        pch = NA, outline = outline,
+        main = main
+    )
+    beeswarm::beeswarm(formula, data = x, col = col, add = TRUE, pch = 19L)
+}
 ```
@@
     .hist(imp_data[[v]], main = paste0("Histogram of ", v), xlab = v)
+    .qqnorm(imp_data[[v]], yname = v)
 }
 ```
+
+# Zlog Data (imputed)
+
+## Laboratory Measurements
+
+```{r histograms_labs_zlog, results = "asis", message = FALSE, warning = FALSE}
+for (v in labv) {
+    cat("\n\n##", v , "\n\n")
+    .hist(zlog_data[[v]], main = paste0("Histogram of ", v), xlab = v)
+    .qqnorm(zlog_data[[v]], yname = v)
+}
+```
+
+# Raw vs Zlog Data
+
+## Compare Sexes
+
+```{r histograms_rvsz_sex, results = "asis"}
+old.par <- par(no.readonly = TRUE)
+par(mfrow = c(1L, 2L))
+log_data <- imp_data
+log_data[labv] <- lapply(log_data[labv], log)
+col <- cividis(2L)
+for (v in labv) {
+    cat('\n\n### ', v, '\n\n')
+    fm <- as.formula(paste0(v, " ~ Sex"))
+    .bp(log_data, fm, main = "log()")
+    .bp(zlog_data, fm, main = "zlog()")
+}
+par(old.par)
+```
+
+## Compare Status
+
+```{r histograms_rvsz_status, results = "asis"}
+old.par <- par(no.readonly = TRUE)
+par(mfrow = c(1L, 2L))
+for (v in labv) {
+    cat('\n\n### ', v, '\n\n')
+    fm <- as.formula(paste0(v, " ~ Deceased"))
+    .bp(log_data, fm, main = "log()")
+    .bp(zlog_data, fm, main = "zlog()")
+}
+par(old.par)
+```
+
+## Compare Sexes : Status
+
+```{r histograms_rvsz_sex_status, results = "asis"}
+old.par <- par(no.readonly = TRUE)
+par(mfrow = c(1L, 2L))
+col <- cividis(4L)
+for (v in labv) {
+    cat('\n\n### ', v, '\n\n')
+    fm <- as.formula(paste0(v, " ~ Sex + Deceased"))
+    .bp(log_data, fm, main = "log()")
+    .bp(zlog_data, fm, main = "zlog()")
+}
+par(old.par)
+```

### 2021-06-11 code/01-data.R

@@ -27,5 +27,12 @@ tg_data <- list(
                     invokeRestart("muffleWarning")
             }
         )
-    })
+    }),
+    tar_target(zlog_data, {
+        zlog_data <- imp_data
+        iconvert <- colnames(zlog_data) %in%
+            c("Age", "Sex", grep("_[SECQ]$", colnames(imp_data), value = TRUE))
+        zlog_data[iconvert] <- zlog_df(imp_data[iconvert], ref_data)
+        zlog_data
+    }, packages = "zlog")
 )

### 2021-06-11 analysis/histograms.Rmd
fix: remove useless headings
fix: size and color of the beeswarm plots

--- 2021-06-11 analysis/histograms.Rmd
@@
 
-```{r histograms_setup}
+```{r histograms_setup, message = FALSE}
 library("targets")
+library("beeswarm")
+library("viridisLite")
 tar_load(imp_data)
+tar_load(zlog_data)
 
@@
 )hist(..., main = main, xlab = xlab, col = col)
+
+.qqnorm <- function(y, ..., yname = deparse(substitute(y)),
+    ylab = paste0("Sample Quantiles (", yname, ")"),
+    col = palette.colors(2L)[2L]) {
+    qqnorm(y, ..., ylab = ylab, col = col, pch = 19L)
+    qqline(y, ..., col = col)
+}
+
+.bp <- function(x, formula, col = cividis(ncol(x)), main = NULL,
+                outline = FALSE) {
+    boxplot(
+        formula, data = x, col = gsub("..$", "40", col),
+        pch = NA, outline = outline,
+        main = main
+    )
+    beeswarm(formula, data = x, col = col, add = TRUE, pch = 19L)
+}
 ```
@@
 
-## Laboratory Measurements
-
 ```{r histograms_labs, results = "asis", message = FALSE, warning = FALSE}
@@
     .hist(imp_data[[v]], main = paste0("Histogram of ", v), xlab = v)
+    .qqnorm(imp_data[[v]], yname = v)
 }
 ```
+
+# Zlog Data (imputed)
+
+```{r histograms_labs_zlog, results = "asis", message = FALSE, warning = FALSE}
+for (v in labv) {
+    cat("\n\n##", v , "\n\n")
+    .hist(zlog_data[[v]], main = paste0("Histogram of ", v), xlab = v)
+    .qqnorm(zlog_data[[v]], yname = v)
+}
+```
+
+# Raw vs Zlog Data
+
+## Compare Sexes
+
+```{r histograms_rvsz_sex, results = "asis", fig.width = 12, fig.height = 12}
+old.par <- par(no.readonly = TRUE)
+par(mfrow = c(1L, 2L))
+log_data <- imp_data
+log_data[labv] <- lapply(log_data[labv], log)
+col <- cividis(2L)
+for (v in labv) {
+    cat('\n\n### ', v, '\n\n')
+    fm <- as.formula(paste0(v, " ~ Sex"))
+    .bp(log_data, fm, col = col, main = "log()")
+    .bp(zlog_data, fm, col = col, main = "zlog()")
+}
+par(old.par)
+```
+
+## Compare Status
+
+```{r histograms_rvsz_status, results = "asis", fig.width = 12, fig.height = 12}
+old.par <- par(no.readonly = TRUE)
+par(mfrow = c(1L, 2L))
+for (v in labv) {
+    cat('\n\n### ', v, '\n\n')
+    fm <- as.formula(paste0(v, " ~ Deceased"))
+    .bp(log_data, fm, col = col, main = "log()")
+    .bp(zlog_data, fm, col = col, main = "zlog()")
+}
+par(old.par)
+```
+
+## Compare Sexes : Status
+
+```{r histograms_rvsz_sex_status, results = "asis", fig.width = 12, fig.height = 12}
+old.par <- par(no.readonly = TRUE)
+par(mfrow = c(1L, 2L))
+col <- cividis(4L)
+for (v in labv) {
+    cat('\n\n### ', v, '\n\n')
+    fm <- as.formula(paste0(v, " ~ Sex + Deceased"))
+    .bp(log_data, fm, col = col, main = "log()")
+    .bp(zlog_data, fm, col = col, main = "zlog()")
+}
+par(old.par)
+```

### 2021-06-11 analysis/boxplots.Rmd
refactor: move boxplots in separate report

---
title: "Boxplots"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r boxplots_setup, message = FALSE}
library("targets")
library("beeswarm")
library("viridisLite")
tar_load(imp_data)
tar_load(zlog_data)

.bp <- function(x, formula, col = cividis(ncol(x)), main = NULL,
                outline = FALSE) {
    boxplot(
        formula, data = x, col = gsub("..$", "40", col),
        pch = NA, outline = outline,
        main = main
    )
    beeswarm(formula, data = x, col = col, add = TRUE, pch = 19L)
}
```

# Raw vs Zlog Data

## Compare Sexes

```{r boxplots_rvsz_sex, results = "asis", fig.width = 12, fig.height = 12}
old.par <- par(no.readonly = TRUE)
par(mfrow = c(1L, 2L))
log_data <- imp_data
labv <- sort(grep("_[SECQ]$", colnames(imp_data), value = TRUE))
log_data[labv] <- lapply(log_data[labv], log)
col <- cividis(2L)
for (v in labv) {
    cat('\n\n### ', v, '\n\n')
    fm <- as.formula(paste0(v, " ~ Sex"))
    .bp(log_data, fm, col = col, main = "log()")
    .bp(zlog_data, fm, col = col, main = "zlog()")
}
par(old.par)
```

## Compare Status

```{r boxplots_rvsz_status, results = "asis", fig.width = 12, fig.height = 12}
old.par <- par(no.readonly = TRUE)
par(mfrow = c(1L, 2L))
for (v in labv) {
    cat('\n\n### ', v, '\n\n')
    fm <- as.formula(paste0(v, " ~ Deceased"))
    .bp(log_data, fm, col = col, main = "log()")
    .bp(zlog_data, fm, col = col, main = "zlog()")
}
par(old.par)
```

## Compare Sexes : Status

```{r boxplots_rvsz_sex_status, results = "asis", fig.width = 12, fig.height = 12}
old.par <- par(no.readonly = TRUE)
par(mfrow = c(1L, 2L))
col <- cividis(4L)
for (v in labv) {
    cat('\n\n### ', v, '\n\n')
    fm <- as.formula(paste0(v, " ~ Sex + Deceased"))
    .bp(log_data, fm, col = col, main = "log()")
    .bp(zlog_data, fm, col = col, main = "zlog()")
}
par(old.par)
```

### 2021-06-11 analysis/histograms.Rmd
refactor: move boxplots in separate report

--- 2021-06-11 analysis/histograms.Rmd
@@
 
-```{r histograms_setup}
+```{r histograms_setup, message = FALSE}
 library("targets")
 tar_load(imp_data)
+tar_load(zlog_data)
 
@@
 )hist(..., main = main, xlab = xlab, col = col)
+
+.qqnorm <- function(y, ..., yname = deparse(substitute(y)),
+    main = paste0("Normal Q-Q Plot (", yname, ")"),
+    ylab = paste0("Sample Quantiles (", yname, ")"),
+    col = palette.colors(2L)[2L]) {
+    qqnorm(y, ..., main = main, ylab = ylab, col = col, pch = 19L)
+    qqline(y, ..., col = col)
+}
 ```
@@
     .hist(imp_data[[v]], main = paste0("Histogram of ", v), xlab = v)
+    .qqnorm(imp_data[[v]], yname = v)
 }
 ```
+
+# Zlog Data (imputed)
+
+```{r histograms_labs_zlog, results = "asis", message = FALSE, warning = FALSE}
+for (v in labv) {
+    cat("\n\n##", v , "\n\n")
+    .hist(zlog_data[[v]], main = paste0("Histogram of ", v), xlab = v)
+    .qqnorm(zlog_data[[v]], yname = v)
+}
+```

### 2021-06-11 code/03-reports.R

@@ -13,6 +13,19 @@ tg_reports <- list(
         format = "file"
     ),
     tar_target(
+        report_boxplots,
+        command = {
+            list(site_conf)
+            !!tar_knitr_deps_expr(file.path("analysis", "boxplots.Rmd"))
+            workflowr::wflow_build(
+                file.path("analysis", "boxplots.Rmd"),
+                view = FALSE
+           )
+           file.path("analysis", "boxplots.Rmd")
+        },
+        format = "file"
+    ),
+    tar_target(
         report_missing,
         command = {
             list(site_conf)

### 2021-06-11 analysis/tables.Rmd
feat: add skimr tables

---
title: "Tables"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: console
---

```{r histograms_setup, message = FALSE}
library("targets")
library("skimr")
tar_load(raw_data)
tar_load(imp_data)
tar_load(zlog_data)
```

# Raw Data

```{r tables_skim_raw}
skim(raw_data)
```

# Raw Data (imputed)

```{r tables_skim_imp}
skim(imp_data)
```

# Zlog Data

```{r tables_skim_zlog}
skim(zlog_data)
```

### 2021-06-11 code/03-reports.R

@@ -1,5 +1,18 @@
 tg_reports <- list(
     tar_target(
+        report_tables,
+        command = {
+            list(site_conf)
+            !!tar_knitr_deps_expr(file.path("analysis", "tables.Rmd"))
+            workflowr::wflow_build(
+                file.path("analysis", "tables.Rmd"),
+                view = FALSE
+           )
+           file.path("analysis", "tables.Rmd")
+        },
+        format = "file"
+    ),
+    tar_target(
         report_histograms,
         command = {
             list(site_conf)

### 2021-06-11 analysis/linecharts.Rmd
feat: add linecharts report

---
title: "Linecharts"
output:
  workflowr::wflow_html:
    toc: true
    fig.width: 12
    fig.height: 8
editor_options:
  chunk_output_type: console
---

```{r linecharts_setup, message = FALSE}
library("targets")
tar_load(zlog_data)
```

# Compare Mean Zlog Values vs Status

```{r linecharts_zlogmeanplot}
labv <- sort(grep("_[SECQ]$", colnames(zlog_data), value = TRUE))

## divide data.frame by dead/alive
s <- split(zlog_data[labv], zlog_data$Deceased)
names(s) <- c("survived", "dead")

## calculate mean standardized lab values
s <- lapply(s, colMeans, na.rm = TRUE)
o <- order(s$dead)

## comparison plot
col <- palette.colors(2L)

## keep old graphic parameters and restore them afterwards
old.par <- par(no.readonly = TRUE)

par(mar = c(7.1, 4.1, 4.1, 2.1))

plot(
    s$dead[o], type = "b", pch = 20, lwd = 2, col = col[1],
    axes = FALSE, ann = FALSE
)
lines(s$survived[o], type = "b", pch = 20, lwd = 2, col = col[2])
legend(
    "bottomright",
    legend = c("dead", "survived"),
    col = col, lwd = 2, pch = 20, bty = "n"
)
title(
    main = "Mortality Status vs Mean Standardized Laboratory Values", adj = 0
)
title(xlab = "Laboratory Measurements", adj = 1L, line = 5)
title(ylab = "Mean Standardized Values", adj = 1L)
r <- range(unlist(s))
axis(
    2, at = seq(from = floor(r[1]), to = ceiling(r[2])),
    lwd.ticks = 0, col = "#808080"
)
axis(
    1, at = seq_along(o), labels = names(s$dead[o]), las = 2,
    lwd.ticks = 0L, col = "#808080"
)
par(old.par)
```

### 2021-06-11 code/03-reports.R

@@ -52,6 +52,19 @@ tg_reports <- list(
         format = "file"
     ),
     tar_target(
+        report_linecharts,
+        command = {
+            list(site_conf)
+            !!tar_knitr_deps_expr(file.path("analysis", "linecharts.Rmd"))
+            workflowr::wflow_build(
+                file.path("analysis", "linecharts.Rmd"),
+                view = FALSE
+           )
+           file.path("analysis", "linecharts.Rmd")
+        },
+        format = "file"
+    ),
+    tar_target(
         report_corrplot,
         command = {
             list(site_conf)

### 2021-07-02 code/03-reports.R
feat: add dotcharts

@@ -52,6 +52,19 @@ tg_reports <- list(
         format = "file"
     ),
     tar_target(
+        report_dotcharts,
+        command = {
+            list(site_conf)
+            !!tar_knitr_deps_expr(file.path("analysis", "dotcharts.Rmd"))
+            workflowr::wflow_build(
+                file.path("analysis", "dotcharts.Rmd"),
+                view = FALSE
+           )
+           file.path("analysis", "dotcharts.Rmd")
+        },
+        format = "file"
+    ),
+    tar_target(
         report_linecharts,
         command = {
             list(site_conf)

### 2021-07-02 analysis/index.Rmd
refactor: remove about page

@@
 
-Please find all project details in the [About](about.html) page.
+# Acknowledgment
+
+This work is part of the [AMPEL](https://ampel.care/en/)
+(Analysis and Reporting System for the Improvement of Patient Safety through
+Real-Time Integration of Laboratory Findings) project.
+
+This measure is co-funded with tax revenues based on the budget adopted by
+the members of the Saxon State Parliament.

### 2021-07-02 code/99-site.R

@@ -3,21 +3,6 @@
 tg_site <- list(
     tar_file(site_conf, file.path("analysis", "_site.yml")),
     tar_target(
-        site_about,
-        command = {
-            list(site_conf)
-            !!tar_knitr_deps_expr(file.path("analysis", "about.Rmd"))
-            # Build
-            workflowr::wflow_build(
-                file.path("analysis", "about.Rmd"),
-                view = FALSE
-            )
-            # Return file name
-            file.path("analysis", "about.Rmd")
-        },
-        format = "file"
-    ),
-    tar_target(
         site_index,
         command = {
             list(site_conf)

### 2021-07-02 scripts/R.sh
feat: add pipeline page

@@ -10,7 +10,7 @@
 # SINGULARITYENV_R_CLI_NUM_COLORS=1 \
 WORKAROUND='assignInNamespace("win10_build", function()0L, ns = "cli")'
 
-REXPR=${1:-"${WORKAROUND}; targets::tar_make()"}
+REXPR=${1:-"${WORKAROUND}; targets::tar_make(); workflowr::wflow_build(rprojroot::find_rstudio_root_file('analysis', 'pipeline.Rmd'), view = FALSE)"}
 PREPEND_PATH=""
 BIND_PATH="/var/run"
 

### 2021-07-02 scripts/R.sh
refactor: remove useless comment and set timestamp reporter

@@ -22,9 +22,8 @@ if ! [[ ${HOSTNAME} =~ (x1|t400) ]] ; then
     BIND_PATH="/etc/slurm/,/usr/lib64/libmunge.so.2,${BIND_PATH},$(echo ${PREPEND_PATH} | sed 's#:#\n#g; s#\(bin\|lib\)\/\?##g' | uniq | sed ':a;N;$!ba;s#\n#,#g')"
 fi
 
-# cli:::win10_build fails on guix generated containers, see
-#
 SINGULARITYENV_LOCPATH=/lib/locale/2.31/ \
+    SINGULARITYENV_TAR_MAKE_REPORTER="timestamp" \
     SINGULARITYENV_TZ="UTC" \
     SINGULARITYENV_PREPEND_PATH="${PREPEND_PATH}" \
     SINGULARITY_BIND="${BIND_PATH}" \

### 2021-07-02 analysis/index.Rmd
feat: add first citations

--- 2021-06-04 analysis/index.Rmd
@@
   chunk_output_type: console
+bibliography:
+  - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'bibliography.bib')`"
+  - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'rpackages', 'home.bib')`"
 ---
 
-Please find all project details in the [About](about.html) page.
+```{r, include=FALSE}
+knitr::write_bib(
+    c("targets", "workflowr"),
+    file = rprojroot::find_rstudio_root_file(
+        "analysis", "bibliography", "rpackages", "home.bib"
+    )
+)
+```
+
+## Acknowledgment
+
+### AMPEL
+
+This work is part of the [AMPEL](https://ampel.care/en/)
+(Analysis and Reporting System for the Improvement of Patient Safety through
+Real-Time Integration of Laboratory Findings) project [@eckelt2020].
+
+This measure is co-funded with tax revenues based on the budget adopted by
+the members of the Saxon State Parliament.
+
+### R Packages
+
+This site is generated by combining the
+[workflowr](https://jdblischak.github.io/workflowr/) [@workflowr2019] and the
+[targets](https://docs.ropensci.org/targets/) [@targets2021] package.
+
+A visual network of the pipeline is depicted [here](pipeline.html).
+
+## References

### 2021-07-02 analysis/missing.Rmd
refactor: use ameld::plot_dots

@@
 tar_load(raw_data)
-
-.dotplot <- function(x, xlim = c(0, length(x)),
-                     col = palette.colors(2L)[2L], pch = 19L) {
-    old.par <- par(no.readonly = TRUE)
-    on.exit(par(old.par))
-
-    n <- length(x)
-
-    mai <- par("mai")
-    w <- max(strwidth(names(x), "inch"), na.rm = TRUE) + 1/16
-    if (mai[2L] < w)
-        mai[2L] <- mai[4L] + w # taken from dotchart
-    par(mai = mai)
-
-    plot(NA, xlim = xlim, ylim = c(0L, n + 1L), axes = FALSE)
-    title(main = "Missing Data", adj = 0L)
-    title(xlab = "Frequency", adj = 1L)
-    y <- seq_len(n)
-    mtext(
-        names(x), at = y, adj = 0L, side = 2L, las = 2L,
-        line = (w + 0.1) / par("csi"), cex = 0.8
-    )
-    abline(h = y, col = "#808080", lty = "dotted", lwd = 1L)
-    points(x, y, col = col, pch = pch)
-    axis(1L, lwd.ticks = 0L, col = "#808080")
-}
 
@@
 
-.dotplot(sort(na), xlim = c(0L, nrow(m)))
+plot_dots(sort(na), xlim = c(0L, nrow(m)), main = "Missing Data")
 ```

### 2021-07-02 analysis/histograms.Rmd
feat: add MELD histograms

@@
 ```
+
+# MELD Scores (imputed)
+
+```{r histograms_meld, results = "asis", message = FALSE, warning = FALSE}
+meldv <- c(paste0("ScoreMeld", c("", "Unos", "Na", "NaUnos")), "ProbMeldPlus7")
+for (v in meldv) {
+    cat("\n\n##", v , "\n\n")
+    .hist(zlog_data[[v]], main = paste0("Histogram of ", v), xlab = v)
+    .qqnorm(zlog_data[[v]], yname = v)
+}
+```

### 2021-07-02 code/01-data.R

@@ -15,7 +15,7 @@ tg_data <- list(
         set_missing_limits(r)
     }, packages = "zlog"),
     tar_target(imp_data, {
-        withCallingHandlers(
+        imp <- withCallingHandlers(
             impute_df(raw_data, ref_data, method = "logmean"),
             warning = function(w) {
                 if(w$message != paste0(
@@ -27,9 +27,69 @@ tg_data <- list(
                     invokeRestart("muffleWarning")
             }
         )
+        ## assume no dialysis for missing values
+        imp$Dialysis[is.na(imp$Dialysis)] <- 0L
+        imp
+    }),
+    tar_target(meld_data, {
+        m <- imp_data
+        m$ScoreMeld <- meld(
+            creatinine = as_metric(m$CRE_S, "creatinine"),
+            bilirubin = as_metric(m$BILI_S, "bilirubin"),
+            inr = m$INR_C,
+            dialysis = m$Dialysis,
+            cause = ifelse(m$Ethyltoxic == 1, "ethyltoxisch", "other")
+        )
+        m$ScoreMeldUnos <- meld(
+            creatinine = as_metric(m$CRE_S, "creatinine"),
+            bilirubin = as_metric(m$BILI_S, "bilirubin"),
+            inr = m$INR_C,
+            dialysis = m$Dialysis,
+            cause = "other"
+        )
+        m$ScoreMeldNa <- meld_na(
+            creatinine = as_metric(m$CRE_S, "creatinine"),
+            bilirubin = as_metric(m$BILI_S, "bilirubin"),
+            inr = m$INR_C,
+            sodium = m$NA_S,
+            dialysis = m$Dialysis,
+            cause = ifelse(m$Ethyltoxic == 1, "ethyltoxisch", "other")
+        )
+        m$ScoreMeldNaUnos <- meld_na(
+            creatinine = as_metric(m$CRE_S, "creatinine"),
+            bilirubin = as_metric(m$BILI_S, "bilirubin"),
+            inr = m$INR_C,
+            sodium = m$NA_S,
+            dialysis = m$Dialysis,
+            type = "UNOS"
+        )
+        m$ProbMeldPlus7 <- meld_plus7(
+            creatinine = as_metric(m$CRE_S, "creatinine"),
+            bilirubin = as_metric(m$BILI_S, "bilirubin"),
+            inr = m$INR_C,
+            sodium = m$NA_S,
+            albumin = m$ALB_S / 10,
[…]

### 2021-07-02 analysis/linecharts.Rmd
feat: add plot MeldCategory vs mean std values

@@
 
-```{r linecharts_zlogmeanplot}
+```{r linecharts_zlogmeanplot_status}
 labv <- sort(grep("_[SECQ]$", colnames(zlog_data), value = TRUE))
@@
 
-par(mar = c(7.1, 4.1, 4.1, 2.1))
+par(mar = c(7.1, 4.1, 4.1, 2.1), cex = 0.5)
 
@@
 ```
+
+# Compare Mean Zlog Values vs MeldCategory
+
+```{r linecharts_zlogmeanplot_meld}
+## divide data.frame by MELD
+s <- split(zlog_data[labv], zlog_data$MeldCategory)
+
+## calculate mean standardized lab values
+s <- lapply(s, colMeans, na.rm = TRUE)
+n <- length(s)
+o <- order(s[[n]])
+
+## comparison plot
+col <- rev(palette.colors(n))
+
+## keep old graphic parameters and restore them afterwards
+old.par <- par(no.readonly = TRUE)
+
+par(mar = c(7.1, 4.1, 4.1, 2.1), cex = 0.5)
+
+plot(
+    s[[n]][o], type = "b", pch = 20, lwd = 2, col = col[n],
+    axes = FALSE, ann = FALSE
+)
+for (i in seq_len(n - 1)) {
+    lines(s[[i]][o], type = "b", pch = 20, lwd = 2, col = col[i])
+}
+legend(
+    "bottomright",
+    legend = names(s),
+    col = col, lwd = 2, pch = 20, bty = "n"
+)
+title(
+    main = "MELD Category vs Mean Standardized Laboratory Values", adj = 0
+)
+title(xlab = "Laboratory Measurements", adj = 1L, line = 5)
+title(ylab = "Mean Standardized Values", adj = 1L)
+r <- range(unlist(s))
+axis(
+    2, at = seq(from = floor(r[1]), to = ceiling(r[2])),
+    lwd.ticks = 0, col = "#808080"
+)
+axis(
+    1, at = seq_along(o), labels = names(s[[n]][o]), las = 2,
+    lwd.ticks = 0L, col = "#808080"
+)
+abline(h = 0, col = "#808080", lty = 2)
+par(old.par)
+```

### 2021-07-02 analysis/survivalplots.Rmd
feat: add survival plots

---
title: "Survival Plots"
output:
  workflowr::wflow_html:
    toc: true
    fig.width: 12
    fig.height: 8
editor_options:
  chunk_output_type: console
---

```{r survplots_setup, message = FALSE}
library("targets")
library("ameld")
library("survival")
tar_load(zlog_data)

srv <- Surv(zlog_data$DaysAtRisk, zlog_data$Deceased)
col <- palette.colors(nlevels(zlog_data$MeldCategory))
```

# Survival

```{r survplots_total}
srvfit <- survfit(srv ~ 1)
plot_surv(
    srvfit,
    main = "Kaplan-Meier Survival Estimate",
    col = col[2]
)
```

```{r survplots_90d}
times <- c(0, 7, 14, 30, 60, 90)

## calculate risk tables
sm <- summary(srvfit, times = times)
nrisk <- as.matrix(sm$n.risk)
ncumevents <- as.matrix(cumsum(sm$n.event))
rownames(nrisk) <- rownames(ncumevents) <- times

## keep old graphic parameters and restore them afterwards
old.par <- par(no.readonly = TRUE)

layout(matrix(1:3, nrow = 3), height = c(5, 1.5, 1.5))
plot_surv(
    srvfit,
    main = "Kaplan-Meier Survival Estimate to day 90",
    times = times,
    xmax = 90,
    col = col[2]
)
par(mar = c(5.1, 5.1, 1.1, 2.1))
plot_table(
    nrisk, at = times, main = "Number at risk",
    xaxis = FALSE, cex.text = 1.5, ylabels = FALSE
)
par(mar = c(5.1, 5.1, 1.1, 2.1))
plot_table(
    ncumevents, at = times, main = "Cumulative number of events",
    xaxis = FALSE, cex.text = 1.5, ylabels = FALSE
)
par(old.par)
```

# Survival vs MELD Category

```{r survplots_meld_total}
srvfit <- survfit(srv ~ MeldCategory, data = zlog_data)
col <- rev(col)
plot_surv(
    srvfit,
    main = "Kaplan-Meier Survival Estimate",
    col = col
)
legend(
    "bottomright",
    legend = levels(zlog_data$MeldCategory),
    col = col, lwd = 1, pch = 3, bty = "n"
)
```

```{r survplots_meld_90d, fig.width = 12, fig.height = 12}
## calculate risk tables
sm <- summary(srvfit, times = times)
ncumevents <- nrisk <- matrix(
    NA_real_,
    nrow = length(times), ncol = nlevels(zlog_data$MeldCategory),
    dimnames = list(times, levels(zlog_data$MeldCategory))
)
nrisk[][seq_along(sm$n.risk)] <- sm$n.risk
ncumevents[][seq_along(sm$n.event)] <- sm$n.event
ncumevents <- apply(ncumevents, MARGIN = 2, cumsum)

## keep old graphic parameters and restore them afterwards
old.par <- par(no.readonly = TRUE)

layout(matrix(1:3, nrow = 3), height = c(5, 2, 2))
par(mar = c(5.1, 4.1, 1.1, 2.1))
plot_surv(
    srvfit,
    main = "Kaplan-Meier Survival Estimate to day 90",
    times = times,
    xmax = 90,
    col = col
)
legend(
    "bottomright",
    legend = levels(zlog_data$MeldCategory),
    col = col, lwd = 1, pch = 3, bty = "n"
)
par(mar = c(5.1, 5.1, 1.1, 2.1))
plot_table(
    nrisk, at = times, main = "Number at risk",
    xaxis = FALSE, cex.text = 1.5, col = col
)
par(mar = c(5.1, 5.1, 1.1, 2.1))
plot_table(
    ncumevents, at = times, main = "Cumulative number of events",
    xaxis = FALSE, cex.text = 1.5, col = col
)
par(old.par)
```

### 2021-07-02 code/03-reports.R

@@ -78,6 +78,19 @@ tg_reports <- list(
         format = "file"
     ),
     tar_target(
+        report_survivalplots,
+        command = {
+            list(site_conf)
+            !!tar_knitr_deps_expr(file.path("analysis", "survivalplots.Rmd"))
+            workflowr::wflow_build(
+                file.path("analysis", "survivalplots.Rmd"),
+                view = FALSE
+           )
+           file.path("analysis", "survivalplots.Rmd")
+        },
+        format = "file"
+    ),
+    tar_target(
         report_corrplot,
         command = {
             list(site_conf)

### 2021-07-14 scripts/R.sh
chore: upgrade singularity container with latest guix and fix slurm

@@ -10,22 +10,12 @@
 # SINGULARITYENV_R_CLI_NUM_COLORS=1 \
 WORKAROUND='assignInNamespace("win10_build", function()0L, ns = "cli")'
 
-REXPR=${1:-"${WORKAROUND}; targets::tar_make(); workflowr::wflow_build(rprojroot::find_rstudio_root_file('analysis', 'pipeline.Rmd'), view = FALSE)"}
-PREPEND_PATH=""
-BIND_PATH="/var/run"
-
-## on an HPC
-if ! [[ ${HOSTNAME} =~ (x1|t400) ]] ; then
-    module load singularity slurm
-    PREPEND_PATH=$(IFS=:; for i in ${PATH}; do if [[ $i =~ (slurm|munge|singularity) ]] ; then echo -n "${i}:"; fi; done;)
-    PREPEND_PATH=${PREPEND_PATH%:}
-    BIND_PATH="/etc/slurm/,/usr/lib64/libmunge.so.2,${BIND_PATH},$(echo ${PREPEND_PATH} | sed 's#:#\n#g; s#\(bin\|lib\)\/\?##g' | uniq | sed ':a;N;$!ba;s#\n#,#g')"
-fi
+REXPR=${1:-"${WORKAROUND}; targets::tar_make_future(); workflowr::wflow_build(rprojroot::find_rstudio_root_file('analysis', 'pipeline.Rmd'), view = FALSE)"}
+BIND_PATH="/etc/passwd,/etc/group,/var/run,${SSH_AUTH_SOCK}"
 
 SINGULARITYENV_LOCPATH=/lib/locale/2.31/ \
     SINGULARITYENV_TAR_MAKE_REPORTER="timestamp" \
     SINGULARITYENV_TZ="UTC" \
-    SINGULARITYENV_PREPEND_PATH="${PREPEND_PATH}" \
     SINGULARITY_BIND="${BIND_PATH}" \
-    singularity run container/ampel-leipzig-meld.sif \
-    --vanilla --silent -e "${REXPR}"
+    singularity exec container/ampel-leipzig-meld.squashfs \
+    R --vanilla --silent -e "${REXPR}"

### 2021-07-14 scripts/Rscript.sh

@@ -1,0 +1,14 @@
+#!/bin/bash
+
+# should run on a HPC
+
+. /usr/share/Modules/init/bash
+module load singularity slurm
+
+BIND_PATH="/etc/group,/etc/passwd,/etc/slurm,/var/run"
+
+SINGULARITYENV_LOCPATH=/lib/locale/2.31/ \
+    SINGULARITYENV_TAR_MAKE_REPORTER="timestamp" \
+    SINGULARITYENV_TZ="UTC" \
+    SINGULARITY_BIND="${BIND_PATH}" \
+    singularity run container/ampel-leipzig-meld.squashfs "$@"

### 2021-07-14 analysis/benchmarks.Rmd
feat: add mlr benchmarks

---
title: "Benchmarks"
output:
  workflowr::wflow_html:
    toc: true
    fig.width: 12
    fig.height: 8
editor_options:
  chunk_output_type: console
---

```{r benchmarks_setup, message = FALSE}
library("targets")
library("mlr3viz")
library("viridis")
library("ggplot2")
tar_load(bmrk_results)
```

```{r benchmarks_plot_c_harrell}
autoplot(bmrk_results) +
    geom_boxplot(aes(fill = learner_id)) +
    geom_jitter(position = position_jitter(0.2)) +
    scale_fill_viridis(discrete = TRUE) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
```

### 2021-07-14 _targets.R

@@ -1,17 +1,51 @@
 library("targets")
 library("tarchetypes")
 
-tar_option_set(packages = c(
-    "ameld",
-    "zlog"
-))
+library("future")
+library("future.batchtools")
+library("future.callr")
+
+#Sys.setenv(RUNLOCAL = 1)
+Sys.setenv(DEBUGME = "batchtools")
+lgr::get_logger("mlr3")$set_threshold("info")
+lgr::get_logger("bbotk")$set_threshold("info")
+
+login <- future::tweak(
+    future::remote,
+    workers = "brain",
+    rscript = "scripts/Rscript.sh",
+    rshopts = if (nchar(Sys.getenv("DEBUGME"))) "-v" else ""
+)
+
+slurm <- future::tweak(
+    future.batchtools::batchtools_slurm,
+    template = "scripts/slurm_batchtools.tmpl",
+    workers = 4L,
+    resources = list(
+        logfile = "logs/job%s.log",
+        partition = "batch,compute,pinky,snowball",
+        ncpu = 8L,
+        mcpu = 16L * 1024L,
+        walltime = 4L * 60L * 60L
+    )
+)
+
+if (nchar(Sys.getenv("RUNLOCAL"))) {
+    future::plan(callr, workers = 3)
+} else {
+    future::plan(list(login, slurm, future::tweak(callr, workers = 8L)))
+}
 
 source("code/01-data.R")
+source("code/02-benchmarking.R")
 source("code/03-reports.R")
 source("code/99-site.R")
 
 list(
    tg_data,
+   tg_tasks,
+   tg_benchmarks,
+   tg_benchmark_results,
    tg_reports,
    tg_site
 )

### 2021-07-14 code/02-benchmarking.R

@@ -1,0 +1,244 @@
+library("mlr3")
+library("mlr3extralearners")
+library("mlr3learners")
+library("mlr3proba")
+library("mlr3tuning")
+library("paradox")
+
+## learners
+lrn_cox <- lrn("surv.coxph", id = "cox")
+
+lrn_lasso <- lrn("surv.cv_glmnet", id = "lasso", alpha = 1L)
+
+lrn_ridge <- lrn("surv.cv_glmnet", id = "ridge", alpha = 0L)
+
+lrn_elasticnet <- lrn("surv.cv_glmnet", id = "elasticnet")
+
+lrn_penlasso <- lrn("surv.penalized", id = "penlasso",
+    lambda2 = 0, standardize = TRUE, trace = FALSE
+)
+
+lrn_penridge <- lrn("surv.penalized", id = "penridge",
+    lambda1 = 0, standardize = TRUE, trace = FALSE
+)
+
+lrn_ranger <- lrn("surv.ranger", id = "ranger",
+    # ?ranger, 3 for survival
+    min.node.size = 3L
+)
+
+lrn_rfsrc <- lrn("surv.rfsrc",
+    # ?randomForestSRC, nsplit = 0, randomized splitting
+    nsplit = 0
+)
+
+lrn_xgboost <- lrn("surv.xgboost", max_depth = 6)
+
+lrn_svmregression <- lrn("surv.svm", id = "svm-regression",
+    type = "regression",
+    gamma.mu = 0 # would be trained later, is required by ctor
+)
+
+lrn_svmvanbelle1 <- lrn("surv.svm", id = "svm-vanbelle1",
+    type = "vanbelle1",
+    diff.meth = "makediff3",
+    kernel = "rbf_kernel",
+    opt.meth = "quadprog",
+    gamma.mu = 0 # would be trained later, is required by ctor
+)
+
+## params
+ps_lasso <- ParamSet$new(
+    params = list(
+        ParamDbl$new("s", lower = 0L, upper = 1L)
+    )
+)
+
+ps_ridge <- ParamSet$new(
+    params = list(
+        ParamDbl$new("s", lower = 0L, upper = 1L)
+    )
+)
+
+ps_elasticnet <- ParamSet$new(
+    params = list(
+        ParamDbl$new("alpha", lower = 0L, upper = 1L),
+        ParamDbl$new("s", lower = 0L, upper = 1L)
+    )
+)
+
+ps_penlasso <- ParamSet$new(
[…]

### 2021-07-14 code/01-data.R
refactor: add deployment arguments

@@ -1,98 +1,118 @@
 tg_data <- list(
     tar_target(raw_data, {
-        data(eldd, package = "ameld")
-        eldd
-    }),
+            data(eldd, package = "ameld")
+            eldd
+        },
+        packages = "ameld",
+        deployment = "main"
+    ),
     tar_target(ref_data, {
-        ## prepare reference data for zlog normalisation
-        data(eldr, package = "ameld")
-        r <- eldr[c("Code", "AgeDays", "Sex", "LowerLimit", "UpperLimit")]
-        names(r) <- c("param", "age", "sex", "lower", "upper")
-        r$age <- r$age / 365.25
+            ## prepare reference data for zlog normalisation
+            data(eldr, package = "ameld")
+            r <- eldr[c("Code", "AgeDays", "Sex", "LowerLimit", "UpperLimit")]
+            names(r) <- c("param", "age", "sex", "lower", "upper")
+            r$age <- r$age / 365.25
 
-        ## set missing lower limits to 0.15 of the upper according to
-        ## Haeckel et al. 2015
-        set_missing_limits(r)
-    }, packages = "zlog"),
+            ## set missing lower limits to 0.15 of the upper according to
+            ## Haeckel et al. 2015
+            set_missing_limits(r)
+        },
+        packages = "zlog",
+        deployment = "main"
+    ),
     tar_target(imp_data, {
-        imp <- withCallingHandlers(
-            impute_df(raw_data, ref_data, method = "logmean"),
-            warning = function(w) {
-                if(w$message != paste0(
-                    "No reference for column(s): DaysAtRisk, Deceased, LTx, ",
-                    "Cirrhosis, ALF, Ethyltoxic, HBV, HCV, AIH, PBC, PSC, ",
-                    "NASH, Cryptogenic, Dialysis, GIB, HCC, SBP"))
-                       stop(w$message)
-                else
-                    invokeRestart("muffleWarning")
-            }
-        )
-        ## assume no dialysis for missing values
-        imp$Dialysis[is.na(imp$Dialysis)] <- 0L
-        imp
-    }),
+            imp <- withCallingHandlers(
+                impute_df(raw_data, ref_data, method = "logmean"),
+                warning = function(w) {
+                    if(w$message != paste0(
[…]

### 2021-07-14 code/03-reports.R

@@ -17,7 +17,8 @@ tg_reports <- tar_eval(
             !!tar_knitr_deps_expr(rpt)
             workflowr::wflow_build(rpt, view = FALSE)
             rpt
-        }, format = "file"
+        }, format = "file",
+        deployment = "main"
     )
 )
 

### 2021-07-14 code/99-site.R

@@ -13,7 +13,8 @@ tg_site <- list(
             )
             file.path("analysis", "index.Rmd")
         },
-        format = "file"
+        format = "file",
+        deployment = "main"
     ),
     tar_target(
         site_license,
@@ -27,6 +28,7 @@ tg_site <- list(
 
             file.path("analysis", "license.Rmd")
         },
-        format = "file"
+        format = "file",
+        deployment = "main"
     )
 )

### 2021-07-14 _targets.R
chore: increase number of workers

@@ -20,7 +20,7 @@ login <- future::tweak(
 slurm <- future::tweak(
     future.batchtools::batchtools_slurm,
     template = "scripts/slurm_batchtools.tmpl",
-    workers = 4L,
+    workers = 8L,
     resources = list(
         logfile = "logs/job%s.log",
         partition = "batch,compute,pinky,snowball",
@@ -33,7 +33,7 @@ slurm <- future::tweak(
 if (nchar(Sys.getenv("RUNLOCAL"))) {
     future::plan(callr, workers = 3)
 } else {
-    future::plan(list(login, slurm, future::tweak(callr, workers = 8L)))
+    future::plan(list(login, slurm, callr))
 }
 
 source("code/01-data.R")

### 2021-07-15 scripts/R.sh
chore: increase number of allowed transient workers

@@ -10,7 +10,7 @@
 # SINGULARITYENV_R_CLI_NUM_COLORS=1 \
 WORKAROUND='assignInNamespace("win10_build", function()0L, ns = "cli")'
 
-REXPR=${1:-"${WORKAROUND}; targets::tar_make_future(); workflowr::wflow_build(rprojroot::find_rstudio_root_file('analysis', 'pipeline.Rmd'), view = FALSE)"}
+REXPR=${1:-"${WORKAROUND}; targets::tar_make_future(workers = 4L); workflowr::wflow_build(rprojroot::find_rstudio_root_file('analysis', 'pipeline.Rmd'), view = FALSE)"}
 BIND_PATH="/etc/passwd,/etc/group,/var/run,${SSH_AUTH_SOCK}"
 
 SINGULARITYENV_LOCPATH=/lib/locale/2.31/ \

### 2021-08-01 analysis/article.Rmd
feat: add first article draft with pandoc filters

---
editor_options:
  chunk_output_type: console
title: "A Model to Predict Survival in Patients with End-Stage Liver Disease"
subtitle: >
    Comparision against MELD, MELD-Na, MELD-Plus7.
abstract: >
  **Background**:
      TODO

  **Methods**:
      TODO

  **Results**:
      TODO

  **Conclusion**:
      TODO

author:
    - Sebastian Gibb:
        email: mail@sebastiangibb.de
        institute: [ILM, KAINS]
        correspondence: true
    - Thorsten Kaiser:
        email: thorsten.kaiser@medizin.uni-leipzig.de
        institute: ILM
        correspondence: false
institute:
    - ILM: Institute of Laboratory Medicine, Clinical Chemistry and Molecular Diagnostics, University Hospital Leipzig, Paul-List-Str. 13-15, D-04103 Leipzig, Germany.
    - KAINS: Anesthesiology and Intensive Care Medicine, University Hospital Greifswald, Ferdinand-Sauerbruch-Straße, D-17475 Greifswald, Germany.
date: "`r Sys.Date()`"
bibliography:
  - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'bibliography.bib')`"
  - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'rpackages', 'article.bib')`"
output:
  workflowr::wflow_html:
    pandoc_args:
      - --filter=pandoc-crossref
        # pandoc_args doesn't support r evaluation
      - "--lua-filter=pandoc/lua-filters/scholarly-metadata.lua"
      - "--lua-filter=pandoc/lua-filters/author-info-blocks.lua"
---

```{r, include=FALSE}
knitr::write_bib(
    c("targets", "workflowr"),
    file = rprojroot::find_rstudio_root_file(
        "analysis", "bibliography", "rpackages", "article.bib"
    )
)
```

# Introduction

Liver cirrhosis is the terminal result of fibrotic remodeling
of liver tissue due to chronic damage. An organ failure is often
irreversible and the only available therapy is liver transplantation.
However, the shortage in grafts for transplantation from deceased donors
requires risk stratification and precise allocation rules.
The allocation of liver transplantation in most countries is based on
disease severity determined by the model of end-stage liver disease (MELD)
[@malinchoc2000; @wiesner2003; @optn_policies2021].
The MELD score estimates the patients' 3-month mortality risk based on
laboratory results, namely bilirubin, creatinine and international normalized
ratio (INR). In general the MELD score is extended by the sodium level
(MELD-Na score) as this was found to be an important additional risk factor in
liver cirrhosis [@kim2008; @optn_policies2021].
The MELD was initially developed for predicting the survival of patients
undergoing transjugular intrahepatic portosystemic shunts.
Afterwards it was revalidated for predicting mortality risk in patients awaiting
a liver transplantation.
Although the MELD score should be an objective allocation score especially the
creatinine and INR are highly dependent on the utilized laboratory methods
[@trotter2004; @cholongitas2007]. Patients with identical disease state could
have very different MELD scores and thus get different priority on the liver
transplantation waiting list.
Furthermore often, e.g. for acute-on-chronic liver failure, the MELD score
underestimates the mortality risk [@hernaez2020].

There have been some attempts to use the data extracted from more than 300.000
electronic medical records from two hospitals in the United States to improve
the MELD score [@kartoun2017]. The derived MELD-Plus7 and MELD-Plus9 risk score
add albumin, white blood cell count, age and total cholesterol and length of
stay to the MELD-Na variables. Despite its published prediction improvement the
MELD-Plus scores are not used for transplant allocation yet.

As depicted by MELD-Plus risk scores better predictive scores often need more
variables and are more complicated. To reduce the risk of overlooking or
incorrectly calculating and interpreting the results,
clinical decision support systems may be used and could improve patient safety.
The research project on digital laboratory medicine (AMPEL) develops a
clinical decision support system based on laboratory diagnostics that should
support clinical practitioners in interpreting the laboratory results and taking
the necessary medical interventions [@eckelt2020].

This study aims to find clinical and laboratory values that improve the risk
stratification for liver transplantation over classical MELD, MELD-Na and
MELD-Plus scores and could be implemented as part of the AMPEL clinical decision
support system.

# Material and methods

# Results

# Discussion

# References

### 2021-08-01 analysis/supplement.Rmd
feat: add first article draft with pandoc filters

---
title: "Supplement"
output:
  workflowr::wflow_html:
    toc: false
editor_options:
  chunk_output_type: console
---

```{r supplement_setup, message = FALSE}
library("targets")
```

### 2021-08-01 analysis/pandoc/lua-filters/fetch.sh

@@ -1,0 +1,8 @@
+#!/bin/sh
+CURLOPT="--silent --remote-name"
+FILTERS="scholarly-metadata author-info-blocks"
+GITHUBURL="https://raw.githubusercontent.com/pandoc/lua-filters/master"
+
+for FILTER in ${FILTERS}; do
+    curl ${CURLOPT} ${GITHUBURL}/${FILTER}/${FILTER}.lua
+done

### 2021-08-01 code/02-benchmarking.R
fix: suppress package startup messages

@@ -1,9 +1,9 @@
-library("mlr3")
-library("mlr3extralearners")
-library("mlr3learners")
-library("mlr3proba")
-library("mlr3tuning")
-library("paradox")
+suppressPackageStartupMessages(library("mlr3"))
+suppressPackageStartupMessages(library("mlr3extralearners"))
+suppressPackageStartupMessages(library("mlr3learners"))
+suppressPackageStartupMessages(library("mlr3proba"))
+suppressPackageStartupMessages(library("mlr3tuning"))
+suppressPackageStartupMessages(library("paradox"))
 
 ## learners
 lrn_cox <- lrn("surv.coxph", id = "cox")

### 2021-08-02 analysis/article.Rmd
refactor: use bookdown::html_document2 instead of workflowr::wflow_html to allow crossreferences

@@
 ---
+# R markdown template taken from
+# Dan Holmes <dtholmes@mail.ubc.ca>
+# https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
 editor_options:
@@
 output:
-  workflowr::wflow_html:
+  bookdown::html_document2:
     pandoc_args:
-      - --filter=pandoc-crossref
         # pandoc_args doesn't support r evaluation

### 2021-08-02 code/make.R
refactor: unitise Rscript scripts

@@ -1,0 +1,15 @@
+# The cli package has an internal function cli:::win10_build
+# that uses utils::sessionInfo()$running, which returns osVersion.
+# osVersion is NULL in the container and `tar_make` will throw an error
+# because cli::win10_build fails because grepl("^Win...", osVersion) won't work.
+# This overwrites win10_build and just returns 0L
+# https://github.com/r-lib/cli/issues/273
+# We could also use
+# SINGULARITYENV_R_CLI_NUM_COLORS=1 \
+assignInNamespace("win10_build", function()0L, ns = "cli")
+
+targets::tar_make_future(workers = 4L)
+workflowr::wflow_build(
+        rprojroot::find_rstudio_root_file("analysis", "pipeline.Rmd"),
+        view = FALSE
+)

### 2021-08-02 scripts/R.sh

@@ -1,21 +1,0 @@
-#!/bin/bash
-
-# The cli package has an internal function cli:::win10_build
-# that uses utils::sessionInfo()$running, which returns osVersion.
-# osVersion is NULL in the container and `tar_make` will throw an error
-# because cli::win10_build fails because grepl("^Win...", osVersion) won't work.
-# This overwrites win10_build and just returns 0L
-# https://github.com/r-lib/cli/issues/273
-# We could also use
-# SINGULARITYENV_R_CLI_NUM_COLORS=1 \
-WORKAROUND='assignInNamespace("win10_build", function()0L, ns = "cli")'
-
-REXPR=${1:-"${WORKAROUND}; targets::tar_make_future(workers = 4L); workflowr::wflow_build(rprojroot::find_rstudio_root_file('analysis', 'pipeline.Rmd'), view = FALSE)"}
-BIND_PATH="/etc/passwd,/etc/group,/var/run,${SSH_AUTH_SOCK}"
-
-SINGULARITYENV_LOCPATH=/lib/locale/2.31/ \
-    SINGULARITYENV_TAR_MAKE_REPORTER="timestamp" \
-    SINGULARITYENV_TZ="UTC" \
-    SINGULARITY_BIND="${BIND_PATH}" \
-    singularity exec container/ampel-leipzig-meld.squashfs \
-    R --vanilla --silent -e "${REXPR}"

### 2021-08-02 scripts/Rscript.sh

@@ -1,14 +1,23 @@
 #!/bin/bash
 
-# should run on a HPC
+BIND_PATH="/etc/passwd,/etc/group,/var/run"
 
-. /usr/share/Modules/init/bash
-module load singularity slurm
-
-BIND_PATH="/etc/group,/etc/passwd,/etc/slurm,/var/run"
+# on HPC?
+MODULEPATH=/usr/share/Modules/init/bash
+if [ -f "${MODULEPATH}" ] ; then
+    SINGULARITYVER=3.4.2
+    . "${MODULEPATH}"
+    module load singularity/${SINGULARITYVER} slurm
+    SINGULARITYPATH=/opt/singularity-${SINGULARITYVER}/bin/singularity
+    BIND_PATH="${BIND_PATH},/etc/slurm"
+else # local
+    BIND_PATH="${BIND_PATH},${SSH_AUTH_SOCK}"
+    SINGULARITYPATH=/usr/bin/singularity
+fi
 
 SINGULARITYENV_LOCPATH=/lib/locale/2.31/ \
     SINGULARITYENV_TAR_MAKE_REPORTER="timestamp" \
     SINGULARITYENV_TZ="UTC" \
     SINGULARITY_BIND="${BIND_PATH}" \
-    singularity run container/ampel-leipzig-meld.squashfs "$@"
+    ${SINGULARITYPATH} run container/ampel-leipzig-meld.squashfs "${@}"
+exit 0

### 2021-08-02 code/01-data.R
fix: update ameld to get correct Long/ShortDescription

@@ -9,8 +9,12 @@ tg_data <- list(
     tar_target(ref_data, {
             ## prepare reference data for zlog normalisation
             data(eldr, package = "ameld")
-            r <- eldr[c("Code", "AgeDays", "Sex", "LowerLimit", "UpperLimit")]
-            names(r) <- c("param", "age", "sex", "lower", "upper")
+            r <- eldr[c(
+                "Code", "AgeDays", "Sex", "LowerLimit", "UpperLimit",
+                "LongDescription", "ShortDescription"
+            )]
+            names(r) <-
+                c("param", "age", "sex", "lower", "upper", "long", "short")
             r$age <- r$age / 365.25
 
             ## set missing lower limits to 0.15 of the upper according to

### 2021-08-02 analysis/index.Rmd
refactor: move common yaml headers into _site.yml

@@
 output:
-  workflowr::wflow_html:
+  bookdown::html_document2:
     toc: false
-editor_options:
-  chunk_output_type: console
 bibliography:

### 2021-08-02 analysis/benchmarks.Rmd
refactor: move common yaml headers into _site.yml
fix: working directory for targets

@@
 output:
-  workflowr::wflow_html:
-    toc: true
+  bookdown::html_document2:
     fig.width: 12
     fig.height: 8
-editor_options:
-  chunk_output_type: console
 ---
@@
 library("ggplot2")
+
+withr::with_dir(rprojroot::find_rstudio_root_file(), {
     tar_load(bmrk_results)
+})
 ```

### 2021-08-02 analysis/boxplots.Rmd
refactor: move common yaml headers into _site.yml
fix: working directory for targets

@@
 title: "Boxplots"
-output:
-  workflowr::wflow_html:
-    toc: true
-editor_options:
-  chunk_output_type: console
 ---
@@
 library("viridisLite")
+
+withr::with_dir(rprojroot::find_rstudio_root_file(), {
     tar_load(imp_data)
     tar_load(zlog_data)
+})
 

### 2021-08-02 analysis/corrplot.Rmd
fix: suppress package startup messages
fix: working directory for targets

@@
 output:
-  workflowr::wflow_html:
-    toc: true
+  bookdown::html_document2:
     fig.width: 12
     fig.height: 12
-editor_options:
-  chunk_output_type: console
 ---
 
-```{r corrplot_setup}
+```{r corrplot_setup, message = FALSE}
 library("corrplot")
 library("targets")
+
+withr::with_dir(rprojroot::find_rstudio_root_file(), {
     tar_load(imp_data)
+})
 ```

### 2021-08-02 analysis/histograms.Rmd
fix: working directory for targets

@@
 title: "Histograms"
-output:
-  workflowr::wflow_html:
-    toc: true
-editor_options:
-  chunk_output_type: console
 ---
@@
 library("targets")
+
+withr::with_dir(rprojroot::find_rstudio_root_file(), {
     tar_load(imp_data)
     tar_load(zlog_data)
+})
 

### 2021-08-02 analysis/linecharts.Rmd
refactor: move common yaml headers into _site.yml
fix: working directory for targets

@@
 library("targets")
+
+withr::with_dir(rprojroot::find_rstudio_root_file(), {
     tar_load(zlog_data)
+})
 ```

### 2021-08-02 analysis/missing.Rmd
fix: load missing ameld package
fix: working directory for targets

@@
 output:
-  workflowr::wflow_html:
-    toc: true
+  bookdown::html_document2:
     fig.width: 10
     fig.height: 10
-editor_options:
-  chunk_output_type: console
 ---
@@
 library("targets")
+library("ameld")
+
+withr::with_dir(rprojroot::find_rstudio_root_file(), {
     tar_load(raw_data)
+})
 

### 2021-08-02 analysis/survivalplots.Rmd
fix: working directory for targets

@@
 library("survival")
+
+withr::with_dir(rprojroot::find_rstudio_root_file(), {
     tar_load(zlog_data)
+})
 

### 2021-08-02 analysis/tables.Rmd
fix: working directory for targets

@@
 title: "Tables"
-output:
-  workflowr::wflow_html:
-    toc: true
-editor_options:
-  chunk_output_type: console
 ---
@@
 library("skimr")
+
+withr::with_dir(rprojroot::find_rstudio_root_file(), {
     tar_load(raw_data)
@@
     tar_load(zlog_data)
+})
 ```

### 2021-08-03 code/make.R
refactor: remove now useless cli workaround

@@ -1,14 +1,4 @@
-# The cli package has an internal function cli:::win10_build
-# that uses utils::sessionInfo()$running, which returns osVersion.
-# osVersion is NULL in the container and `tar_make` will throw an error
-# because cli::win10_build fails because grepl("^Win...", osVersion) won't work.
-# This overwrites win10_build and just returns 0L
-# https://github.com/r-lib/cli/issues/273
-# We could also use
-# SINGULARITYENV_R_CLI_NUM_COLORS=1 \
-assignInNamespace("win10_build", function()0L, ns = "cli")
-
-targets::tar_make_future(workers = 4L)
+targets::tar_make_future()
 workflowr::wflow_build(
         rprojroot::find_rstudio_root_file("analysis", "pipeline.Rmd"),
         view = FALSE

### 2021-08-03 analysis/article.Rmd
feat: add flowchart and discussion notes

--- 2021-08-01 analysis/article.Rmd
@@
 ---
-editor_options:
-  chunk_output_type: console
+# R markdown template taken from
+# Dan Holmes <dtholmes@mail.ubc.ca>
+# https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
 title: "A Model to Predict Survival in Patients with End-Stage Liver Disease"
@@
 output:
-  workflowr::wflow_html:
-    pandoc_args:
-      - --filter=pandoc-crossref
-        # pandoc_args doesn't support r evaluation
-      - "--lua-filter=pandoc/lua-filters/scholarly-metadata.lua"
-      - "--lua-filter=pandoc/lua-filters/author-info-blocks.lua"
+  bookdown::html_document2:
+    pandoc_args: # pandoc_args doesn't support r evaluation
+      - --lua-filter=pandoc/lua-filters/scholarly-metadata.lua
+      - --lua-filter=pandoc/lua-filters/author-info-blocks.lua
 ---
 
-```{r, include=FALSE}
+```{r knitr_setup, include=FALSE}
 knitr::write_bib(
@@
 )
+```
+
+```{r article_setup, include=FALSE}
+library("targets")
+
+tar_load(raw_data)
+tar_load(bmrk_results)
 ```
@@
 
+## Study population
+
+```{r excluded, include = FALSE}
+n_excluded <- attr(raw_data, "excluded")
+n_analysed <- nrow(raw_data)
+n_total <- sum(n_analysed, n_excluded)
+
+cn_labs <- grep("_[CESQ]$", colnames(raw_data), value = TRUE)
+```
+
+In a retrospective cohort study we followed `r n_total` consecutive
+patients, who were recruited during the evaluation process for
+liver transplantation at the University Hospital of Leipzig from
+`r paste(format(attr(raw_data, "period"), "%B %Y"), collapse = " to ")`.
+For each patient we recorded `r ncol(raw_data)` variables.
+Among them are age, sex, etiology of liver disease, complications as listed in
+Table \@ref(tab:tableone) and `r length(cn_labs)` laboratory measurements.
+
+```{r flowchart, results = "asis", echo = FALSE, fig.width = 5, fig.height = 5, fig.cap = "Flowchart Inclusion/Exclusion."}
+graph <- "
+digraph flowchart {
+    graph [splines = false];
+
+    // invisible nodes
+    node [shape = point, width = 0, height = 0];
+    iAR [style = 'invis'];
+
+    // visible nodes
+    node [shape = box];
+    A [label = '@@1'];
+    E [label = '@@3'];
+    R [label = '@@2'];
+
+    // edge statements
+    A -> iAR [arrowhead = none];
+    {rank = same; iAR -> E [minlen = 5]};
+    iAR -> R;
+}
+
+[1]: paste0('Total Records\\n(n = ', n_total, ')')
+[2]: paste0('Analysed Records\\n(n = ', n_analysed, ')')
+[3]: str_excluded
+"
+str_excluded <- paste0(
+    'Excluded (n = ', sum(n_excluded), ')\\\\l',
+    paste0(
+        '- ',
+        paste(
+            c(
+                'Younger than 18 years:',
+                'Liver Transplantation before visit:',
+                'Follow-up missing/invalid:'
+            ),
+            n_excluded
+        ),
+        collapse = '\\\\l'
+    ),
+    '\\\\l'
+)
+
+DiagrammeR::grViz(graph)
+```
+
+We excluded `r sum(n_excluded)` patients from our analysis
+(Figure \@ref(fig:flowchart)).
+`r if (n_excluded["Children"] == 1) "One was" else paste(n_excluded["Children"], "were")` younger than 18 years.
+`r n_excluded["LiverTransplantation"]` had a liver transplantation before and
+in `r n_excluded["LostFollowUp"]` cases the follow up data were missing or
+invalid.
+
+The Ethics Committee at the Leipzig University Faculty of
+Medicine approved the retrospective usage of the data for our study
+(reference number: 039/14ff).
+
+## MELD scores
+
+MELD and MELD-Na were calculated as described in
+@optn_policies2021 using the following formulas:
+$MELD = 10 * (0.957 \ln(creatinine [mg/dl]) + 0.378 \ln(bilirubin [mg/dl]) + 1.120 * \ln(INR) + 0.643)$,
+Creatinine, bilirubin and INR values lower than 1.0 mg/dl were set to 1.0 mg/dl.
+The maximum allowed creatinine was 4.0 mg/dl. If patients received dialysis,
+creatinine was set to 4.0 mg/dl.
+For MELD-Na MELD was calculated as above and recalculated if greater than 11 using:
+$MELD\text{-}Na = MELD_i + 1.32 * (137 - Na [mmol/l]) - (0.033 * MELD_i * (137 - Na [mmol/l]))$
+
+Serum sodium values lower than 125 mmol/l or higher than 137 mmol/l were set
+to 125 mmol/l and 137 mmol/l, respectively.
+
+The MELD-Plus7 risk score was calculated as described in @kartoun2017:
+$$
+\begin{aligned}
+L = 8.53499496 &+ \\
+    2.59679650 &* \log_{10}(1 + creatinine [mg/dl]) + \\
+    2.06503238 &* \log_{10}(1 + bilirubin [mg/dl]) + \\
+    2.99724802 &* \log_{10}(1 + INR) - \\
+    6.47834101 &* \log_{10}(1 + sodium [mmol/l]) - \\
+    6.34990436 &* \log_{10}(1 + albumin [g/l]) + \\
+    1.92811726 &* \log_{10}(1 + wbc [th/cumm]) + \\
+    0.04070442 &* age [years]
+\end{aligned}
+$$
+
+$$MELD\text{-}Plus = \frac{\exp(L)}{1 + \exp(L)}$$
+
+The MELD-Plus9 score was ignored because the length of stay information was not
+available for our cohort and the model itself wasn't superior to MELD-Plus7 in
+its original publication [@kartoun2017].
+
+## Machine learning
+
+```{r ml, include = FALSE}
+n_lrns <- nrow(bmrk_results$learners)
+#n_cv <- c(inner.folds
+#    outer = unlist(
+#        bmrk_results$resamplings$resampling[[1]]$param_set$get_values()
+#    )
+#)
+```
+
 # Results
@@
 
+In this retrospective analysis of `r n_analysed` consecutive patients,
+who were recruited during the evaluation process for liver transplantation at
+the University Hospital of Leipzig we compared
+`r n_lrns` different statistical and machine learning algorithms.
+
+...
+
+- higher mortality than predicted, compare @hernaez2020
+- adding sodium doesn't add much/any discrimination information
+    - as already seen in @kartoun2017
+- @kartoun2017 uses > 60 variables mostly clinical stuff, just few laboratory values, none of the clinical stuff remains
+- general low sample size thus simpler models/fewer predictors are preferred.
+- comparison of different predictors of MELD-Plus7 (+MELD-Plus9) vs our features
+    - 90 day mortality *post* discharge
+    - poorer performance in own validation set (lower and more uniform disease severity, also lower MELD)
+- machine learning algorithms worse than statistical learning (lower c-index, higher variance/instability)
+    - additive effects (favour statistical models)
+    - low sample size, for ML more than 200 events per predictor variable necessary [@ploeg2014]
+- national register required
+
+## Limitations
+
+- comparable mortality?
+- retrospective analysis
+- single center
+- unknown cause of death (death with vs of end-stage liver disease)
+- bias caused by LTx censorship?
+
 # References

### 2021-08-04 _targets.R
chore: rename reports code file

@@ -38,7 +38,7 @@ if (nchar(Sys.getenv("RUNLOCAL"))) {
 
 source("code/01-data.R")
 source("code/02-benchmarking.R")
-source("code/03-reports.R")
+source("code/09-reports.R")
 source("code/99-site.R")
 
 list(

### 2021-08-04 analysis/article.Rmd
feat: add table 1 and table observed vs expected mortality

@@
 library("targets")
+library("ameld")
+
+tar_load(bmrk_results)
 
 tar_load(raw_data)
-tar_load(bmrk_results)
+tar_load(labelled_meld_data)
+
+tar_load(tbl1)
+tar_load(tbl_observed_vs_expected_mortality)
 ```
@@
 Among them are age, sex, etiology of liver disease, complications as listed in
-Table \@ref(tab:tableone) and `r length(cn_labs)` laboratory measurements.
+Table \@ref(tab:tbl1) and `r length(cn_labs)` laboratory measurements.
 
@@
 
+## Baseline characteristics
+
+```{r tbl1, results = "asis", echo = FALSE, message = FALSE}
+tbl1
+```
+
+<!-- TODO:
+use `pattern` argument to provide "{n} ({p}%)" for categorial/dichotomos
+variables after this PR is merged
+https://github.com/ddsjoberg/gtsummary/pull/951
+-->
+
+```{r tbl1_helper_func, echo = FALSE}
+.tbl1 <- function(variable, column = "Overall", ...) {
+    id <- gtsummary::inline_text(
+        tbl1, variable = all_of(variable), column = "tbl_id1"
+    )
+    if (column == "Overall")
+        column <- "stat_0"
+    gtsummary::inline_text(
+        tbl1$tbls[[id]],
+        variable = all_of(variable), column = all_of(column), ...
+    )
+}
+```
+
+In this study we analysed a cohort of `r n_analysed` patients evaluated for
+liver transplantation (Tab. \@ref(tab:tbl1)).
+`r sprintf("%i (%s)", sum(raw_data$Sex == "male"), gtsummary::style_percent(mean(raw_data$Sex == "male"), symbol = TRUE))`
+of them were men.
+The median (IQR) follow-up times was `r .tbl1("DaysAtRisk")` days.
+Within the follow-up time `r .tbl1("LTx")` patients received a liver transplant
+and were censored for the analysis beyond this time point.
+The median (IQR) age of the cohort was `r .tbl1("Age")` years.
+
+Most of the patients presented with cirrhosis
+(`r .tbl1("Cirrhosis")`).
+The remaining patients had an acute liver failure
+(`r .tbl1("ALF")`) or
+a chronic liver insufficiency
+(`r .tbl1("CLI")`).
+At evaluation time the median (IQR) MELD and MELD-Na score were
+`r .tbl1("ScoreMeldUnos")` and
+`r .tbl1("ScoreMeldNaUnos")`, respectively.
+Within 90 days `r .tbl1("M90")` patients died.
+In nearly two third of all patients alcohol was the main cause
+(`r .tbl1("Ethyltoxic")`) of their end-stage liver disease.
+The remaining patients had viral infections
+(hepatitis B and C viruse `r .tbl1("HBV")` and `r .tbl1("HCV")`, respectively),
+autoimmune processes (autoimmune hepatitis `r .tbl1("AIH")`,
+primary sclerosing cirrhosis `r .tbl1("PSC")`),
+primary biliary cirrhosis `r .tbl1("PBC")` or unknown causes
+(non-alcoholic steatohepatitis `r .tbl1("NASH")` and
+cryptogenic hepatitis `r .tbl1("Cryptogenic")`).
+The most common complications among all patients were gastrointestinal
+bleedings in `r .tbl1("GIB")` patients, followed by
+hepatocellular carcinoma (HCC) in `r .tbl1("HCC")` and
+spontaneous bacterial peritonitis (SBP) in `r .tbl1("SBP")` patients.
+Just `r .tbl1("Dialysis")` patients required a renal dialysis.
+
+## MELD scores
+
+According to @wiesner2003 we divided our cohort into
+`r nlevels(labelled_meld_data$MeldCategory)` MELD-score risk
+categories, with `r enum(table(labelled_meld_data$MeldCategory))` patients,
+respectively.
+As depicted in Figure \@ref(fig:observed-vs-meld-mortality-plot) and Table
+\@ref(tab:tbl-observed-vs-expected-mortality) our observed 90 day mortality was
+`r paste(round(range(tbl_observed_vs_expected_mortality$SMR[-1]), 1), collapse = " to ")`
+times higher than the predicted one. Just in the group with the lowest MELD
+scores the mortality was much lower.
+
+```{r tbl-observed-vs-expected-mortality, results = "asis", echo = FALSE, message = FALSE}
+knitr::kable(
+    cbind(
+        row.names(tbl_observed_vs_expected_mortality),
+        tbl_observed_vs_expected_mortality
+    ),
+    row.names = FALSE,
+    col.names =
+        c(
+            "MELD category",
+            "Observed deaths (n)",
+            "Expected deaths (n)",
+            "Standardized mortality ratio (SMR)",
+            "Observed mortality (%)",
+            "Expected mortality (%)"
+        ),
+    caption = paste0(
+        "Observed vs MELD-expected 90 day mortality. ",
+        "MELD mortality values are taken from @wiesner2003. ",
+        "All patients censored before day 90 are ignored for ",
+        " the calculation of the MELD-expected deaths. ",
+        "SMR, Standardized mortality ratio = observed deaths/expected deaths."
+    ),
+    digits = 1
+)
+```
+
 # Discussion

### 2021-08-04 analysis/tables.Rmd
Revert "fix: working directory for targets"
This reverts commit cac14f630848b078c39aee7344862d418fcdcb52.
feat: add table 1 and table observed vs expected mortality

@@
 library("skimr")
-
-withr::with_dir(rprojroot::find_rstudio_root_file(), {
+library("zlog")
 tar_load(raw_data)
@@
 tar_load(zlog_data)
-})
 ```
@@
 ```
+
+# Raw Data (imputed) with Zlog Colors
+
+```{r tables_raw_zlog, results = "asis", echo = FALSE, message = FALSE}
+k <- kableExtra::kable_classic(
+    kableExtra::kbl(
+        raw_data,
+        digits = 2,
+        escape = FALSE
+    ),
+    "basic"
+)
+for (i in grep("_[CEQS]$", colnames(raw_data)))
+    k <- kableExtra::column_spec(k, i, background = zcol(zlog_data[[i]]))
+k |> kableExtra::scroll_box(width = "100%", height = "500px")
+```
+
+# Table One
+
+```{r tbl1, results = "asis", echo = FALSE, message = FALSE}
+tar_read(tbl1)
+```

### 2021-08-04 _targets.R

@@ -38,6 +38,7 @@ if (nchar(Sys.getenv("RUNLOCAL"))) {
 
 source("code/01-data.R")
 source("code/02-benchmarking.R")
+source("code/08-tables.R")
 source("code/09-reports.R")
 source("code/99-site.R")
 
@@ -46,6 +47,7 @@ list(
    tg_tasks,
    tg_benchmarks,
    tg_benchmark_results,
+   tg_tables,
    tg_reports,
    tg_site
 )

### 2021-08-04 code/01-data.R

@@ -1,3 +1,60 @@
+.add_meld_columns <- function(x) {
+    x$ScoreMeld <- meld(
+        creatinine = as_metric(x$CRE_S, "creatinine"),
+        bilirubin = as_metric(x$BILI_S, "bilirubin"),
+        inr = x$INR_C,
+        dialysis = x$Dialysis,
+        cause = ifelse(x$Ethyltoxic == 1, "ethyltoxisch", "other")
+    )
+    x$ScoreMeldUnos <- meld(
+        creatinine = as_metric(x$CRE_S, "creatinine"),
+        bilirubin = as_metric(x$BILI_S, "bilirubin"),
+        inr = x$INR_C,
+        dialysis = x$Dialysis,
+        cause = "UNOS"
+    )
+    x$ScoreMeldNa <- meld_na(
+        creatinine = as_metric(x$CRE_S, "creatinine"),
+        bilirubin = as_metric(x$BILI_S, "bilirubin"),
+        inr = x$INR_C,
+        sodium = x$NA_S,
+        dialysis = x$Dialysis,
+        cause = ifelse(x$Ethyltoxic == 1, "ethyltoxisch", "other")
+    )
+    x$ScoreMeldNaUnos <- meld_na(
+        creatinine = as_metric(x$CRE_S, "creatinine"),
+        bilirubin = as_metric(x$BILI_S, "bilirubin"),
+        inr = x$INR_C,
+        sodium = x$NA_S,
+        dialysis = x$Dialysis,
+        type = "UNOS"
+    )
+    x$ProbMeldPlus7 <- meld_plus7(
+        creatinine = as_metric(x$CRE_S, "creatinine"),
+        bilirubin = as_metric(x$BILI_S, "bilirubin"),
+        inr = x$INR_C,
+        sodium = x$NA_S,
+        albumin = x$ALB_S / 10,
+        wbc = x$B_WBC_E,
+        age = x$Age
+    )
+    x$MeldCategory <- cut(
+        x$ScoreMeldUnos,
+        breaks = c(-Inf, seq(10, 40, by=10), Inf),
+        labels = c(
+            paste0(
+                "[", floor(min(x$ScoreMeldUnos, na.rm = TRUE)), ",9]"
+            ),
+            "[10,20)", "[20,30)", "[30,40)",
+            paste0(
+                "[40,", ceiling(max(x$ScoreMeldUnos, na.rm = TRUE)), ")"
+            )
+        ),
+        right = FALSE
+    )
+    x
+}
+
[…]

### 2021-08-04 code/08-tables.R

@@ -1,0 +1,199 @@
+tg_tables <- list(
+    tar_target(tbl1, {
+        theme_gtsummary_journal(journal = "jama")
+        theme_gtsummary_compact()
+        tbGeneral <- labelled_meld_data |>
+            select(
+                Sex,
+                DaysAtRisk, Age, LTx
+            ) |>
+            tbl_summary(
+                by = Sex,
+                missing_text = "(Missing)"
+            ) |>
+            add_overall()
+
+        tbLabs <- labelled_meld_data |>
+            select(
+                Sex,
+                BILI_S, CYSC_S, CRE_S, INR_C, NA_S,
+                B_WBC_E, CRE_S, IL6_S,
+                ALB_S, PROT_S, CHOLG_S,
+                ALAT_S, ASAT_S
+            ) |>
+            tbl_summary(
+                by = Sex,
+                missing_text = "(Missing)"
+            ) |>
+            add_overall()
+
+        tbMeld <- labelled_meld_data |>
+            select(
+               Sex,
+               ScoreMeldUnos,
+               MeldCategory,
+               ScoreMeldNaUnos, ProbMeldPlus7,
+            ) |>
+            tbl_summary(
+                by = Sex,
+                missing_text = "(Missing)"
+            ) |>
+            add_overall()
+
+        tbEntities <- labelled_meld_data |>
+            select(
+               Sex,
+               Cirrhosis, ALF, CLI
+            ) |>
+            tbl_summary(
+                by = Sex,
+                missing_text = "(Missing)"
+            ) |>
+            add_overall()
+
+        tbEtiology <- labelled_meld_data |>
+            select(
+               Sex,
+               Ethyltoxic, HBV, HCV, AIH, PBC, PSC, NASH, Cryptogenic
+            ) |>
+            tbl_summary(
+                by = Sex,
+                missing_text = "(Missing)"
+            ) |>
+            add_overall()
+
+        tbComplications <- labelled_meld_data |>
+            select(
+               Sex,
+               Dialysis, GIB, HCC, SBP
+            ) |>
+            tbl_summary(
+                by = Sex,
+                missing_text = "(Missing)"
+            ) |>
+            add_overall()
+
+        ## generate mortality table stub
+        ## (tbl_survfit doesn't support (cumulative) events)
+        tbMortality <- labelled_meld_data |>
+            select(
+               Sex,
+               Age
+            ) |>
+            tbl_summary(
+                by = Sex,
[…]

### 2021-08-07 code/01-data.R
fix: MELD calculation (cause)

@@ -4,7 +4,7 @@
         bilirubin = as_metric(x$BILI_S, "bilirubin"),
         inr = x$INR_C,
         dialysis = x$Dialysis,
-        cause = ifelse(x$Ethyltoxic == 1, "ethyltoxisch", "other")
+        cause = ifelse(x$Ethyltoxic == 1, "ethyltoxic", "other")
     )
     x$ScoreMeldUnos <- meld(
         creatinine = as_metric(x$CRE_S, "creatinine"),
@@ -19,7 +19,7 @@
         inr = x$INR_C,
         sodium = x$NA_S,
         dialysis = x$Dialysis,
-        cause = ifelse(x$Ethyltoxic == 1, "ethyltoxisch", "other")
+        cause = ifelse(x$Ethyltoxic == 1, "ethyltoxic", "other")
     )
     x$ScoreMeldNaUnos <- meld_na(
         creatinine = as_metric(x$CRE_S, "creatinine"),
@@ -191,7 +191,9 @@ tg_data <- list(
     tar_target(zlog_data, {
             zlog_data <- meld_data
             iconvert <- colnames(zlog_data) %in%
-                c("Age", "Sex", grep("_[SECQ]$", colnames(imp_data), value = TRUE))
+                c("Age", "Sex",
+                  grep("_[SECQ]$", colnames(imp_data), value = TRUE)
+                )
             zlog_data[iconvert] <- zlog_df(imp_data[iconvert], ref_data)
             zlog_data
         },

### 2021-08-07 analysis/article.Rmd
fix: observed vs expected mortality table

@@
 \@ref(tab:tbl-observed-vs-expected-mortality) our observed 90 day mortality was
-`r paste(round(range(tbl_observed_vs_expected_mortality$SMR[-1]), 1), collapse = " to ")`
+`r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
 times higher than the predicted one. Just in the group with the lowest MELD
@@
 ```{r tbl-observed-vs-expected-mortality, results = "asis", echo = FALSE, message = FALSE}
+tbl_observed_vs_expected_mortality$ObservedMortality <-
+    tbl_observed_vs_expected_mortality$ObservedMortality * 100
+tbl_observed_vs_expected_mortality$ExpectedMortality <-
+    tbl_observed_vs_expected_mortality$ExpectedMortality * 100
 knitr::kable(

### 2021-08-07 code/02-benchmarking.R
refactor: use random instead of grid search with incredible low cross val (HPC currently unavailable)

@@ -158,8 +158,10 @@ ps_svmregression$trafo <- function(x, param_set) {
 }
 
 ## resampling
-inner_cv <- rsmp("cv", folds = 5L)
-outer_cv <- rsmp("repeated_cv", folds = 3L, repeats = 5L)
+#inner_cv <- rsmp("cv", folds = 5L)
+#outer_cv <- rsmp("repeated_cv", folds = 3L, repeats = 10L)
+inner_cv <- rsmp("cv", folds = 3L)
+outer_cv <- rsmp("repeated_cv", folds = 3L, repeats = 1L)
 
 ## tuning
 autoTuner <- function(lrn, prms, rsmp = inner_cv)AutoTuner$new(
@@ -167,8 +169,8 @@ autoTuner <- function(lrn, prms, rsmp = inner_cv)AutoTuner$new(
     search_space = prms,
     resampling = rsmp,
     measure = msr("surv.cindex"),
-    tuner = tnr("grid_search", resolution = 20L, batch_size = 2L),
-    terminator = trm("stagnation", iters = 3L, threshold = 1e-3)
+    tuner = tnr("random_search", batch_size = 2L),
+    terminator = trm("evals", n_evals = 20)
 )
 
 at_lasso <- autoTuner(lrn_lasso, ps_lasso)

### 2021-08-07 scripts/Rscript.sh
fix: set python path and pycox data dir for survivalmodels

@@ -16,6 +16,8 @@ else # local
 fi
 
 SINGULARITYENV_LOCPATH=/lib/locale/2.31/ \
+    SINGULARITYENV_RETICULATE_PYTHON=/bin/python3.8 \
+    SINGULARITYENV_PYCOX_DATA_DIR=${TMPDIR:-/tmp} \
     SINGULARITYENV_TAR_MAKE_REPORTER="timestamp" \
     SINGULARITYENV_TZ="UTC" \
     SINGULARITY_BIND="${BIND_PATH}" \

### 2021-08-07 code/08-tables.R
fix: gtsummary compact theme

@@ -1,7 +1,5 @@
 tg_tables <- list(
     tar_target(tbl1, {
-        theme_gtsummary_journal(journal = "jama")
-        theme_gtsummary_compact()
         tbGeneral <- labelled_meld_data |>
             select(
                 Sex,

### 2021-08-07 analysis/tables.Rmd
fix: gtsummary compact theme
fix: use imp_data instead of raw_data; add popover

--- 2021-08-02 analysis/tables.Rmd
@@
 
+<script>
+$(document).ready(function(){
+    $('[data-toggle="popover"]').popover();
+});
+
+</script>
+
 ```{r histograms_setup, message = FALSE}
@@
 ```
+
+# Raw Data (imputed) with Zlog Colors
+
+```{r tables_raw_zlog, results = "asis", echo = FALSE, message = FALSE}
+k <- kableExtra::kable_classic(
+    kableExtra::kbl(
+        imp_data,
+        digits = 2,
+        escape = FALSE
+    ),
+    "basic"
+)
+for (i in grep("_[CEQS]$", colnames(imp_data)))
+    k <- kableExtra::column_spec(
+            k, i,
+            background = zcol(zlog_data[[i]]),
+            popover = paste0("zlog: ", round(zlog_data[[i]]), 2)
+    )
+k |> kableExtra::scroll_box(width = "100%", height = "500px")
+```
+
+# Table One
+
+```{r tbl1, results = "asis", echo = FALSE, message = FALSE}
+library("gtsummary")
+theme_gtsummary_journal(journal = "jama")
+theme_gtsummary_compact()
+tar_read(tbl1)
+```

### 2021-08-07 analysis/tables.Rmd
fix: zlog rounding

@@
             background = zcol(zlog_data[[i]]),
-            popover = paste0("zlog: ", round(zlog_data[[i]]), 2)
+            popover = paste0("zlog: ", round(zlog_data[[i]], 2))
     )

### 2021-08-08 _targets.R
refactor: split mlr3 settings and use dynamic branching for benchmarking

@@ -20,30 +20,33 @@ login <- future::tweak(
 slurm <- future::tweak(
     future.batchtools::batchtools_slurm,
     template = "scripts/slurm_batchtools.tmpl",
-    workers = 8L,
+    workers = 16L,
     resources = list(
         logfile = "logs/job%s.log",
         partition = "batch,compute,pinky,snowball",
-        ncpu = 8L,
+        ncpu = 16L,
         mcpu = 16L * 1024L,
         walltime = 4L * 60L * 60L
     )
 )
 
 if (nchar(Sys.getenv("RUNLOCAL"))) {
-    future::plan(callr, workers = 3)
+    future::plan(callr, workers = 3L)
 } else {
     future::plan(list(login, slurm, callr))
 }
 
 source("code/01-data.R")
-source("code/02-benchmarking.R")
+source("code/02-mlr-tasks.R")
+source("code/03-mlr-learners.R")
+source("code/04-mlr-benchmarking.R")
 source("code/08-tables.R")
 source("code/09-reports.R")
 source("code/99-site.R")
 
 list(
    tg_data,
+   tg_lrns,
    tg_tasks,
    tg_benchmarks,
    tg_benchmark_results,

### 2021-08-08 code/02-benchmarking.R

@@ -1,246 +1,0 @@
-suppressPackageStartupMessages(library("mlr3"))
-suppressPackageStartupMessages(library("mlr3extralearners"))
-suppressPackageStartupMessages(library("mlr3learners"))
-suppressPackageStartupMessages(library("mlr3proba"))
-suppressPackageStartupMessages(library("mlr3tuning"))
-suppressPackageStartupMessages(library("paradox"))
-
-## learners
-lrn_cox <- lrn("surv.coxph", id = "cox")
-
-lrn_lasso <- lrn("surv.cv_glmnet", id = "lasso", alpha = 1L)
-
-lrn_ridge <- lrn("surv.cv_glmnet", id = "ridge", alpha = 0L)
-
-lrn_elasticnet <- lrn("surv.cv_glmnet", id = "elasticnet")
-
-lrn_penlasso <- lrn("surv.penalized", id = "penlasso",
-    lambda2 = 0, standardize = TRUE, trace = FALSE
-)
-
-lrn_penridge <- lrn("surv.penalized", id = "penridge",
-    lambda1 = 0, standardize = TRUE, trace = FALSE
-)
-
-lrn_ranger <- lrn("surv.ranger", id = "ranger",
-    # ?ranger, 3 for survival
-    min.node.size = 3L
-)
-
-lrn_rfsrc <- lrn("surv.rfsrc",
-    # ?randomForestSRC, nsplit = 0, randomized splitting
-    nsplit = 0
-)
-
-lrn_xgboost <- lrn("surv.xgboost", max_depth = 6)
-
-lrn_svmregression <- lrn("surv.svm", id = "svm-regression",
-    type = "regression",
-    gamma.mu = 0 # would be trained later, is required by ctor
-)
-
-lrn_svmvanbelle1 <- lrn("surv.svm", id = "svm-vanbelle1",
-    type = "vanbelle1",
-    diff.meth = "makediff3",
-    kernel = "rbf_kernel",
-    opt.meth = "quadprog",
-    gamma.mu = 0 # would be trained later, is required by ctor
-)
-
-## params
-ps_lasso <- ParamSet$new(
-    params = list(
-        ParamDbl$new("s", lower = 0L, upper = 1L)
-    )
-)
-
-ps_ridge <- ParamSet$new(
-    params = list(
-        ParamDbl$new("s", lower = 0L, upper = 1L)
-    )
-)
-
-ps_elasticnet <- ParamSet$new(
[…]

### 2021-08-08 code/02-mlr-tasks.R

@@ -1,0 +1,20 @@
+suppressPackageStartupMessages(library("mlr3"))
+suppressPackageStartupMessages(library("mlr3proba"))
+
+tg_tasks <- list(
+    tar_target(
+        zlog_task, {
+            ## exclude precalculated scores
+            zd <- zlog_data[!grepl("Meld", colnames(zlog_data))]
+            zd$Sex <- as.integer(zd$Sex == "male") # female: 0, male: 1
+
+            TaskSurv$new(
+                id = "eldd",
+                backend = zd,
+                time = "DaysAtRisk",
+                event = "Deceased"
+            )
+        },
+        packages = c("mlr3", "mlr3proba")
+    )
+)

### 2021-08-08 code/03-mlr-learners.R

@@ -1,0 +1,160 @@
+suppressPackageStartupMessages(library("mlr3"))
+suppressPackageStartupMessages(library("mlr3extralearners"))
+suppressPackageStartupMessages(library("mlr3learners"))
+suppressPackageStartupMessages(library("mlr3proba"))
+suppressPackageStartupMessages(library("mlr3tuning"))
+suppressPackageStartupMessages(library("paradox"))
+
+tg_lrns <- list(
+    ## while these are small targets that don't need any processing time
+    ## we want to rerun the benchmarking when anything changes
+    tar_target(crossval,
+        list(
+            inner = rsmp("cv", folds = 3L),
+            outer = rsmp("repeated_cv", folds = 3L, repeats = 1L)
+        )
+    ),
+    tar_target(learners, {
+        l <- list(
+            cox = list(
+                lrn = lrn("surv.coxph", id = "cox")
+            ),
+            lasso = list(
+                lrn = lrn("surv.cv_glmnet", id = "lasso", alpha = 1L),
+                ps = ps(s = p_dbl(lower = 0, upper = 1))
+            ),
+            ridge = list(
+                lrn = lrn("surv.cv_glmnet", id = "ridge", alpha = 0L),
+                ps = ps(s = p_dbl(lower = 0, upper = 1))
+            ),
+            elasticnet = list(
+                lrn = lrn("surv.cv_glmnet", id = "elasticnet"),
+                ps = ps(
+                    alpha = p_dbl(lower = 0, upper = 1),
+                    s = p_dbl(lower = 0, upper = 1)
+                )
+            ),
+            penlasso = list(
+                lrn = lrn(
+                    "surv.penalized", id = "penlasso",
+                    lambda2 = 0, standardize = TRUE, trace = FALSE
+                ),
+                ps = ps(
+                    lambda1 = p_dbl(lower = 1, upper = 100)
+                )
+            ),
+            penridge = list(
+                lrn = lrn(
+                    "surv.penalized", id = "penridge",
+                    lambda1 = 0, standardize = TRUE, trace = FALSE
+                ),
+                ps = ps(
+                    lambda2 = p_dbl(lower = 1, upper = 100)
+                )
+            ),
+            ranger = list(
+                lrn = lrn(
+                    "surv.ranger", id = "ranger",
+                    # ?ranger, 3 for survival
+                    min.node.size = 3L
+                ),
+                ps = ps(
[…]

### 2021-08-08 code/04-mlr-benchmarking.R

@@ -1,0 +1,30 @@
+suppressPackageStartupMessages(library("mlr3"))
+suppressPackageStartupMessages(library("mlr3tuning"))
+
+tg_benchmarks <- tar_target(
+    bmrks,
+    benchmark(
+        benchmark_grid(
+            task = zlog_task,
+            learner = learners,
+            resampling = crossval$outer
+        ),
+        store_models = FALSE,
+        store_backends = FALSE
+    ),
+    pattern = map(learners),
+    iteration = "list",
+    packages = c(
+        "mlr3", "mlr3tuning", "mlr3misc", "mlr3proba", "mlr3learners",
+        "ranger"
+    ),
+    deployment = "worker"
+)
+
+tg_benchmark_results <- tar_combine(
+    bmrk_results,
+    tg_benchmarks,
+    command = c(!!!.x),
+    packages = "mlr3",
+    deployment = "worker"
+)

### 2021-09-14 _targets.R
chore: adapt to new slurm on hpc

@@ -20,12 +20,12 @@ login <- future::tweak(
 slurm <- future::tweak(
     future.batchtools::batchtools_slurm,
     template = "scripts/slurm_batchtools.tmpl",
-    workers = 16L,
+    workers = 8L,
     resources = list(
         logfile = "logs/job%s.log",
-        partition = "batch,compute,pinky,snowball",
+        partition = "batch,compute,snowball",
         ncpu = 16L,
-        mcpu = 16L * 1024L,
+        mcpu = 8L * 1024L,
         walltime = 4L * 60L * 60L
     )
 )

### 2021-09-14 _targets.R
chore: add loggers

@@ -5,10 +5,17 @@ library("future")
 library("future.batchtools")
 library("future.callr")
 
+library("lgr")
+
 #Sys.setenv(RUNLOCAL = 1)
 Sys.setenv(DEBUGME = "batchtools")
-lgr::get_logger("mlr3")$set_threshold("info")
-lgr::get_logger("bbotk")$set_threshold("info")
+tf <- rprojroot::find_rstudio_root_file("logs", "lgr.log")
+lgr$add_appender(AppenderFile$new(tf), name = "file")
+lgr$set_threshold("all")
+get_logger("mlr3")$set_threshold("all")
+get_logger("bbotk")$set_threshold("all")
+get_logger("mlr3")$add_appender(AppenderFile$new(tf), name = "file")
+get_logger("bbotk")$add_appender(AppenderFile$new(tf), name = "file")
 
 login <- future::tweak(
     future::remote,
@@ -31,7 +38,7 @@ slurm <- future::tweak(
 )
 
 if (nchar(Sys.getenv("RUNLOCAL"))) {
-    future::plan(callr, workers = 3L)
+    future::plan(callr, workers = 7L)
 } else {
     future::plan(list(login, slurm, callr))
 }

### 2021-09-14 scripts/Rscript.sh
fix: adapt to new slurm on hpc

@@ -3,13 +3,13 @@
 BIND_PATH="/etc/passwd,/etc/group,/var/run"
 
 # on HPC?
-MODULEPATH=/usr/share/Modules/init/bash
-if [ -f "${MODULEPATH}" ] ; then
+MODULESPATH=/usr/share/Modules/init/bash
+if [ -f "${MODULESPATH}" ] ; then
     SINGULARITYVER=3.4.2
-    . "${MODULEPATH}"
-    module load singularity/${SINGULARITYVER} slurm
+    . "${MODULESPATH}"
+    module load singularity/${SINGULARITYVER}
     SINGULARITYPATH=/opt/singularity-${SINGULARITYVER}/bin/singularity
-    BIND_PATH="${BIND_PATH},/etc/slurm"
+    BIND_PATH="${BIND_PATH},/etc/slurm/slurm.conf"
 else # local
     BIND_PATH="${BIND_PATH},${SSH_AUTH_SOCK}"
     SINGULARITYPATH=/usr/bin/singularity
@@ -21,5 +21,6 @@ SINGULARITYENV_LOCPATH=/lib/locale/2.31/ \
     SINGULARITYENV_TAR_MAKE_REPORTER="timestamp" \
     SINGULARITYENV_TZ="UTC" \
     SINGULARITY_BIND="${BIND_PATH}" \
-    ${SINGULARITYPATH} run container/ampel-leipzig-meld.squashfs "${@}"
+    ${SINGULARITYPATH} run container/ampel-leipzig-meld.squashfs \
+    "--vanilla ${@}"
 exit 0

### 2021-09-14 analysis/benchmarks.Rmd
feat: first nnet tests

--- 2021-07-14 analysis/benchmarks.Rmd
@@
 
+# Overview
+
 ```{r benchmarks_plot_c_harrell}
@@
 ```
+
+# Aggregated performance
+
+```{r benchmarks_aggr_plot_c_harrell}
+autoplot(bmrk_aggr, type = "box", meas = "harrell_c") +
+    geom_boxplot(aes(fill = learner_id)) +
+    geom_jitter(position = position_jitter(0.2)) +
+    scale_fill_viridis(discrete = TRUE) +
+    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+```
+
+# Critical difference
+
+```{r benchmarks_aggr_critdiff_c_harrell}
+autoplot(
+    bmrk_aggr, type = "cd", meas = "harrell_c", minimize = FALSE, style = 2,
+    p.value = 0.1
+) + scale_color_viridis(discrete = TRUE)
+```

### 2021-09-14 code/03-mlr-learners.R

@@ -11,8 +11,9 @@ tg_lrns <- list(
     tar_target(crossval,
         list(
             inner = rsmp("cv", folds = 3L),
-            outer = rsmp("repeated_cv", folds = 3L, repeats = 1L)
-        )
+            outer = rsmp("repeated_cv", folds = 3L, repeats = 2L)
+        ),
+        deployment = "main"
     ),
     tar_target(learners, {
         l <- list(
@@ -78,7 +79,7 @@ tg_lrns <- list(
                     # ?randomForestSRC ceiling(sqrt(length(variables))) = 7
                     # in our case
                     mtry = p_int(lower = 4L, upper = 8L),
-                    splitrule = p_fct(c("logrank", "C"))
+                    splitrule = p_fct(c("logrank", "logrankscore"))
                 )
             ),
             xgboost = list(
@@ -141,6 +142,58 @@ tg_lrns <- list(
                     }
                     p
                 }
+            ),
+            # initial settings as described in
+            # https://towardsdatascience.com/neural-networks-for-survival-analysis-in-r-1e0421584ab
+            coxtime = list(
+                lrn = lrn(
+                    "surv.coxtime", id = "coxtime",
+                    frac = 1/3,
+                    early_stopping = TRUE, epochs = 1, optimizer = "adam"
+                ),
+                ps = {
+                    p <- ps(
+                        dropout = p_dbl(lower = 0.1, upper = 0.7, default = 0.5),
+                        weight_decay = p_dbl(lower = 0, upper = 0.4, default = 0.1),
+                        learning_rate = p_dbl(lower = 0, upper = 0.3, default = 0.1),
+                        ## nodes in a layer
+                        nodes = p_int(lower = 2L, upper = 32L),
+                        ## number of hidden layers
+                        k = p_int(lower = 1L, upper = 4L)
+                    )
+                    p$trafo <- function(x, param_set) {
+                        ## use same number of nodes in each layer
+                        x$num_nodes = rep(x$nodes, x$k)
+                        x$nodes <- x$k <- NULL
+                        x
+                    }
+                    p
+                }
+            ),
+            deepsurv = list(
+                lrn = lrn(
+                    "surv.deepsurv", id = "deepsurv",
+                    frac = 1/3,
+                    early_stopping = TRUE, epochs = 1, optimizer = "adam"
+                ),
+                ps = {
+                    p <- ps(
+                        dropout = p_dbl(lower = 0.1, upper = 0.7, default = 0.5),
[…]

### 2021-09-14 code/04-mlr-benchmarking.R

@@ -1,30 +1,44 @@
 suppressPackageStartupMessages(library("mlr3"))
+suppressPackageStartupMessages(library("mlr3benchmark"))
+suppressPackageStartupMessages(library("mlr3proba"))
 suppressPackageStartupMessages(library("mlr3tuning"))
 
 tg_benchmarks <- tar_target(
     bmrks,
     benchmark(
         benchmark_grid(
-            task = zlog_task,
+            task = tasks,
             learner = learners,
             resampling = crossval$outer
         ),
         store_models = FALSE,
-        store_backends = FALSE
+        store_backends = TRUE
     ),
-    pattern = map(learners),
+    pattern = cross(tasks, learners),
     iteration = "list",
     packages = c(
-        "mlr3", "mlr3tuning", "mlr3misc", "mlr3proba", "mlr3learners",
-        "ranger"
+        "mlr3", "mlr3tuning", "mlr3misc", "mlr3proba", "mlr3learners"
     ),
     deployment = "worker"
 )
 
-tg_benchmark_results <- tar_combine(
-    bmrk_results,
-    tg_benchmarks,
-    command = c(!!!.x),
-    packages = "mlr3",
-    deployment = "worker"
+tg_benchmark_results <- list(
+    tar_target(
+        bmrk_results,
+        do.call(c, bmrks),
+        packages = "mlr3",
+        deployment = "worker"
+    ),
+    tar_target(
+        bmrk_aggr,
+        as.BenchmarkAggr(
+            bmrk_results,
+            measures = list(
+                msr("surv.cindex"),
+                msr("surv.cindex", weight_meth = "G2")
+            )
+        ),
+        packages = c("mlr3", "mlr3proba", "mlr3benchmark"),
+        deployment = "worker"
+    )
 )

### 2021-09-14 code/01-data.R
feat: create ln_data task

@@ -188,6 +188,14 @@ tg_data <- list(
         packages = "ameld",
         deployment = "main"
     ),
+    tar_target(ln_data, {
+            l <- meld_data
+            iconvert <- grep("_[CEQS]$", colnames(imp_data))
+            l[iconvert] <- lapply(l[iconvert], function(y)log(y + 1))
+            l
+        },
+        deployment = "main"
+    ),
     tar_target(zlog_data, {
             zlog_data <- meld_data
             iconvert <- colnames(zlog_data) %in%

### 2021-09-14 code/02-mlr-tasks.R

@@ -1,20 +1,27 @@
 suppressPackageStartupMessages(library("mlr3"))
 suppressPackageStartupMessages(library("mlr3proba"))
 
+.create_tsk <- function(d, id) {
+    ## exclude precalculated scores
+    d <- d[!grepl("Meld", colnames(d))]
+    d$Sex <- as.integer(d$Sex == "male") # female: 0, male: 1
+
+    TaskSurv$new(
+        id = id,
+        backend = d,
+        time = "DaysAtRisk",
+        event = "Deceased"
+    )
+}
+
 tg_tasks <- list(
     tar_target(
-        zlog_task, {
-            ## exclude precalculated scores
-            zd <- zlog_data[!grepl("Meld", colnames(zlog_data))]
-            zd$Sex <- as.integer(zd$Sex == "male") # female: 0, male: 1
-
-            TaskSurv$new(
-                id = "eldd",
-                backend = zd,
-                time = "DaysAtRisk",
-                event = "Deceased"
-            )
-        },
+        tasks,
+        mapply(
+            FUN = .create_tsk,
+            d = list(zlog_data, ln_data),
+            id = paste(c("zlog", "ln"), "eldd", sep = "_")
+        ),
         packages = c("mlr3", "mlr3proba")
     )
 )

### 2021-09-21 analysis/benchmarks.Rmd
fix: load bmrk_aggr and set p.value

@@
 tar_load(bmrk_results)
+tar_load(bmrk_aggr)
+
+p.value <- 0.2
 ```
@@
 
+Aggregated performance across all `nrow(bmrk_results$tasks)` datasets.
+
 ```{r benchmarks_aggr_plot_c_harrell}
@@
 
+# Pairwise comparison
+
+*p*-value is set to `r p.value`
+(otherwise an error is thrown because of insignificant results).
+
+```{r benchmarks_aggr_pairwise_harrell_c}
+autoplot(
+    bmrk_aggr, type = "fn", meas = "harrell_c", p.value = p.value,
+    col = palette.colors(2L)[2L]
+)
+```
+
 # Critical difference
+
+*p*-value is set to `r p.value`
+(otherwise an error is thrown because of insignificant results).
 
@@
 autoplot(
-    bmrk_aggr, type = "cd", meas = "harrell_c", minimize = FALSE, style = 2,
-    p.value = 0.1
+    bmrk_aggr, type = "cd", meas = "harrell_c", style = 2, p.value = p.value,
+    minimize = FALSE
 ) + scale_color_viridis(discrete = TRUE)

### 2021-10-15 _targets.R
refactor: increase hpc requirements

@@ -27,13 +27,12 @@ login <- future::tweak(
 slurm <- future::tweak(
     future.batchtools::batchtools_slurm,
     template = "scripts/slurm_batchtools.tmpl",
-    workers = 8L,
+    workers = 16L,
     resources = list(
-        logfile = "logs/job%s.log",
         partition = "batch,compute,snowball",
-        ncpu = 16L,
-        mcpu = 8L * 1024L,
-        walltime = 24L * 60L * 60L # seconds
+        ncpu = 64,
+        mcpu = 2L * 1024L,
+        walltime = 48L * 60L * 60L # seconds
     )
 )
 

### 2021-10-15 code/03-mlr-learners.R
refactor: shrink range for gamma in svm

@@ -124,8 +124,8 @@ tg_lrns <- list(
                 ),
                 ps = {
                     p <- ps(
-                        gamma.mu = p_dbl(lower = -5, upper = 5, default = 0),
-                        kernel.pars = p_dbl(lower = -5, upper = 5, default = 0)
+                        gamma.mu = p_dbl(lower = -3, upper = 3, default = 0),
+                        kernel.pars = p_dbl(lower = -3, upper = 3, default = 0)
                     )
                     p$trafo <- function(x, param_set) {
                         x$gamma.mu <- 2^x$gamma.mu

### 2021-10-15 code/03-mlr-learners.R
refactor: set deepsurv tuning params to the one mentioned in original publication

@@ -155,58 +155,30 @@ tg_lrns <- list(
                     p
                 }
             ),
-            # initial settings as described in
-            # https://towardsdatascience.com/neural-networks-for-survival-analysis-in-r-1e0421584ab
-            coxtime = list(
-                lrn = lrn(
-                    "surv.coxtime", id = "coxtime",
-                    frac = 1/3,
-                    early_stopping = TRUE, epochs = 10, optimizer = "adam"
-                ),
-                ps = {
-                    p <- ps(
-                        dropout =
-                            p_dbl(lower = 0.1, upper = 0.7, default = 0.5),
-                        weight_decay =
-                            p_dbl(lower = 0, upper = 0.4, default = 0.1),
-                        learning_rate =
-                            p_dbl(lower = 0, upper = 0.3, default = 0.1),
-                        ## nodes in a layer
-                        nodes = p_int(lower = 2L, upper = 32L),
-                        ## number of hidden layers
-                        k = p_int(lower = 1L, upper = 4L)
-                    )
-                    p$trafo <- function(x, param_set) {
-                        ## use same number of nodes in each layer
-                        x$num_nodes = rep(x$nodes, x$k)
-                        x$nodes <- x$k <- NULL
-                        x
-                    }
-                    p
-                }
-            ),
             deepsurv = list(
                 lrn = lrn(
                     "surv.deepsurv", id = "deepsurv",
-                    frac = 1/3,
-                    early_stopping = TRUE, epochs = 10, optimizer = "adam"
+                    activation = "leakyrelu",
+                    frac = 0.3, early_stopping = TRUE,
+                    epochs = 100, optimizer = "adam"
                 ),
                 ps = {
                     p <- ps(
                         dropout =
-                            p_dbl(lower = 0.1, upper = 0.7, default = 0.5),
+                            p_dbl(lower = 0, upper = 0.7, default = 0.5),
+                        ## 0.4, 0.2, 0.1, 0.05, 0.02, 0.01, 0
                         weight_decay =
-                            p_dbl(lower = 0, upper = 0.4, default = 0.1),
+                            p_int(lower = 0, upper = 6, default = 2),
                         learning_rate =
-                            p_dbl(lower = 0, upper = 0.3, default = 0.1),
-                        ## nodes in a layer
-                        nodes = p_int(lower = 2L, upper = 32L),
-                        ## number of hidden layers
[…]

### 2021-10-15 code/03-mlr-learners.R
refactor: reset tuner/terminator and preprocessing pipeline

@@ -186,6 +186,21 @@ tg_lrns <- list(
                 }
             )
         )
+        ## tuner
+        tuner <-
+            tnr("grid_search", resolution = 4L, batch_size = 4L)
+            #tnr("random_search", batch_size = 5L)
+
+        ## terminator
+        terminator <- trm("combo", list(
+                trm("evals", n_evals = 50L),
+                trm("run_time", secs = 90L),
+                trm("stagnation", iters = 3L, threshold = 1e-3),
+                trm("stagnation_batch", n = 1L, threshold = 1e-3)
+            ),
+            any = TRUE
+        )
+
         ## autotuner
         l <- lapply(l, function(ll) {
             if (is.null(ll$ps))
@@ -196,27 +211,19 @@ tg_lrns <- list(
                     search_space = ll$ps,
                     resampling = crossval$inner,
                     measure = msr("surv.cindex"),
-                    tuner =
-                        tnr("grid_search", resolution = 20L, batch_size = 2L),
-                    terminator =
-                        trm("stagnation", iters = 3L, threshold = 1e-3)
+                    tuner = tuner,
+                    terminator = terminator
                 )
         })
 
-        ## preprocessing (is part of the crossval now)
-        # TODO: add scaling to ln_data task (not to zlog_data)
-        #po_nop <- po("nop", "nothing")
-        #po_scale <- po("scale")
-        #choices <- c("nothing", "scale")
-        #g <-
-        #    po("removeconstants") %>>%
-        #        po("branch", choices) %>>%
-        #            gunion(list(po_nop, po_scale)) %>>%
-        #        po("unbranch", choices)
-
-        #l <- lapply(l, function(ll) {
-        #    as_learner(g %>>% po("learner", ll))
-        #})
+        ### preprocessing (is part of the crossval now)
+        l <- c(
+            l,
+            lapply(
+                l,
+                function(ll)as_learner(po("scale") %>>% po("learner", ll))
+            )
+        )
         l
     },
     iteration = "list",

### 2021-10-15 code/03-mlr-learners.R
refactor: use incredibly low cv folds b/c hpc is down

@@ -11,8 +11,8 @@ tg_lrns <- list(
     ## we want to rerun the benchmarking when anything changes
     tar_target(crossval,
         list(
-            inner = rsmp("cv", folds = 5L),
-            outer = rsmp("repeated_cv", folds = 3L, repeats = 5L)
+            inner = rsmp("cv", folds = 2L),
+            outer = rsmp("repeated_cv", folds = 2L, repeats = 1L)
         ),
         deployment = "main"
     ),

### 2021-10-15 analysis/article.Rmd
fix: gtsummary compact theme
feat: add ml method part

@@
 knitr::write_bib(
-    c("targets", "workflowr"),
+    c(
+        "base",
+        "targets", "workflowr",
+        "mlr3", "mlr3proba", "zlog",
+        "survival", "glmnet", "penalized", "ranger", "randomForestSRC",
+        "xgboost", "survivalsvm", "survivalmodels"
+    ),
     file = rprojroot::find_rstudio_root_file(
@@
 library("targets")
+library("ggplot2")
+library("mlr3viz")
+library("viridis")
 library("ameld")
@@
 tar_load(bmrk_results)
+tar_load(crossval)
 
@@
 ```{r ml, include = FALSE}
-n_lrns <- nrow(bmrk_results$learners)
+n_lrns <- sum(!grepl("^scale..*", bmrk_results$learners$learner_id))
 #n_cv <- c(inner.folds
@@
 
+All statistical and machine learning analyses were performed using R version
+`r paste0(sessionInfo()$R.version[c("major", "minor")], collapse = ".")`
+[@R-base].
+Prior to the analyses all laboratory values were zlog transformed
+[@hoffmann2017; @R-zlog].
+We compared `r n_lrns` different statistical and machine learning
+algorithms using the mlr3 framework
+[@mlr3; @R-mlr3; @mlr3proba2021; @R-mlr3proba].
+The used algorithms are designed for analysis of survival data:
+Cox proportional hazards regression model
+[@cox1972; @survival-book; @R-survival],
+penalized regressions,
+namely two different implementations of the
+lasso (least absolute shrinkage and selection operator) regression
+[@tibshirani1997; @glmnet2011; @R-glmnet; @penalized2010; @R-penalized],
+tow different implementations of the
+ridge regression [@glmnet2011; @R-glmnet; @penalized2010; @R-penalized], and
+the elastic net regression [@glmnet2011; @R-glmnet],
+two different random forest implementations
+[@ranger2017; @R-ranger; @randomForestSRC2008; @R-randomForestSRC],
+extreme gradient boosting [@chen2016; @R-xgboost],
+two different support vector machines [@vanbelle2011; @R-survivalsvm]
+and a neural network [@katzman2018; @R-survivalmodels].
+
+To choose the best model we used nested resampling to avoid feature- and
+model-selection-bias.
+Therefore, we applied a `r crossval$inner$param_set$values$folds`-fold inner
+and a `r crossval$outer$param_set$values$folds`-fold outer cross-validation
+that was repeated `r crossval$outer$param_set$values$repeats` times.
+Subsequently we selected the model with the
+highest median concordance index [@harrell1982].
+
+TODO: selected final model, analysis, feature selection
+
 # Results
@@
 ```{r tbl1, results = "asis", echo = FALSE, message = FALSE}
+library("gtsummary")
+theme_gtsummary_journal(journal = "jama")
+theme_gtsummary_compact()
 tbl1
@@
 
+```{r benchmark-plot, fig.cap = "Benchmark results of machine learning algorithms."}
+r <- bmrk_results$clone()
+id <- grep("^scale", x = r$learners$learner_id, invert = TRUE, value = TRUE)
+r$filter(task_id = "zlog_eldd", learner_id = id)
+autoplot(r) +
+    geom_boxplot(aes(fill = learner_id)) +
+    geom_jitter(position = position_jitter(0.2)) +
+    scale_fill_viridis(discrete = TRUE) +
+    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+```
+
+TODO: mention best ml algorithm (\@ref(fig:benchmark-plot))
+
 # Discussion

### 2021-10-15 analysis/benchmarks.Rmd
refactor: increase figure width
refactor: reset figure width

@@
   bookdown::html_document2:
-    fig.width: 12
+    fig.width: 18
     fig.height: 8
@@
 
-```{r benchmarks_plot_c_harrell}
+```{r benchmarks_plot_c_harrell, fig.width = 18}
 autoplot(bmrk_results) +

### 2021-10-15 analysis/article.Rmd
fix: remove missing cross reference

@@
 respectively.
-As depicted in Figure \@ref(fig:observed-vs-meld-mortality-plot) and Table
-\@ref(tab:tbl-observed-vs-expected-mortality) our observed 90 day mortality was
+As shown in Table \@ref(tab:tbl-observed-vs-expected-mortality)
+our observed 90 day mortality was
 `r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`

### 2021-10-19 code/08-tables.R
fix: percent sign in mortality table

@@ -72,7 +72,7 @@ tg_tables <- list(
             add_overall()
 
         ## generate mortality table stub
-        ## (tbl_survfit doesn't support (cumulative) events)
+        ## (tbl_survfit doesn't support (cumulative) events) and t()
         tbMortality <- labelled_meld_data |>
             select(
                Sex,
@@ -111,15 +111,15 @@ tg_tables <- list(
             label = paste0("Within ", times, " days"),
             stat_label = "n (%)",
             stat_0 = paste0(
-                n_overall, " (", style_percent(m_overall), ")"
+                n_overall, " (", style_percent(m_overall, symbol = TRUE), ")"
             ),
             stat_1 = paste0(
                 n_sex[, "Female"],
-                " (", style_percent(m_sex[, "Female"]), ")"
+                " (", style_percent(m_sex[, "Female"], symbol = TRUE), ")"
             ),
             stat_2 = paste0(
                 n_sex[, "Male"],
-                " (", style_percent(m_sex[, "Male"]), ")"
+                " (", style_percent(m_sex[, "Male"], symbol = TRUE), ")"
             )
         )
 
@@ -165,7 +165,9 @@ tg_tables <- list(
         tbl1 |>
         modify_caption(
             paste0(
-                "Baseline characteristics.\n",
+                "Baseline characteristics.<br />",
+                "Values are given as ",
+                "median (lower quartile, upper quartile) or n (percent).<br />",
                 paste0(
                     Filter(
                         nchar,

### 2021-10-19 analysis/article.Rmd
feat: describe zlog in more detail

@@
 
-## Machine learning
+## Data processing and machine learning
 
@@
 
-All statistical and machine learning analyses were performed using R version
+All data processing, statistical and machine learning analyses were performed
+using R version
 `r paste0(sessionInfo()$R.version[c("major", "minor")], collapse = ".")`
 [@R-base].
-Prior to the analyses all laboratory values were zlog transformed
-[@hoffmann2017; @R-zlog].
+Prior to the analyses all laboratory values were *zlog* transformed
+as described in @hoffmann2017 and implemented in @R-zlog.
+
+$$
+z = (\log(x) - \frac{\log(LL) + \log(UL)}{2}) * \frac{3.92}{\log(UL) - \log(LL)}
+$$
+
+Where $LL$ and $UL$ are the lower and upper reference limit of
+the corresponding laboratory diagnostic, respectively.
+Missing laboratory measurements were replaced with a *zlog* value of zero.
+In similar manner if information about complications or
+comorbidities were missing we treat them as not present.
+
 We compared `r n_lrns` different statistical and machine learning
@@
 Subsequently we selected the model with the
-highest median concordance index [@harrell1982].
+highest median concordance index [@harrell1982]. If multiple models were
+comparable we chose the simpler (more penalized) one.
 

### 2021-10-19 analysis/article.Rmd
feat: cite gtsummary

@@
         "targets", "workflowr",
+        "gtsummary",
         "mlr3", "mlr3proba", "zlog",
@@
 We compared `r n_lrns` different statistical and machine learning
-algorithms using the mlr3 framework
+algorithms using the *mlr3* framework
 [@mlr3; @R-mlr3; @mlr3proba2021; @R-mlr3proba].
@@
 TODO: selected final model, analysis, feature selection
+
+Summary tables were built using the *gtsummary* package
+[@gtsummary; @R-gtsummary].
 

### 2022-02-10 scripts/Rscript.sh
fix: adapt to new singularity version on HPC

@@ -6,7 +6,11 @@ BIND_PATH="/etc/passwd,/etc/group,/var/run"
 MODULESPATH="/usr/share/[Mm]odules/init/bash"
 if [ -f ${MODULESPATH} ] ; then
     . ${MODULESPATH}
-    SINGULARITYVER=3.4.2
+    SINGULARITYVER=3.8.2
+    if [ ! -d "/opt/singularity-${SINGULARITYVER}" ] ; then
+        SINGULARITYVER=3.4.2
+    fi
+    SINGULARITYPATH=/opt/singularity-${SINGULARITYVER}/bin/singularity
     module load singularity/${SINGULARITYVER}
     SINGULARITYPATH=/opt/singularity-${SINGULARITYVER}/bin/singularity
     BIND_PATH="${BIND_PATH},/var/run/slurm/conf/slurm.conf:/etc/slurm/slurm.conf"
@@ -16,14 +20,12 @@ else # local
     SINGULARITYPATH=/usr/bin/singularity
 fi
 
-SINGULARITYENV_LOCPATH=/lib/locale/2.31/ \
-    SINGULARITYENV_RETICULATE_PYTHON=/bin/python3.8 \
-    SINGULARITYENV_PYCOX_DATA_DIR=${TMPDIR:-/tmp} \
-    SINGULARITY_TMPDIR=${TMPDIR:-/tmp} \
-    SINGULARITYENV_TMPDIR=${TMPDIR:-/tmp} \
-    SINGULARITYENV_TAR_MAKE_REPORTER="timestamp" \
-    SINGULARITYENV_TZ="UTC" \
+SINGULARITY_TMPDIR=${TMPDIR:-/tmp}
     SINGULARITY_BIND="${BIND_PATH}" \
     ${SINGULARITYPATH} run container/ampel-leipzig-meld.squashfs \
-    "--vanilla ${@}"
+    "--no-save --no-restore --no-site-file --no-init-file ${@}"
+    # don't use --vanilla to read .Renviron to set TMP and PYTHONDIR which
+    # is not possible via SINGULARITY_TMPDIR/SINGULARITYENV_*
+    # in the current singularity 3.8.2 on the HPC
+    #"--vanilla ${@}"
 exit 0

### 2022-02-10 code/03-mlr-learners.R
feat: add coxtime learner

@@ -155,6 +155,30 @@ tg_lrns <- list(
                     p
                 }
             ),
+            coxtime = list(
+                lrn = lrn(
+                    "surv.deepsurv", id = "coxtime",
+                    activation = "leakyrelu",
+                    #frac = 0.3, # we run our own nested cv
+                    early_stopping = TRUE,
+                    epochs = 100, optimizer = "adam"
+                ),
+                ps = {
+                    p <- ps(
+                        dropout = p_dbl(lower = 0, upper = 1),
+                        weight_decay = p_dbl(lower = 0, upper = 0.5),
+                        learning_rate = p_dbl(lower = 0, upper = 1),
+                        nodes = p_int(lower = 1, upper = 32),
+                        k = p_int(lower = 1, upper = 4)
+                    )
+                    p$trafo <- function(x, param_set) {
+                        x$num_nodes <- rep(x$nodes, x$k)
+                        x$nodes <- x$k <- NULL
+                        x
+                    }
+                    p
+                }
+            ),
             deepsurv = list(
                 lrn = lrn(
                     "surv.deepsurv", id = "deepsurv",

### 2022-02-10 code/03-mlr-learners.R
refactor: change deepsurv's search space

@@ -183,22 +183,20 @@ tg_lrns <- list(
                 lrn = lrn(
                     "surv.deepsurv", id = "deepsurv",
                     activation = "leakyrelu",
-                    frac = 0.3, early_stopping = TRUE,
+                    #frac = 0.3, # we run our own nested cv
+                    early_stopping = TRUE,
                     epochs = 100, optimizer = "adam"
                 ),
                 ps = {
                     p <- ps(
-                        dropout =
-                            p_dbl(lower = 0, upper = 0.7, default = 0.5),
+                        dropout = p_dbl(lower = 0, upper = 0.7),
                         ## 0.004, 0.002, 0.001, 0.0005, 0.0002, 0.0001, 0
-                        weight_decay =
-                            p_int(lower = 0, upper = 6, default = 2),
-                        learning_rate =
-                            p_dbl(lower = 0, upper = 0.2, default = 0.1),
+                        weight_decay = p_int(lower = 0, upper = 6),
+                        learning_rate = p_dbl(lower = 0, upper = 0.2),
                         ## nodes in a layer (2, 4, 8, 16, 32, 64, 128)
-                        nodes = p_int(lower = 0L, upper = 7L, default = 5L),
+                        nodes = p_int(lower = 0L, upper = 7L),
                         ## number of hidden layers
-                        k = p_int(lower = 0L, upper = 4L, default = 2L)
+                        k = p_int(lower = 1L, upper = 4L)
                     )
                     p$trafo <- function(x, param_set) {
                         x$weight_decay <-

### 2022-02-10 code/03-mlr-learners.R
refactor: code rearrangement

@@ -11,8 +11,20 @@ tg_lrns <- list(
     ## we want to rerun the benchmarking when anything changes
     tar_target(crossval,
         list(
-            inner = rsmp("cv", folds = 2L),
-            outer = rsmp("repeated_cv", folds = 2L, repeats = 1L)
+            inner = rsmp("cv", folds = 3L),
+            outer = rsmp("repeated_cv", folds = 3L, repeats = 10L)
+        ),
+        deployment = "main"
+    ),
+    tar_target(terminator,
+        trm("combo",
+            list(
+                trm("evals", n_evals = 50L),
+                trm("stagnation", iters = 3L, threshold = 1e-2),
+                trm("stagnation_batch", n = 1L, threshold = 1e-2),
+                trm("run_time", secs = 60)
+            ),
+            any = TRUE
         ),
         deployment = "main"
     ),
@@ -210,19 +222,7 @@ tg_lrns <- list(
             )
         )
         ## tuner
-        tuner <-
-            tnr("grid_search", resolution = 4L, batch_size = 4L)
-            #tnr("random_search", batch_size = 5L)
-
-        ## terminator
-        terminator <- trm("combo", list(
-                trm("evals", n_evals = 50L),
-                trm("run_time", secs = 90L),
-                trm("stagnation", iters = 3L, threshold = 1e-3),
-                trm("stagnation_batch", n = 1L, threshold = 1e-3)
-            ),
-            any = TRUE
-        )
+        tuner <- tnr("grid_search", resolution = 10L, batch_size = 5L)
 
         ## autotuner
         l <- lapply(l, function(ll) {
@@ -239,14 +239,15 @@ tg_lrns <- list(
                 )
         })
 
-        ### preprocessing (is part of the crossval now)
-        l <- c(
+        ## preprocessing (is part of the crossval now)
+        l <- c(l, lapply(
             l,
-            lapply(
-                l,
-                function(ll)as_learner(po("scale") %>>% po("learner", ll))
-            )
-        )
+            function(ll)as_learner(po("scale") %>>% po("learner", ll))
+        ))
+
+        ## name learners
+        names(l) <- vapply(l, "[[", NA_character_, "id")
+
         l
     },
     iteration = "list",

### 2022-02-10 code/04-mlr-benchmarking.R
feat: name measurements

@@ -36,8 +36,8 @@ tg_benchmark_results <- list(
         as.BenchmarkAggr(
             bmrk_results,
             measures = list(
-                msr("surv.cindex"),
-                msr("surv.cindex", weight_meth = "G2")
+                msr("surv.cindex", id = "harrell"),
+                msr("surv.cindex", id = "uno", weight_meth = "G2")
             )
         ),
         packages = c("mlr3", "mlr3proba", "mlr3benchmark"),

### 2022-02-10 analysis/article.Rmd
refactor: switch to grid search

@@
 tar_load(crossval)
+tar_load(tuner)
 
@@
 two different support vector machines [@vanbelle2011; @R-survivalsvm]
-and a neural network [@katzman2018; @R-survivalmodels].
+and two neural networks [@katzman2018; @kvamme2019; @R-survivalmodels].
 
@@
 that was repeated `r crossval$outer$param_set$values$repeats` times.
+We searched the hyperparameter space using a grid search with a resolution of
+`r tuner$param_set$get_values()$resolution`.
 Subsequently we selected the model with the

### 2022-02-10 analysis/article.Rmd
chore: remove comments

@@
 n_lrns <- sum(!grepl("^scale..*", bmrk_results$learners$learner_id))
-#n_cv <- c(inner.folds
-#    outer = unlist(
-#        bmrk_results$resamplings$resampling[[1]]$param_set$get_values()
-#    )
-#)
 ```

### 2022-02-10 analysis/benchmarks.Rmd
feat: add benchmark table and plots

@@
 library("targets")
+library("data.table")
+library("mlr3")
+library("mlr3misc")
+library("mlr3proba")
 library("mlr3viz")
@@
 
+agg <- bmrk_results$aggregate(msr("surv.cindex", id = "harrell"))
+agg[, `:=`
+    (nr = NULL, resample_result = NULL, resampling_id = NULL, iters = NULL)
+]
+setorder(agg, task_id, -harrell)
 p.value <- 0.2
@@
 
+## Table
+
+```{r benchmarks_table, results = "asis"}
+lapply(split(agg, agg$task_id), knitr::kable, digits = 4)
+```
+
+## Boxplots
+
 ```{r benchmarks_plot_c_harrell, fig.width = 18}
+m <- agg[, max(harrell), by = task_id]
 autoplot(bmrk_results) +
+    geom_boxplot(aes(fill = learner_id)) +
+    geom_jitter(position = position_jitter(0.2)) +
+    scale_fill_viridis(discrete = TRUE) +
+    geom_hline(
+        aes(yintercept = V1), linetype = "dashed", color = "red", data = m
+    ) +
+    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+```
+
+```{r benchmarks_plot_c_uno, fig.width = 18}
+autoplot(
+    bmrk_results,
+    measure = msr("surv.cindex", id = "uno", weight_meth = "G2")) +
     geom_boxplot(aes(fill = learner_id)) +
@@
 
-Aggregated performance across all `nrow(bmrk_results$tasks)` datasets.
+Aggregated performance across all `r nrow(bmrk_results$tasks)` datasets.
 
 ```{r benchmarks_aggr_plot_c_harrell}
-autoplot(bmrk_aggr, type = "box", meas = "harrell_c") +
+autoplot(bmrk_aggr, type = "box", meas = "harrell") +
+    geom_boxplot(aes(fill = learner_id)) +
+    geom_jitter(position = position_jitter(0.2)) +
+    scale_fill_viridis(discrete = TRUE) +
+    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+```
+
+```{r benchmarks_aggr_plot_c_uno}
+autoplot(bmrk_aggr, type = "box", meas = "uno") +
     geom_boxplot(aes(fill = learner_id)) +
@@
 autoplot(
-    bmrk_aggr, type = "fn", meas = "harrell_c", p.value = p.value,
+    bmrk_aggr, type = "fn", meas = "harrell", p.value = p.value,
     col = palette.colors(2L)[2L]
@@
 autoplot(
-    bmrk_aggr, type = "cd", meas = "harrell_c", style = 2, p.value = p.value,
+    bmrk_aggr, type = "cd", meas = "harrell", style = 2, p.value = p.value,
     minimize = FALSE

### 2022-02-10 code/03-mlr-learners.R
refactor: turn tuner into a target

@@ -28,6 +28,11 @@ tg_lrns <- list(
         ),
         deployment = "main"
     ),
+    tar_target(tuner,
+        #tuner <- tnr("random_search", batch_size = 10L)
+        tuner <- tnr("grid_search", resolution = 20L, batch_size = 10L),
+        deployment = "main"
+    ),
     tar_target(learners, {
         l <- list(
             cox = list(
@@ -221,8 +226,6 @@ tg_lrns <- list(
                 }
             )
         )
-        ## tuner
-        tuner <- tnr("grid_search", resolution = 10L, batch_size = 5L)
 
         ## autotuner
         l <- lapply(l, function(ll) {

### 2022-03-10 code/make.R
refactor: use timestamp reporter

@@ -1,4 +1,6 @@
-targets::tar_make_future()
+targets::tar_make_future(
+        reporter = "timestamp"
+)
 workflowr::wflow_build(
         rprojroot::find_rstudio_root_file("analysis", "pipeline.Rmd"),
         view = FALSE

### 2022-03-10 analysis/article.Rmd
feat: describe model development in methods section

@@
 tar_load(tbl_observed_vs_expected_mortality)
+
+tar_load(ameldcfg)
 ```
@@
 
-TODO: selected final model, analysis, feature selection
+After selecting the final model we train and validated it as described in
+@harrell1996.
+Briefly we develop our model and predict survival at `r ameldcfg$times` days.
+The predictions were grouped into equal sized intervals so that there
+are `r ameldcfg$m` patients in each interval.
+For each group the mean survival probability and the difference to the
+Kaplan-Meier estimate was calculated.
+The same procedure was repeated for `r ameldcfg$nboot` bootstrap samples.
+Subsequently the differences between predicted survival and Kaplan-Meier
+estimates of the bootstrap samples were averaged and added to the original
+model difference to obtain a bias-corrected estimated between predicted and
+observed survival.
+
+The final model was compared against the MELD, MELD-Na and MELD-Plus7 score
+using the area under the time-dependent ROC curve (AUC)
+based on the nonparametric inverse probability of censoring weighting estimator
+(IPCW) [@blanche2013].
 
@@
 ```
-
-<!-- TODO:
-use `pattern` argument to provide "{n} ({p}%)" for categorial/dichotomos
-variables after this PR is merged
-https://github.com/ddsjoberg/gtsummary/pull/951
--->
 

### 2022-03-11 code/01-data.R
fix: adapt to ameld 0.0.17

@@ -29,7 +29,8 @@
         dialysis = x$Dialysis,
         type = "unos"
     )
-    x$ProbMeldPlus7 <- meld_plus7(
+    x$ProbMeldUnos <- pmeld(x$ScoreMeldUnos)
+    x$ProbMeldPlus7 <- pmeld_plus7(
         creatinine = as_metric(x$CRE_S, "creatinine"),
         bilirubin = as_metric(x$BILI_S, "bilirubin"),
         inr = x$INR_C,

### 2022-03-11 analysis/article.Rmd
feat: add co2eq estimation

@@
 
+```{r co2e}
+tdp <- c(
+    ## brain/batch Intel(R) Xeon(R) CPU E5-2650 v4 @ 2.20GHz
+    ## https://ark.intel.com/content/www/us/en/ark/products/91767/intel-xeon-processor-e52650-v4-30m-cache-2-20-ghz.html
+    batch = 105,
+    ## brain/snowball Intel(R) Xeon(R) Gold 6240 CPU @ 2.60GHz
+    ## https://www.intel.com/content/www/us/en/products/sku/192443/intel-xeon-gold-6240-processor-24-75m-cache-2-60-ghz.html
+    snowball = 150
+)
+
+## Umweltbundesamt
+## Climate Change 45/2021 Entwicklung der spezifischen Kohlendioxid-Emissionen des deutschen Strommix in den Jahren 1990 - 2020
+## Erscheinungsjahr Mai 2021, Autor(en): Petra Icha, Dr. Thomas Lauf, Gunter Kuhs
+## p. 8
+## https://www.umweltbundesamt.de/sites/default/files/medien/5750/publikationen/2021-05-26_cc-45-2021_strommix_2021_0.pdf
+kgco2kwh <- 0.471
+
+elapsed_runtime <- sum(
+    subset(tar_meta(), !type %in% c("function", "object", "branch"))$seconds
+)
+
+kgco2e <- tdp["snowball"] / 1000 * elapsed_runtime / 3600 * kgco2kwh
+kgco2e
+
+## Umweltbundesamt
+## Ökologische Bewertung von Verkehrsarten
+## Erscheinungsjahr August 2020, Autor(en): Michel Allekotte, Fabian Bergk, Kirsten Biemann, Carolin Deregowski, Wolfram Knörr, Hans-Jörg-Althaus, Daniel Sutter, Thomas Bergmann
+## p. 214-215
+## https://www.umweltbundesamt.de/sites/default/files/medien/479/publikationen/texte_156-2020_oekologische_bewertung_von_verkehrsarten_0.pdf
+## Gesamt CO2eq (nur Verbrauch und Energiebereitstellung) je Personen km,
+## A2, Tab. 69: PKW TTW 131 gCO2eq/Pkm, A3, Tab. 71.: PKW WTT 22 gCO2eq/Pkm
+carkgco2e <- 153/ 1000
+km <- kgco2e / carkgco2e
+```
+
+The calculations were conducted using the high performance computing (HPC)
+cluster of the University Greifswald.
+In 2018 the average carbon efficiency of the German power grid was
+`r kgco2kwh` kgCO~2~eq/kWh [@icha2021].
+A single run of the complete calculation pipeline took
+`r sprintf("%.1f", elapsed_runtime / 3600)` hours of computation on an
+Intel(R) Xeon(R) Gold 6240 CPU (thermal design power: 150 W).
+The total estimated emissions were at least
+`r sprintf("%.2f", kgco2e)` kgCO~2~eq
+(ignoring development, test runs, etc.).
+This is equivalent to `r sprintf("%.1f", km)` km driven by an average car
+[@allekotte2021].
+
+# Acknowledgement
+
+The study was supported by the AMPEL project (www.ampel.care).
+This project is co-financed through public funds according to the budget
+decided by the Saxon State Parliament under the
+RL eHealthSax 2017/18 grant number 100331796.
+The funder provided support in the form of
+salaries for author SG
+but did not have any additional role in the study design,
+data collection and analysis, decision to publish, or preparation
+of the manuscript.
+We thank the whole AMPEL team for their support,
+and Stefan Kemnitz for his technical support regarding the use of the
+HPC cluster of the University Greifswald.
+
 # References

### 2022-03-11 analysis/article.Rmd
fix: cite timeROC package

@@
         "survival", "glmnet", "penalized", "ranger", "randomForestSRC",
-        "xgboost", "survivalsvm", "survivalmodels"
+        "xgboost", "survivalsvm", "survivalmodels", "timeROC"
     ),
@@
 based on the nonparametric inverse probability of censoring weighting estimator
-(IPCW) [@blanche2013].
+(IPCW) [@blanche2013, @R-timeROC].
 

### 2022-03-15 analysis/article.Rmd
fix: co2eq calculation

@@
 
-```{r benchmark-plot, fig.cap = "Benchmark results of machine learning algorithms."}
+```{r benchmark-plot, fig.cap = "Benchmark results of machine learning algorithms.", echo = FALSE}
 r <- bmrk_results$clone()
@@
 
-```{r co2e}
+```{r co2e, include = FALSE}
 tdp <- c(
-    ## brain/batch Intel(R) Xeon(R) CPU E5-2650 v4 @ 2.20GHz
+    ## brain/(batch|login) Intel(R) Xeon(R) CPU E5-2650 v4 @ 2.20GHz
     ## https://ark.intel.com/content/www/us/en/ark/products/91767/intel-xeon-processor-e52650-v4-30m-cache-2-20-ghz.html
-    batch = 105,
+    login = 105,
     ## brain/snowball Intel(R) Xeon(R) Gold 6240 CPU @ 2.60GHz
     ## https://www.intel.com/content/www/us/en/products/sku/192443/intel-xeon-gold-6240-processor-24-75m-cache-2-60-ghz.html
+    ## TDP 150, contains 2 processors
+    ## (but even with workers = 72, we barely use more than 36, so 150 instead of 300 W should be fine)
     snowball = 150
@@
 
-elapsed_runtime <- sum(
-    subset(tar_meta(), !type %in% c("function", "object", "branch"))$seconds
-)
+tm <- tar_meta(fields = c("name", "type", "seconds"))
+tm <- tm[tm$type %in% c("pattern", "stem"),]
+tm$node <-
+    ifelse(grepl("^boot|^bmrks$|^arcvob$", tm$name), "snowball", "login")
 
-kgco2e <- tdp["snowball"] / 1000 * elapsed_runtime / 3600 * kgco2kwh
+tm$runtime <- tm$seconds *
+    ifelse(tm$node == "snowball", 32, 1) # on snowball we use 32 workers
+kgco2e <- sum((tm$runtime / 3600) * (tdp[tm$node] / 1000)) * kgco2kwh
 kgco2e
@@
 ## A2, Tab. 69: PKW TTW 131 gCO2eq/Pkm, A3, Tab. 71.: PKW WTT 22 gCO2eq/Pkm
-carkgco2e <- 153/ 1000
+carkgco2e <- 153 / 1000
 km <- kgco2e / carkgco2e
@@
 `r kgco2kwh` kgCO~2~eq/kWh [@icha2021].
-A single run of the complete calculation pipeline took
-`r sprintf("%.1f", elapsed_runtime / 3600)` hours of computation on an
-Intel(R) Xeon(R) Gold 6240 CPU (thermal design power: 150 W).
+A single run of the complete calculation pipeline took roughly
+`r sprintf("%.1f", sum(tm$runtime) / 3600)` hours of computation on an
+Intel(R) Xeon(R) E5-2650 v4 and
+Intel(R) Xeon(R) Gold 6240 CPU (thermal design power: 105 and 150 W).
 The total estimated emissions were at least

### 2022-03-17 _targets.R
chore: reset future params, callr fails on HPC and remote is deprecated

@@ -3,7 +3,6 @@ library("tarchetypes")
 
 library("future")
 library("future.batchtools")
-library("future.callr")
 
 library("lgr")
 
@@ -18,28 +17,40 @@ get_logger("mlr3")$add_appender(AppenderFile$new(tf), name = "file")
 get_logger("bbotk")$add_appender(AppenderFile$new(tf), name = "file")
 
 login <- future::tweak(
-    future::remote,
+    future::cluster,
     workers = "brain",
+    homogeneous = FALSE,
     rscript = "scripts/Rscript.sh",
     rshopts = if (nchar(Sys.getenv("DEBUGME"))) "-v" else ""
+)
+
+resources <- list(
+    workers = 32L,
+    partition = "snowball",
+    ncpu = 36,
+    mcpu = 1L * 1024L,
+    walltime = 4L * 60L * 60L # seconds
 )
 
 slurm <- future::tweak(
     future.batchtools::batchtools_slurm,
     template = "scripts/slurm_batchtools.tmpl",
-    workers = 16L,
-    resources = list(
-        partition = "batch,compute,snowball",
-        ncpu = 64,
-        mcpu = 2L * 1024L,
-        walltime = 48L * 60L * 60L # seconds
-    )
+    workers = resources$workers,
+    resources = resources
+)
+
+node <- future::tweak(
+    multisession,
+    workers = resources$ncpu
 )
 
 if (nchar(Sys.getenv("RUNLOCAL"))) {
-    future::plan(callr, workers = 7L)
+    future::plan(list(
+        future::tweak(multisession, workers = 2L),
+        future::tweak(multicore, workers = 3L))
+    )
 } else {
-    future::plan(list(login, slurm, callr))
+    future::plan(list(login, slurm, node))
 }
 
 source("code/01-data.R")

### 2022-03-17 analysis/benchmarks.Rmd
feat: add elastic net bootstrap and timeROC evaluations

@@
 setorder(agg, task_id, -harrell)
-p.value <- 0.2
+p.value <- 0.05
 ```

### 2022-03-17 analysis/elasticnet.Rmd
feat: add elastic net bootstrap and timeROC evaluations

---
title: "elasticnet"
output:
  bookdown::html_document2:
    fig.width: 18
    fig.height: 8
---

```{r elasticnet_setup, message = FALSE}
library("targets")
library("ameld")
library("viridisLite")
tar_load(arcvob)
tar_load(bootrcv)
tar_load(bootarcv)
#tar_load(bootarcv9)
tar_load(amelddata)
tar_load(ameldcfg)
tar_load(zlog_data)
```

# Tuning alpha

```{r arcvob}
arcvob
```

```{r arcvob-plot-alpha}
plot(arcvob)
```

```{r arcvob-plot-lambdamin}
plot(arcvob, what = "lambda.min")
```

```{r arcvob-plot-lambda1se}
plot(arcvob, what = "lambda.1se")
```

# Bootstrapping

## rcv.glmnet

```{r bootrcv}
plot(bootrcv, what = "calibration")

ps <- lapply(
    zlog_data[paste0("SurvProbMeld", c("Unos", "NaUnos", "Plus7"))],
    function(p) {
        ctpnts <- cutpoints(p, n = ameldcfg$m)
        f <- cut(p, ctpnts, include.lowest = TRUE)
        list(
            predicted = groupmean(p, f = f),
            observed = observed_survival(
                amelddata$y, f = f, times = ameldcfg$times
            )
        )
    }
)
names(ps) <- c("MELD", "MELD-Na", "MELD-Plus7")
col <- viridisLite::viridis(6)[4:6]

for (i in seq_along(ps)) {
    lines(
        ps[[i]]$predicted, ps[[i]]$observed, col = col[i], type = "b", pch = 19
    )
}
legend("topleft", col = col, legend = names(ps), pch = 19, bty = "n")

plot(bootrcv, what = "selected", cex = 0.5)
plot(bootrcv$fit$glmnet.fit, xvar = "norm")
plot(bootrcv$fit$glmnet.fit, xvar = "lambda")
plot(bootrcv$fit$glmnet.fit, xvar = "dev")
```

## arcv.glmnet

```{r bootarcv}
a <- c(table(sapply(bootarcv$models, function(m)m$fit$alpha)))
plot(bootarcv, what = "calibration")
```
```{r bootarcv2, fig.height = 12}
plot(bootarcv, what = "selected")
plot_dots(a, main = "Selected Alpha Values")
```

## arcv9.glmnet

```{r bootarcv9}
#a <- c(table(sapply(bootarcv9$models, function(m)m$fit$alpha)))
#plot(bootarcv9 , what = "calibration")
#plot(bootarcv9, what = "selected")
#plot_dots(a, main = "Selected Alpha Values")
```

### 2022-03-17 _targets.R

@@ -57,17 +57,22 @@ source("code/01-data.R")
 source("code/02-mlr-tasks.R")
 source("code/03-mlr-learners.R")
 source("code/04-mlr-benchmarking.R")
+source("code/05-ameld-elasticnet.R")
+source("code/06-timeroc.R")
 source("code/08-tables.R")
 source("code/09-reports.R")
 source("code/99-site.R")
 
 list(
    tg_data,
-   tg_lrns,
    tg_tasks,
+   tg_tables,
+   tg_lrns,
    tg_benchmarks,
    tg_benchmark_results,
-   tg_tables,
+   tg_ameld,
+   tg_timeroc,
+   tg_timeroc_psurv,
    tg_reports,
    tg_site
 )

### 2022-03-17 code/05-ameld-elasticnet.R

@@ -1,0 +1,126 @@
+tg_ameld <- list(
+    tar_target(ameldcfg, list(
+        alpha = seq(0, 1, len = 11)^3,
+        nboot = 200L,
+        nrepcv = 100L,
+        nfolds = 3L,
+        m = 50,
+        times = 90,
+        standardize = FALSE
+    ),
+    deployment = "main"
+    ),
+    tar_target(amelddata, {
+        d <- zlog_data
+        ## exclude precalculated scores
+        d <- d[!grepl("Meld", colnames(d))]
+        srv <- Surv(d$DaysAtRisk, d$Deceased)
+        d$DaysAtRisk <- d$Deceased <- NULL
+        d <- data.matrix(d)
+        list(x = d, y = srv)
+    },
+    packages = c("survival"),
+    deployment = "main"
+    ),
+    tar_target(arcvob, {
+        arcv.glmnet(
+            x = amelddata$x, y = amelddata$y,
+            family = "cox",
+            alpha = ameldcfg$alpha,
+            nrepcv = ameldcfg$nrepcv,
+            nfolds = ameldcfg$nfolds,
+            balanced = TRUE,
+            standardize = ameldcfg$standardize,
+            trace.it = FALSE
+        )
+    },
+    packages = c("ameld", "survival"),
+    deployment = "worker"
+    ),
+    tar_target(bootrcv, {
+        bootrcv <- bootstrap(
+            x = amelddata$x, y = amelddata$y,
+            fun = rcv.glmnet,
+            family = "cox",
+            nboot = ameldcfg$nboot,
+            nfolds = ameldcfg$nfolds,
+            nrepcv = ameldcfg$nrepcv,
+            standardize = ameldcfg$standardize,
+            m = ameldcfg$m,
+            times = ameldcfg$times
+        )
+        bootrcv
+    },
+    packages = c(
+        "ameld", "future", "future.batchtools", "glmnet", "survival"
+    ),
+    deployment = "worker"
+    ),
+    tar_target(psurvbootrcv, {
+        predict(
+            bootrcv$fit, x = amelddata$x, y = amelddata$y, newx = amelddata$x,
+            type = "survival", times = ameldcfg$times, s = "lambda.1se"
+        )[1,]
+    }),
+    tar_target(bootarcv, {
+        selarcv <- function(...) {
+            a <- arcv.glmnet(...)
[…]

### 2022-03-17 code/06-timeroc.R

@@ -1,0 +1,60 @@
+cols <- c("SurvProbMeldUnos", "SurvProbMeldNaUnos", "SurvProbMeldPlus7")
+nms <- c("MELD", "MELDNa", "MELDPlus7")
+
+smbs <- rlang::syms(paste0("timeROC_", nms))
+
+tg_timeroc <- tar_eval(
+    values = list(s = smbs, col = cols),
+    tar_target(
+        s,
+        command = {
+            tr <- timeROC::timeROC(
+                T = zlog_data$DaysAtRisk,
+                delta = zlog_data$Deceased,
+                marker = 1 - zlog_data[[col]],
+                cause = 1,
+                times = c(1, 3, 7, 30, 60, 90, 180, 365),
+                iid = TRUE
+            )
+            tr$confint <- confint(object = tr, level = 0.95, n.sim = 3000)
+            tr
+        },
+        packages = c("survival", "timeROC"),
+        deployment = "main"
+    )
+)
+
+tg_timeroc_psurv <- list(
+    tar_target(timeROC_RCV, {
+        tr <- timeROC::timeROC(
+            T = zlog_data$DaysAtRisk,
+            delta = zlog_data$Deceased,
+            marker = 1 - psurvbootrcv,
+            cause = 1,
+            times = c(1, 3, 7, 30, 60, 90, 180, 365),
+            iid = TRUE
+        )
+        tr$confint <- confint(object = tr, level = 0.95, n.sim = 3000)
+        tr
+    },
+    packages = c("survival", "timeROC"),
+    deployment = "main"
+    ),
+    tar_target(timeROC_ARCV, {
+        tr <- timeROC::timeROC(
+            T = zlog_data$DaysAtRisk,
+            delta = zlog_data$Deceased,
+            marker = 1 - psurvbootarcv,
+            cause = 1,
+            times = c(1, 3, 7, 30, 60, 90, 180, 365),
+            iid = TRUE
+        )
+        tr$confint <- confint(object = tr, level = 0.95, n.sim = 3000)
+        tr
+    },
+    packages = c("survival", "timeROC"),
+    deployment = "main"
+    )
+)
+
+rm("cols", "nms", "smbs")

### 2022-03-17 analysis/corrplot.Rmd
Revert "fix: working directory for targets"
This reverts commit cac14f630848b078c39aee7344862d418fcdcb52.
feat: list correlations in table

@@
 library("targets")
-
-withr::with_dir(rprojroot::find_rstudio_root_file(), {
 tar_load(imp_data)
-})
 ```
@@
 # Correlation
+
+## Plot
 
@@
 ```
+
+## Table
+
+```{r cortbl}
+ut <- upper.tri(cr)
+nms <- paste(
+    colnames(cr)[col(cr)[ut]],
+    rownames(cr)[row(cr)[ut]],
+    sep = ":"
+)
+tbl <- data.frame(Values = nms, Correlation = cr[ut])
+tbl <- tbl[order(-abs(tbl$Correlation)),]
+knitr::kable(tbl, header = TRUE, row.names = FALSE, digits = 3L)
+```

### 2022-03-17 analysis/histograms.Rmd
Revert "fix: working directory for targets"
This reverts commit cac14f630848b078c39aee7344862d418fcdcb52.
fix: MELD probailities

--- 2021-06-11 analysis/histograms.Rmd
@@
 ```
+
+# MELD Scores (imputed)
+
+```{r histograms_meld, results = "asis", message = FALSE, warning = FALSE}
+meldv <-
+    c(paste0("ScoreMeld", c("", "Unos", "Na", "NaUnos")),
+      paste0("SurvProbMeld", c("Unos", "NaUnos", "Plus7")))
+for (v in meldv) {
+    cat("\n\n##", v , "\n\n")
+    .hist(zlog_data[[v]], main = paste0("Histogram of ", v), xlab = v)
+    .qqnorm(zlog_data[[v]], yname = v)
+}
+```

### 2022-03-17 code/01-data.R

@@ -29,8 +29,9 @@
         dialysis = x$Dialysis,
         type = "unos"
     )
-    x$ProbMeldUnos <- pmeld(x$ScoreMeldUnos)
-    x$ProbMeldPlus7 <- pmeld_plus7(
+    x$SurvProbMeldUnos <- pmeld(x$ScoreMeldUnos)
+    x$SurvProbMeldNaUnos <- pmeld(x$ScoreMeldNaUnos)
+    x$SurvProbMeldPlus7 <- pmeld_plus7(
         creatinine = as_metric(x$CRE_S, "creatinine"),
         bilirubin = as_metric(x$BILI_S, "bilirubin"),
         inr = x$INR_C,
@@ -154,8 +155,6 @@ tg_data <- list(
                 attr(x$ScoreMeldUnos, "label") <- "MELD score"
             attr(x$ScoreMeldNa, "label") <-
                 attr(x$ScoreMeldNaUnos, "label") <- "MELD-Na score"
-            attr(x$ProbMeldPlus7, "label") <- "MELD-Plus7 risk score"
-
             attr(x$MeldCategory, "label") <- "MELD Category"
 
             ## labels where the column names are identical to the abbreviation

### 2022-03-17 code/08-tables.R

@@ -30,7 +30,7 @@ tg_tables <- list(
                Sex,
                ScoreMeldUnos,
                MeldCategory,
-               ScoreMeldNaUnos, ProbMeldPlus7,
+               ScoreMeldNaUnos,
             ) |>
             tbl_summary(
                 by = Sex,

### 2022-03-17 code/03-mlr-learners.R
fix: grid search settings

@@ -11,7 +11,16 @@ tg_lrns <- list(
     ## we want to rerun the benchmarking when anything changes
     tar_target(crossval,
         list(
-            inner = rsmp("cv", folds = 3L),
+            # works
+             # 3, 3,2, GS 4
+             # 5, 3,2, GS 4
+             # 5, 3,5, GS 8
+             # 5, 3,10, GS 10
+            # doesn't work/finish
+             # 5, 3,10 GS 12 (penridge > 4 h)
+             # 3, 3,10 GS 12 (coxtime 'obj' must have column names)
+             # 3, 3,10 GS 10 (coxtime 'obj' must have column names)
+            inner = rsmp("cv", folds = 5L),
             outer = rsmp("repeated_cv", folds = 3L, repeats = 10L)
         ),
         deployment = "main"
@@ -19,7 +28,7 @@ tg_lrns <- list(
     tar_target(terminator,
         trm("combo",
             list(
-                trm("evals", n_evals = 50L),
+                trm("evals", n_evals = 60L),
                 trm("stagnation", iters = 3L, threshold = 1e-2),
                 trm("stagnation_batch", n = 1L, threshold = 1e-2),
                 trm("run_time", secs = 60)
@@ -29,8 +38,7 @@ tg_lrns <- list(
         deployment = "main"
     ),
     tar_target(tuner,
-        #tuner <- tnr("random_search", batch_size = 10L)
-        tuner <- tnr("grid_search", resolution = 20L, batch_size = 10L),
+        tuner <- tnr("grid_search", resolution = 10L, batch_size = 5L),
         deployment = "main"
     ),
     tar_target(learners, {

### 2022-03-17 code/04-mlr-benchmarking.R

@@ -29,7 +29,7 @@ tg_benchmark_results <- list(
         bmrk_results,
         do.call(c, bmrks),
         packages = "mlr3",
-        deployment = "worker"
+        deployment = "main"
     ),
     tar_target(
         bmrk_aggr,
@@ -41,6 +41,6 @@ tg_benchmark_results <- list(
             )
         ),
         packages = c("mlr3", "mlr3proba", "mlr3benchmark"),
-        deployment = "worker"
+        deployment = "main"
     )
 )

### 2022-05-26 analysis/tables.Rmd
fix: column index in zlog colored table

@@
     k <- kableExtra::column_spec(
-            k, i,
+            k, i + 1,
             background = zcol(zlog_data[[i]]),

### 2022-06-06 analysis/article.Rmd
feat: add docx export

@@
   bookdown::html_document2:
+    pandoc_args: # pandoc_args doesn't support r evaluation
+      - --lua-filter=pandoc/lua-filters/scholarly-metadata.lua
+      - --lua-filter=pandoc/lua-filters/author-info-blocks.lua
+  bookdown::word_document2:
     pandoc_args: # pandoc_args doesn't support r evaluation
@@
 
-```{r flowchart, results = "asis", echo = FALSE, fig.width = 5, fig.height = 5, fig.cap = "Flowchart Inclusion/Exclusion."}
+```{r flowchart, results = "asis", eval = knitr::is_html_output(), echo = FALSE, fig.width = 5, fig.height = 5, fig.cap = "Flowchart Inclusion/Exclusion."}
 graph <- "

### 2022-06-06 code/01-data.R
refactor: disable age scaling

@@ -203,6 +203,10 @@ tg_data <- list(
                   grep("_[SECQ]$", colnames(imp_data), value = TRUE)
                 )
             zlog_data[iconvert] <- zlog_df(imp_data[iconvert], ref_data)
+            ## additionally scale Age
+            #agelimits <- c(32, 74)
+            #zlog_data$Age <- zlog(zlog_data$Age, agelimits)
+            #attr(zlog_data, "age_limits") <- agelimits
             zlog_data
         },
         packages = "zlog",

### 2022-06-06 analysis/benchmarks.Rmd
fix: bootstraping elastic net

@@
 
-```{r benchmarks_plot_c_harrell, fig.width = 18}
+```{r benchmarks_plot_c_harrell, fig.width = 18, fig.height = 12}
 m <- agg[, max(harrell), by = task_id]
@@
         aes(yintercept = V1), linetype = "dashed", color = "red", data = m
-    ) +
-    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+    )
 ```
 
-```{r benchmarks_plot_c_uno, fig.width = 18}
+```{r benchmarks_plot_c_uno, fig.width = 18, fig.height = 12}
 autoplot(
@@
     geom_jitter(position = position_jitter(0.2)) +
-    scale_fill_viridis(discrete = TRUE) +
-    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+    scale_fill_viridis(discrete = TRUE)
 ```

### 2022-06-06 analysis/elasticnet.Rmd
fix: bootstraping elastic net

@@
 tar_load(bootarcv)
-#tar_load(bootarcv9)
+tar_load(bootarcv7)
+tar_load(bootarcv9)
+tar_load(bootrcv.woICA)
 tar_load(amelddata)
@@
 
+## bootrcv.woICA
+
+```{r bootrcv.woICA}
+plot(bootrcv.woICA, what = "calibration")
+plot(bootrcv.woICA, what = "selected", cex = 0.5)
+```
+
 ## arcv.glmnet
@@
 
+## arcv7.glmnet
+
+```{r bootarcv7}
+a <- c(table(sapply(bootarcv9$models, function(m)m$fit$alpha)))
+plot(bootarcv7 , what = "calibration")
+plot(bootarcv7, what = "selected")
+plot_dots(a, main = "Selected Alpha Values")
+```
+
 ## arcv9.glmnet
@@
 ```{r bootarcv9}
-#a <- c(table(sapply(bootarcv9$models, function(m)m$fit$alpha)))
-#plot(bootarcv9 , what = "calibration")
-#plot(bootarcv9, what = "selected")
-#plot_dots(a, main = "Selected Alpha Values")
+a <- c(table(sapply(bootarcv9$models, function(m)m$fit$alpha)))
+plot(bootarcv9 , what = "calibration")
+plot(bootarcv9, what = "selected")
+plot_dots(a, main = "Selected Alpha Values")
 ```

### 2022-06-06 code/05-ameld-elasticnet.R

@@ -42,6 +42,28 @@ tg_ameld <- list(
             x = amelddata$x, y = amelddata$y,
             fun = rcv.glmnet,
             family = "cox",
+            nboot = 1000,
+            nfolds = 3,
+            nrepcv = 10,
+            standardize = ameldcfg$standardize,
+            m = ameldcfg$m,
+            times = ameldcfg$times
+        )
+        bootrcv
+    },
+    packages = c(
+        "ameld", "future", "future.batchtools", "glmnet", "survival"
+    ),
+    deployment = "worker"
+    ),
+    tar_target(bootrcv.woICA, {
+        x <- amelddata$x[,
+            !colnames(amelddata$x) %in% c("IL6_S", "CYSC_S", "Age")
+        ]
+        bootrcv <- bootstrap(
+            x = x, y = amelddata$y,
+            fun = rcv.glmnet,
+            family = "cox",
             nboot = ameldcfg$nboot,
             nfolds = ameldcfg$nfolds,
             nrepcv = ameldcfg$nrepcv,
@@ -92,35 +114,65 @@ tg_ameld <- list(
             bootarcv$fit, x = amelddata$x, y = amelddata$y, newx = amelddata$x,
             type = "survival", times = ameldcfg$times, s = "lambda.1se"
         )[1,]
-    })#,
-    #tar_target(bootarcv9, {
-    #    selarcv <- function(...) {
-    #        dots <- list(...)
-    #        a <- arcv.glmnet(...)
-    #        a$models[[
-    #            which.min.error(a, s = dots$s, maxnnzero = dots$maxnnzero)
-    #        ]]
-    #    }
+    }),
+    tar_target(bootarcv7, {
+        selarcv <- function(...) {
+            dots <- list(...)
+            a <- arcv.glmnet(...)
+            a$models[[
+                which.min.error(a, s = "lambda.1se", maxnnzero = dots$maxnnzero)
+            ]]
+        }
 
-    #    bootarcv9 <- bootstrap(
-    #        x = amelddata$x, y = amelddata$y,
-    #        fun = selarcv,
-    #        family = "cox",
-    #        alpha = ameldcfg$alpha,
-    #        s = "lambda.1se",
-    #        maxnnzero = 9,
-    #        nboot = ameldcfg$nboot,
-    #        nfolds = ameldcfg$nfolds,
-    #        nrepcv = ameldcfg$nrepcv,
-    #        standardize = ameldcfg$standardize,
-    #        m = ameldcfg$m,
[…]

### 2022-06-06 analysis/article.Rmd
feat: add first discussion draft

@@
 the MELD score [@kartoun2017]. The derived MELD-Plus7 and MELD-Plus9 risk score
-add albumin, white blood cell count, age and total cholesterol and length of
+add albumin, white blood cell count, total cholesterol, age and length of
 stay to the MELD-Na variables. Despite its published prediction improvement the
@@
 
-After selecting the final model we train and validated it as described in
+After selecting the final model we trained and validated it as described in
 @harrell1996.
-Briefly we develop our model and predict survival at `r ameldcfg$times` days.
+Briefly we developed our model and predicted survival at `r ameldcfg$times` days.
 The predictions were grouped into equal sized intervals so that there
@@
 who were recruited during the evaluation process for liver transplantation at
-the University Hospital of Leipzig we compared
-`r n_lrns` different statistical and machine learning algorithms.
+the University Hospital of Leipzig we developed a new model to predict the
+90-day survival probability. Our new model outperforms the MELD-Na score and
+the recently published MELD-Plus risk score.
 
-...
+**NUMBER_OF_VARIABLES**
+
+- more variables, harder to cheat
+- comparison of different predictors of MELD-Plus7 (+MELD-Plus9) vs our features
+
+Sodium is not part of our model.
+In line with the results presented in @kartoun2017 we could show that
+adding sodium to MELD the discrimination performance increases just
+insignificantly.
+
+It has shown that women get lower scores in MELD-Na and therefore
+are less likely getting an urgent listing for liver transplantation
+[@Gaertner2022].
+To address sexual differences in laboratory measurements we applied the *zlog*
+transformation [@hoffmann2017].
+Additionally this transformation yield comparable laboratory measurements
+across different clinical laboratories and methods which could improve the
+objectivity of the allocation process.
+The already nearly normal scaled *zlog*-transformed laboratory measurements
+allow us to avoid the sample depending scaling prior to model development.
+Both should increase the external applicability of our approach.
+
+Additionally to laboratory data we investigated
+`r (ncol(raw_data) - length(cn_labs) - 2)`
+clinical data but none of them were chosen in our model selection process.
+@kartoun2017 used a similar penalizing approach and neither select any clinical
+data. Especially in retrospective analysis the clinical data are often not
+detailed enough.
+
+In our benchmark of `r n_lrns` different machine learning algorithms we have
+seen that the more complex algorithm like boosting and neural networks perform
+worse than simpler ones.
+This may due additive effects that favour regression models or due to our
+relative small sample size.
+In machine learning more than 200 events per predictor variable are necessary
+to achieve good prediction performance [@ploeg2014].
+
+<!--
 
@@
     - low sample size, for ML more than 200 events per predictor variable necessary [@ploeg2014]
+- sex-differences and sample-dependencies (by scaling) addressed by *zlog*.
+- more variables, harder to cheat
 - national register required
+-->
 
 ## Limitations
+
+Our study has limitations. Principally it is a single center, retrospective
+analysis and without an external validation cohort.
+The data were collected during the evaluation process for liver transplantation
+at the University Hospital of Leipzig.
+There was no controlled follow-up process. Instead the follow-up information
+were recored at the next clinical admission.
+Furthermore we excluded all patients who had already a liver transplantation
+which could artificial lower the disease severity of the cohort.
+
+With a median MELD score of `r .tbl1("ScoreMeldUnos")` our
+cohort was healthier than in previous studies that reported
+a score of 14-15 [@kim2008; @kartoun2017].
+However the 90-day mortality rate of `r .tbl1("M90")` was comparable to
+the mortality rates seen in earlier studies that ranges from 6 to 30%
+[@malinchoc2000; @kim2008; @kartoun2017].
+Although our cohort was healthier with a medium mortality rate the classical
+MELD score underestimates the observed mortality
+`r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
+times.
+Unfortunately our sample size is very low in the higher MELD categories which
+yield a poor but better than MELD based prediction performance of our model
+for very ill patients.
+
+In conclusion we developed a new model to predict the 90-day survival of
+patients during the evaluation process for liver transplantation.
+Our model outperforms the established MELD-Na and the recently published
+MELD-Plus risk scores.
+For a wider adoption a prospective multi center study for model
+recalibration and validation is needed.
+
+<!--
+Mortality
+[@malinchoc2000] ~ 30 %
+[@kim2008] ~ 6-7 %
+[@kartoun2017] ~ 16.3 % (90 d, post-discharge-mortality)
+
+random time point for blood sampling
+
+It was not possible to ensure that patients died of end-stage liver disease.
+
+The variables we found and the model itself have to be validated and
+recalibrated in a multi center, prospective trial.
+
+Few patients with high MELD scores, poor calibration for high mortality
 
@@
 - bias caused by LTx censorship?
+-->
 

### 2022-06-06 analysis/article.Rmd
refactor: suppress warning for table1

@@
 
-```{r tbl1, results = "asis", echo = FALSE, message = FALSE}
+```{r tbl1, results = "asis", echo = FALSE, message = FALSE, warning = FALSE}
 library("gtsummary")

### 2022-06-15 _targets.R
feat: add variable importance

@@ -59,6 +59,7 @@ source("code/03-mlr-learners.R")
 source("code/04-mlr-benchmarking.R")
 source("code/05-ameld-elasticnet.R")
 source("code/06-timeroc.R")
+source("code/07-variable-importance.R")
 source("code/08-tables.R")
 source("code/09-reports.R")
 source("code/99-site.R")
@@ -73,6 +74,7 @@ list(
    tg_ameld,
    tg_timeroc,
    tg_timeroc_psurv,
+   tg_variable_importance,
    tg_reports,
    tg_site
 )

### 2022-06-15 code/07-variable-importance.R

@@ -1,0 +1,12 @@
+tg_variable_importance <- list(
+    tar_target(rngr, {
+        ranger(
+            x = amelddata$x, y = amelddata$y,
+            importance = "impurity",
+            num.trees = 1000,
+        )
+    },
+    packages = c("ranger", "survival"),
+    deployment = "main"
+    )
+)

### 2022-06-15 _targets.R
chore: continue on errors

@@ -1,13 +1,15 @@
 library("targets")
 library("tarchetypes")
 
+tar_option_set(error = "continue")
+
 library("future")
 library("future.batchtools")
 
 library("lgr")
 
 #Sys.setenv(RUNLOCAL = 1)
-Sys.setenv(DEBUGME = "batchtools")
+#Sys.setenv(DEBUGME = "batchtools")
 tf <- rprojroot::find_rstudio_root_file("logs", "lgr.log")
 lgr$add_appender(AppenderFile$new(tf), name = "file")
 lgr$set_threshold("all")
@@ -26,8 +28,8 @@ login <- future::tweak(
 
 resources <- list(
     workers = 32L,
-    partition = "snowball",
-    ncpu = 36,
+    partition = c("batch", "snowball"),
+    ncpu = 24, #36,
     mcpu = 1L * 1024L,
     walltime = 4L * 60L * 60L # seconds
 )

### 2022-06-15 code/01-data.R
refactor: keep just columns with less than 0.1 NA

@@ -80,8 +80,10 @@ tg_data <- list(
         deployment = "main"
     ),
     tar_target(imp_data, {
+            ## exclude columns with more than 10% missing values
+            keep <- vapply(raw_data, function(x)mean(is.na(x)) < 0.1, NA)
             imp <- withCallingHandlers(
-                impute_df(raw_data, ref_data, method = "logmean"),
+                impute_df(raw_data[keep], ref_data, method = "logmean"),
                 warning = function(w) {
                     if(w$message != paste0(
                         "No reference for column(s): DaysAtRisk, Deceased, ",

### 2022-06-15 code/04-mlr-benchmarking.R
chore: adapt to new mlr3 structure

@@ -3,6 +3,8 @@ suppressPackageStartupMessages(library("mlr3benchmark"))
 suppressPackageStartupMessages(library("mlr3pipelines"))
 suppressPackageStartupMessages(library("mlr3proba"))
 suppressPackageStartupMessages(library("mlr3tuning"))
+suppressPackageStartupMessages(library("mlr3extralearners"))
+suppressPackageStartupMessages(library("mlr3learners"))
 
 tg_benchmarks <- tar_target(
     bmrks,
@@ -19,7 +21,7 @@ tg_benchmarks <- tar_target(
     iteration = "list",
     packages = c(
         "mlr3", "mlr3tuning", "mlr3misc", "mlr3pipelines", "mlr3proba",
-        "mlr3learners"
+        "mlr3learners", "mlr3extralearners"
     ),
     deployment = "worker"
 )

### 2022-06-15 analysis/elasticnet.Rmd
feat: add analysis of complete cases

--- 2022-03-17 analysis/elasticnet.Rmd
@@
 tar_load(bootrcv)
+tar_load(bootrcvcc)
 tar_load(bootarcv)
-#tar_load(bootarcv9)
+tar_load(bootrcv.woIC)
 tar_load(amelddata)
+tar_load(amelddatacc)
 tar_load(ameldcfg)
 tar_load(zlog_data)
+tar_load(zlog_data_complete_cases)
 ```
@@
 
+## rcv.glmnet (complete cases only)
+
+```{r bootrcvcc}
+plot(bootrcvcc, what = "calibration")
+
+ps <- lapply(
+    zlog_data_complete_cases[paste0("SurvProbMeld", c("Unos", "NaUnos", "Plus7"))],
+    function(p) {
+        ctpnts <- cutpoints(p, n = ameldcfg$m)
+        f <- cut(p, ctpnts, include.lowest = TRUE)
+        list(
+            predicted = groupmean(p, f = f),
+            observed = observed_survival(
+                amelddatacc$y, f = f, times = ameldcfg$times
+            )
+        )
+    }
+)
+names(ps) <- c("MELD", "MELD-Na", "MELD-Plus7")
+col <- viridisLite::viridis(6)[4:6]
+
+for (i in seq_along(ps)) {
+    lines(
+        ps[[i]]$predicted, ps[[i]]$observed, col = col[i], type = "b", pch = 19
+    )
+}
+legend("topleft", col = col, legend = names(ps), pch = 19, bty = "n")
+
+plot(bootrcvcc, what = "selected", cex = 0.5)
+plot(bootrcvcc$fit$glmnet.fit, xvar = "norm")
+plot(bootrcvcc$fit$glmnet.fit, xvar = "lambda")
+plot(bootrcvcc$fit$glmnet.fit, xvar = "dev")
+```
+
+## bootrcv.woIC
+
+Exclude IL-6 and CYSC columns from the data set.
+
+```{r bootrcv.woIC}
+plot(bootrcv.woIC, what = "calibration")
+plot(bootrcv.woIC, what = "selected", cex = 0.5)
+```
+
 ## arcv.glmnet
@@
 ```
-
-## arcv9.glmnet
-
-```{r bootarcv9}
-#a <- c(table(sapply(bootarcv9$models, function(m)m$fit$alpha)))
-#plot(bootarcv9 , what = "calibration")
-#plot(bootarcv9, what = "selected")
-#plot_dots(a, main = "Selected Alpha Values")
-```

### 2022-06-15 code/01-data.R

@@ -213,5 +213,12 @@ tg_data <- list(
         },
         packages = "zlog",
         deployment = "main"
+    ),
+    tar_target(zlog_data_complete_cases, {
+            zd <- zlog_data[complete.cases(raw_data),]
+            zd
+        },
+        packages = "zlog",
+        deployment = "main"
     )
 )

### 2022-06-15 code/05-ameld-elasticnet.R

@@ -22,6 +22,18 @@ tg_ameld <- list(
     packages = c("survival"),
     deployment = "main"
     ),
+    tar_target(amelddatacc, {
+        d <- zlog_data_complete_cases
+        ## exclude precalculated scores
+        d <- d[!grepl("Meld", colnames(d))]
+        srv <- Surv(d$DaysAtRisk, d$Deceased)
+        d$DaysAtRisk <- d$Deceased <- NULL
+        d <- data.matrix(d)
+        list(x = d, y = srv)
+    },
+    packages = c("survival"),
+    deployment = "main"
+    ),
     tar_target(arcvob, {
         arcv.glmnet(
             x = amelddata$x, y = amelddata$y,
@@ -41,10 +53,10 @@ tg_ameld <- list(
         bootrcv <- bootstrap(
             x = amelddata$x, y = amelddata$y,
             fun = rcv.glmnet,
+            nboot = ameldcfg$nboot,
+            nrepcv = ameldcfg$nrepcv,
+            nfolds = ameldcfg$nfolds,
             family = "cox",
-            nboot = 1000,
-            nfolds = 3,
-            nrepcv = 10,
             standardize = ameldcfg$standardize,
             m = ameldcfg$m,
             times = ameldcfg$times
@@ -56,9 +68,28 @@ tg_ameld <- list(
     ),
     deployment = "worker"
     ),
-    tar_target(bootrcv.woICA, {
+    tar_target(bootrcvcc, {
+        bootrcv <- bootstrap(
+            x = amelddatacc$x, y = amelddatacc$y,
+            fun = rcv.glmnet,
+            nboot = ameldcfg$nboot,
+            nrepcv = ameldcfg$nrepcv,
+            nfolds = ameldcfg$nfolds,
+            family = "cox",
+            standardize = ameldcfg$standardize,
+            m = ameldcfg$m,
+            times = ameldcfg$times
+        )
+        bootrcv
+    },
+    packages = c(
+        "ameld", "future", "future.batchtools", "glmnet", "survival"
+    ),
+    deployment = "worker"
+    ),
+    tar_target(bootrcv.woIC, {
         x <- amelddata$x[,
-            !colnames(amelddata$x) %in% c("IL6_S", "CYSC_S", "Age")
+            !colnames(amelddata$x) %in% c("IL6_S", "CYSC_S")
         ]
         bootrcv <- bootstrap(
             x = x, y = amelddata$y,
@@ -82,6 +113,13 @@ tg_ameld <- list(
         predict(
[…]

### 2022-06-15 code/06-timeroc.R

@@ -13,7 +13,7 @@ tg_timeroc <- tar_eval(
                 delta = zlog_data$Deceased,
                 marker = 1 - zlog_data[[col]],
                 cause = 1,
-                times = c(1, 3, 7, 30, 60, 90, 180, 365),
+                times = c(2, 7, 30, 90, 180, 365),
                 iid = TRUE
             )
             tr$confint <- confint(object = tr, level = 0.95, n.sim = 3000)
@@ -31,7 +31,37 @@ tg_timeroc_psurv <- list(
             delta = zlog_data$Deceased,
             marker = 1 - psurvbootrcv,
             cause = 1,
-            times = c(1, 3, 7, 30, 60, 90, 180, 365),
+            times = c(2, 7, 30, 90, 180, 365),
+            iid = TRUE
+        )
+        tr$confint <- confint(object = tr, level = 0.95, n.sim = 3000)
+        tr
+    },
+    packages = c("survival", "timeROC"),
+    deployment = "main"
+    ),
+    tar_target(timeROC_RCVcc, {
+        tr <- timeROC::timeROC(
+            T = zlog_data_complete_cases$DaysAtRisk,
+            delta = zlog_data_complete_cases$Deceased,
+            marker = 1 - psurvbootrcvcc,
+            cause = 1,
+            times = c(2, 7, 30, 90, 180, 365),
+            iid = TRUE
+        )
+        tr$confint <- confint(object = tr, level = 0.95, n.sim = 3000)
+        tr
+    },
+    packages = c("survival", "timeROC"),
+    deployment = "main"
+    ),
+    tar_target(timeROC_MELDNacc, {
+        tr <- timeROC::timeROC(
+            T = zlog_data_complete_cases$DaysAtRisk,
+            delta = zlog_data_complete_cases$Deceased,
+            marker = 1 - zlog_data_complete_cases$SurvProbMeldNaUnos,
+            cause = 1,
+            times = c(2, 7, 30, 90, 180, 365),
             iid = TRUE
         )
         tr$confint <- confint(object = tr, level = 0.95, n.sim = 3000)
@@ -46,7 +76,7 @@ tg_timeroc_psurv <- list(
             delta = zlog_data$Deceased,
             marker = 1 - psurvbootarcv,
             cause = 1,
-            times = c(1, 3, 7, 30, 60, 90, 180, 365),
+            times = c(2, 7, 30, 90, 180, 365),
             iid = TRUE
         )
         tr$confint <- confint(object = tr, level = 0.95, n.sim = 3000)

### 2022-06-15 analysis/article.Rmd
feat: further work on the manuscript

@@
 tar_load(ameldcfg)
+tar_load(rngr)
 ```
@@
 the corresponding laboratory diagnostic, respectively.
-Missing laboratory measurements were replaced with a *zlog* value of zero.
+If more than 10 % of laboratory measurements are missing the variable were
+excluded from further analysis.
+The remainder of missing laboratory measurements were replaced with
+a *zlog* value of zero.
 In similar manner if information about complications or
@@
 based on the nonparametric inverse probability of censoring weighting estimator
-(IPCW) [@blanche2013, @R-timeROC].
+(IPCW) [@blanche2013, @R-timeROC]. Testing was applied as described in @blanche2013. A $p < 0.5$ was considered as statistical significant difference.
 
@@
 
-```{r benchmark-plot, fig.cap = "Benchmark results of machine learning algorithms.", echo = FALSE}
-r <- bmrk_results$clone()
-id <- grep("^scale", x = r$learners$learner_id, invert = TRUE, value = TRUE)
-r$filter(task_id = "zlog_eldd", learner_id = id)
-autoplot(r) +
-    geom_boxplot(aes(fill = learner_id)) +
-    geom_jitter(position = position_jitter(0.2)) +
-    scale_fill_viridis(discrete = TRUE) +
-    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+## RCV model
+
+The penalized regression and random forest models
+result in a similar high median performance in the benchmark process.
+In favour of simpler models we chose a lasso based approach for the final model
+development.
+The classical cox model, the data and computational intense algorithms
+extreme gradient boosting, support vector machines and neural networks yield a
+worse performance.
+
+```{r coeftbl, echo = FALSE}
+tar_load(bootrcv)
+cf <- coef(bootrcv$fit)
+cf <- setNames(cf@x, rownames(cf)[cf@i + 1])
+o <- order(-abs(cf))
+cfs <- cf[o]
+cfnms <- lapply(
+    labelled_meld_data[names(cfs)],
+    function(nm)sub(" *\\[.*$", "", attr(nm, "label"))
+)
+cfnms[!lengths(cfnms)] <- names(cfnms)[!lengths(cfnms)]
+knitr::kable(
+    setNames(cfs, cfnms),
+    col.names = "Coefficient",
+    caption = "Model coefficients",
+    digits = 3
+)
 ```
 
-TODO: mention best ml algorithm (\@ref(fig:benchmark-plot))
+## Model comparison
+
+```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE}
+library("timeROC")
+tar_load(timeROC_MELD)
+tar_load(timeROC_MELDNa)
+tar_load(timeROC_MELDPlus7)
+tar_load(timeROC_RCV)
+
+col <- viridisLite::viridis(5)
+m <- list(
+    RCV = timeROC_RCV,
+    MELD = timeROC_MELD,
+    "MELD-Na" = timeROC_MELDNa,
+    "MELD-Plus7" = timeROC_MELDPlus7
+)
+invisible(plot_surv_roc(m, timepoint = 90, col = col[-2]))
+```
+
+```{r roctest}
+tp <- timeROC::compare(
+    timeROC_RCV, timeROC_MELDNa,
+    adjusted = TRUE
+)$p_values_AUC
+rownames(tp) <- paste(
+    c("Non-adjusted", "Adjusted"), "$p$-values"
+)
+knitr::kable(
+    tp,
+    caption = paste0(
+        "Test comparison between AUC for MELD-Na and RCV model ",
+        "according to @blanche2013 at day ",
+        paste0(timeROC_RCV$times, collapse = ", "), ".",
+        "$p$-values are shown as non-adjusted and adjusted for ",
+        "multiple testing."
+    ),
+    digits = 2
+)
+```
+
+```{r bootrcv, fig.width = 7, fig.height = 7, echo = FALSE}
+tar_load(ameldcfg)
+tar_load(amelddata)
+tar_load(zlog_data)
+par(cex = 0.9)
+plot(bootrcv, what = "calibration")
+
+ps <- lapply(
+    zlog_data[paste0("SurvProbMeld", c("Unos", "NaUnos", "Plus7"))],
+    function(p) {
+        ctpnts <- cutpoints(p, n = ameldcfg$m)
+        f <- cut(p, ctpnts, include.lowest = TRUE)
+        list(
+            predicted = groupmean(p, f = f),
+            observed = observed_survival(
+                amelddata$y, f = f, times = ameldcfg$times
+            )
+        )
+    }
+)
+names(ps) <- c("MELD", "MELD-Na", "MELD-Plus7")
+
+for (i in seq_along(ps)) {
+    lines(
+        ps[[i]]$predicted, ps[[i]]$observed, col = col[i + 2],
+        type = "b", pch = 19
+    )
+}
+legend("topleft", col = col[3:5], legend = names(ps), pch = 19, bty = "n")
+```
+
+```{r basehaz, echo = FALSE}
+bh <- basehaz(
+    bootrcv$fit, x = amelddata$x, y = amelddata$y, s = "lambda.1se",
+    times = timeROC_RCV$times, centered = FALSE
+)
+s0 <- setNames(bh$hazard, bh$time)
+k <- knitr::kable(
+    t(s0),
+    digits = 4,
+    caption = paste0(
+        "Underlying survival function; S~0~: baseline survival estimate"
+    )
+)
+k[3] <- paste("|t [days]", k[3])
+k[4] <- paste0("|:------", k[4])
+k[5] <- paste("|S~0~ (t)", k[5])
+k
+```
+
+The 90-day survival estimate of a given patient could be calculated as shown in
+\@ref(eq:survcalc).
+
+```{r survivalcalc, results = "asis"}
+cfnmsltx <- paste0("\\text{", gsub(" ", "\\ ", cfnms), "}")
+cat(
+    "$$\\begin{align}",
+    "S &= S_{0,90}^{\\exp{LP}} \\\\",
+    "&= ", round(s0["90"], 4), "^{\\exp{(",
+        paste0(
+            round(cfs[1], 4), " * zlog(", cfnmsltx[1], ")",
+            paste0(
+                ifelse(sign(cfs[-1]) > 0, "+", ""),
+                round(cfs[-1], 4), " * zlog(", cfnmsltx[-1], ")",
+                collapse = ""
+            ),
+            collapse = ""
+        ),
+    ")}}",
+    #"(\\#eq:survcalc)",
+    "\\end{align}$$"
+)
+```
+
+Where $S_{0}$ is the baseline survival estimate from
+\@ref(tab:basehaz) and $LP$ is the linear predictor of the coefficients
+in \@ref(tab:coeftbl).
 
@@
 
-**NUMBER_OF_VARIABLES**
-
-- more variables, harder to cheat
-- comparison of different predictors of MELD-Plus7 (+MELD-Plus9) vs our features
-
-Sodium is not part of our model.
-In line with the results presented in @kartoun2017 we could show that
-adding sodium to MELD the discrimination performance increases just
-insignificantly.
+Although the Organ Procurement and Transplantation Network recommends the use
+of MELD-Na over MELD the sodium does not add any information to the score.
+Neither in the bootstrap selection nor in the variable importance of a
+random forest model the sodium gains any important weight
+(in bootstrap selection rank
+`r as.vector(rank(-ameld:::.selvars(bootrcv$models, bootrcv$s))["NA_S"])`
+of
+`r ncol(amelddata$x)` and in random forest variable importance
+`r as.vector(rank(-rngr$variable.importance)["NA_S"])` of
+`r ncol(amelddata$x)`,
+see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
+Already @kartoun2017 have shown that the
+discrimination performance of MELD-Na is insignificantly different from the
+classical MELD.
+Consequentially it is not a component of our model.
 
@@
 are less likely getting an urgent listing for liver transplantation
-[@Gaertner2022].
+[@gaertner2022].
 To address sexual differences in laboratory measurements we applied the *zlog*
@@
 Additionally this transformation yield comparable laboratory measurements
-across different clinical laboratories and methods which could improve the
-objectivity of the allocation process.
+across different clinical laboratories and methods.
+It also reduce the influence of age- and sex-related differences,
+which could improve the objectivity of the allocation process.
 The already nearly normal scaled *zlog*-transformed laboratory measurements
@@
 
-Additionally to laboratory data we investigated
+**NUMBER_OF_VARIABLES**
+
+- comparison of different predictors of MELD/MELD-Plus7 (+MELD-Plus9) vs our features
+
+Results from machine learning can aid in finding clinical meaningful patterns
+and guide further research.
+For example, the systemic inflammatory response syndrome is an
+independent risk factor for poor outcome in patients with cirrhosis
+[@thabut2007].
+However neither MELD nor MELD-Na taking inflammation into account.
+In contrast IL6 is the most important variable in our model. Even if IL6 is
+excluded the algorithms select CRP as one of the most important variables which
+highlights the important role of inflammation in the prognosis of patients
+with end-stage liver disease.
+We and other researchers have already shown that adding inflammatory markers,
+namely IL6 or CRP, to the classical MELD score increase the accuracy of the
+90-day mortality prediction [@remmler2018; @cervoni2016].
+
+Beside laboratory measurements we investigated
 `r (ncol(raw_data) - length(cn_labs) - 2)`
@@
 data. Especially in retrospective analysis the clinical data are often not
-detailed enough.
+detailed enough to keep valuable information.
 
@@
 to achieve good prediction performance [@ploeg2014].
+
+<!--
+As we shown above machine learning can aid in finding clinical meaningful
+patterns.
+-->
 
@@
 
-Our study has limitations. Principally it is a single center, retrospective
-analysis and without an external validation cohort.
+Our study has several limitations.
+Principally it is a single center, retrospective analysis and without an
+external validation cohort.
 The data were collected during the evaluation process for liver transplantation
@@
 which could artificial lower the disease severity of the cohort.
+Next most of our patients suffered from ethyltoxic liver cirrhosis
+(`r .tbl1("Ethyltoxic", pattern = "{p}%")`) which may
+limited the transfer to other etiologies.
 
@@
 
+# Data and software availability
+
+The complete code and reproducible analysis could be found at
+https://ampel-leipzig.github.io/ampel-leipzig-meld/.
+Additionally all data are available in the `ameld` R package at
+https://github.com/ampel-leipzig/ameld for further investigations.
+
 # Acknowledgement

### 2022-06-15 analysis/supplement.Rmd
refactor: move common yaml headers into _site.yml
feat: add first discussion draft
feat: further work on the manuscript

@@
 output:
-  workflowr::wflow_html:
+  bookdown::html_document2:
     toc: false
-editor_options:
-  chunk_output_type: console
 ---
 
+```{r knitr_setup, include=FALSE}
+```
+
 ```{r supplement_setup, message = FALSE}
+library("ameld")
+library("data.table")
 library("targets")
+library("mlr3")
+library("mlr3misc")
+library("mlr3proba")
+library("mlr3viz")
+library("viridis")
+library("ggplot2")
 ```
+
+# Benchmark results of machine learning algorithms
+
+```{r benchmark-plot, fig.cap = "Benchmark results of machine learning algorithms.", echo = FALSE, fig.width = 18, fig.height = 12}
+tar_load(bmrk_results)
+r <- bmrk_results$clone()
+id <- grep("^scale", x = r$learners$learner_id, invert = TRUE, value = TRUE)
+r$filter(task_id = "zlog_eldd", learner_id = id)
+autoplot(r) +
+    geom_boxplot(aes(fill = learner_id)) +
+    geom_jitter(position = position_jitter(0.2)) +
+    scale_fill_viridis(discrete = TRUE) +
+    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+```
+
+```{r benchmark-table}
+agg <- bmrk_results$aggregate(msr("surv.cindex", id = "harrell"))
+agg[, `:=`
+    (nr = NULL, resample_result = NULL, resampling_id = NULL, iters = NULL)
+]
+agg <- subset(agg, task_id == "zlog_eldd" & !grepl("^scale", learner_id))
+agg[, task_id := NULL]
+setorder(agg, -harrell)
+knitr::kable(agg, col.names = c("Learner ID", "Harrell C"))
+```
+
+# AUROC trend
+
+```{r roc-plot, fig.width = 18, fig.height = 12}
+tar_load(timeROC_MELD)
+tar_load(timeROC_MELDNa)
+tar_load(timeROC_MELDPlus7)
+tar_load(timeROC_RCV)
+
+m <- list(
+    MELD = timeROC_MELD,
+    "MELD-Na" = timeROC_MELDNa,
+    "MELD-Plus7" = timeROC_MELDPlus7,
+    RCV = timeROC_RCV
+)
+plot_surv_roc_trend(m)
+```
+
+# Variable importance
+
+```{r vimpbs, fig.width = 8, fig.height = 10, fig.cap = "Variable importance by frequency of bootstrap selections.", echo = FALSE, message = FALSE}
+tar_load(bootrcv)
+plot(bootrcv, what = "selected", cex = 0.5)
+```
+
+```{r vimprf, fig.width = 8, fig.height = 10, fig.cap = "Variable importance by logrank in random forest.", echo = FALSE, message = FALSE}
+tar_load(rngr)
+plot_dots(
+    sort(rngr$variable.importance),
+    main = "Variable importance (random forest)",
+    xlab = rngr$splitrule
+)
+```

### 2022-06-15 code/01-data.R
Revert "refactor: keep just columns with less than 0.1 NA"
This reverts commit 56a5dff2d88e4b6e5b795893457602855846bd45.

@@ -80,10 +80,8 @@ tg_data <- list(
         deployment = "main"
     ),
     tar_target(imp_data, {
-            ## exclude columns with more than 10% missing values
-            keep <- vapply(raw_data, function(x)mean(is.na(x)) < 0.1, NA)
             imp <- withCallingHandlers(
-                impute_df(raw_data[keep], ref_data, method = "logmean"),
+                impute_df(raw_data, ref_data, method = "logmean"),
                 warning = function(w) {
                     if(w$message != paste0(
                         "No reference for column(s): DaysAtRisk, Deceased, ",

### 2022-06-15 analysis/corrplot.Rmd
refactor: use zlog_data instead of imp_data

@@
 library("targets")
-tar_load(imp_data)
+tar_load(zlog_data)
 ```
@@
 ```{r corrplot}
-labv <- grep("_[SECQ]$", colnames(imp_data), value = TRUE)
-cr <- cor(imp_data[labv], use = "pairwise.complete.obs")
+labv <- grep("_[SECQ]$", colnames(zlog_data), value = TRUE)
+cr <- cor(zlog_data[labv], method = "pearson", use = "pairwise.complete.obs")
 corrplot.mixed(

### 2022-06-15 analysis/corrplot.Rmd
feat: add spearman correlation

@@
 
-# Correlation
+# Correlation (pearson)
 
@@
 ```
+
+# Correlation (spearman)
+
+## Plot
+
+```{r corrplot-spearman}
+cr <- cor(zlog_data[labv], method = "spearman", use = "pairwise.complete.obs")
+corrplot.mixed(
+    cr,
+    tl.pos = "lt", tl.col = "black", tl.cex = 0.75,
+    number.cex = 0.25, number.font = 1L,
+    order = "AOE"
+)
+```
+
+## Table
+
+```{r cortbl-spearman}
+ut <- upper.tri(cr)
+nms <- paste(
+    colnames(cr)[col(cr)[ut]],
+    rownames(cr)[row(cr)[ut]],
+    sep = ":"
+)
+tbl <- data.frame(Values = nms, Correlation = cr[ut])
+tbl <- tbl[order(-abs(tbl$Correlation)),]
+knitr::kable(tbl, header = TRUE, row.names = FALSE, digits = 3L)
+```

### 2022-06-16 analysis/article.Rmd
feat: further work on the manuscript

@@
 # https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
-title: "A Model to Predict Survival in Patients with End-Stage Liver Disease"
+title: "A New Model to Predict Survival in Patients with End-Stage Liver Disease"
 subtitle: >
@@
 abstract: >
-  **Background**:
-      TODO
+  **Objectives**:
+      The shortage in grafts for liver transplantation requires risk stratification and allocation rules.
+      This study aims to improve the model of end-stage liver disease (MELD) score for 90-day mortality prediction as an risk stratification tool with the help of machine learning (ML) algorithms.
 
   **Methods**:
-      TODO
+      We retrospectively analysed clinical and laboratory data of 654 patients who were recruited during the evaluation process for liver transplantation at the University Hospital Greifswald.
+      After comparing 13 different ML algorithms in a nested cross-validation setting and selecting the best performing one we build a new model to predict 90-day mortality in patients with end-stage liver disease.
 
   **Results**:
-      TODO
+      Penalized regression algorithms yield the highest prediction performance in our ML algorithm benchmark.
+      In favour of a simpler model we chose the least absolute shrinkage and selection operator (lasso) regression.
+      Beside the classical MELD variables INR and bilirubin the lasso regression selected
+      cystatin c over creatinine, and additionally IL-6, total protein and cholinesterase.
+      The new model offers a better discrimination and calibration than MELD and MELD with sodium (MELD-Na) or the MELD-Plus7 risk score.
 
   **Conclusion**:
-      TODO
+      We provide a new model of end-stage liver disease that outperforms the classical MELD score for 90-day survival prediction.
+
+  **Keywords**:
+      end-stage liver disease; MELD; risk stratification; survival estimation; IL-6.
 
@@
 library("ameld")
+library("timeROC")
 
@@
 tar_load(labelled_meld_data)
+tar_load(zlog_data)
 
@@
 tar_load(ameldcfg)
+tar_load(amelddata)
 tar_load(rngr)
+
+tar_load(timeROC_MELD)
+tar_load(timeROC_MELDNa)
+tar_load(timeROC_MELDPlus7)
+tar_load(timeROC_RCV)
 ```
@@
 the corresponding laboratory diagnostic, respectively.
-If more than 10 % of laboratory measurements are missing the variable were
-excluded from further analysis.
-The remainder of missing laboratory measurements were replaced with
-a *zlog* value of zero.
+Missing laboratory measurements were replaced with a *zlog* value of zero.
 In similar manner if information about complications or
@@
 [@tibshirani1997; @glmnet2011; @R-glmnet; @penalized2010; @R-penalized],
-tow different implementations of the
+two different implementations of the
 ridge regression [@glmnet2011; @R-glmnet; @penalized2010; @R-penalized], and
@@
 @harrell1996.
-Briefly we developed our model and predicted survival at `r ameldcfg$times` days.
+Briefly we developed our model using a `r ameldcfg$nfold` cross-validation
+that was repeated `r ameldcfg$nrepcv` times and finally predicted survival
+at `r ameldcfg$times` days.
 The predictions were grouped into equal sized intervals so that there
@@
 The final model was compared against the MELD, MELD-Na and MELD-Plus7 score
-using the area under the time-dependent ROC curve (AUC)
-based on the nonparametric inverse probability of censoring weighting estimator
-(IPCW) [@blanche2013, @R-timeROC]. Testing was applied as described in @blanche2013. A $p < 0.5$ was considered as statistical significant difference.
+using the area under the time-dependent receiver operating characteristic curve
+(AUROC) based on the nonparametric inverse probability of censoring weighting
+estimator (IPCW) [@blanche2013, @R-timeROC].
+Testing was applied as described in @blanche2013.
+A $p < 0.05$ was considered as statistical significant difference.
 
@@
 As shown in Table \@ref(tab:tbl-observed-vs-expected-mortality)
-our observed 90 day mortality was
+our observed 90-day mortality was
 `r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
@@
     caption = paste0(
-        "Observed vs MELD-expected 90 day mortality. ",
+        "Observed vs MELD-expected 90-day mortality. ",
         "MELD mortality values are taken from @wiesner2003. ",
@@
 
-## RCV model
+## AMPEL MELD
 
@@
 
-## Model comparison
-
-```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE}
-library("timeROC")
-tar_load(timeROC_MELD)
-tar_load(timeROC_MELDNa)
-tar_load(timeROC_MELDPlus7)
-tar_load(timeROC_RCV)
-
-col <- viridisLite::viridis(5)
-m <- list(
-    RCV = timeROC_RCV,
-    MELD = timeROC_MELD,
-    "MELD-Na" = timeROC_MELDNa,
-    "MELD-Plus7" = timeROC_MELDPlus7
-)
-invisible(plot_surv_roc(m, timepoint = 90, col = col[-2]))
-```
-
-```{r roctest}
-tp <- timeROC::compare(
-    timeROC_RCV, timeROC_MELDNa,
-    adjusted = TRUE
-)$p_values_AUC
-rownames(tp) <- paste(
-    c("Non-adjusted", "Adjusted"), "$p$-values"
-)
-knitr::kable(
-    tp,
-    caption = paste0(
-        "Test comparison between AUC for MELD-Na and RCV model ",
-        "according to @blanche2013 at day ",
-        paste0(timeROC_RCV$times, collapse = ", "), ".",
-        "$p$-values are shown as non-adjusted and adjusted for ",
-        "multiple testing."
-    ),
-    digits = 2
-)
-```
-
-```{r bootrcv, fig.width = 7, fig.height = 7, echo = FALSE}
-tar_load(ameldcfg)
-tar_load(amelddata)
-tar_load(zlog_data)
-par(cex = 0.9)
-plot(bootrcv, what = "calibration")
-
-ps <- lapply(
-    zlog_data[paste0("SurvProbMeld", c("Unos", "NaUnos", "Plus7"))],
-    function(p) {
-        ctpnts <- cutpoints(p, n = ameldcfg$m)
-        f <- cut(p, ctpnts, include.lowest = TRUE)
-        list(
-            predicted = groupmean(p, f = f),
-            observed = observed_survival(
-                amelddata$y, f = f, times = ameldcfg$times
-            )
-        )
-    }
-)
-names(ps) <- c("MELD", "MELD-Na", "MELD-Plus7")
-
-for (i in seq_along(ps)) {
-    lines(
-        ps[[i]]$predicted, ps[[i]]$observed, col = col[i + 2],
-        type = "b", pch = 19
-    )
-}
-legend("topleft", col = col[3:5], legend = names(ps), pch = 19, bty = "n")
-```
+The resulting model AMPEL model of end-stage liver disease
+(AMELD) uses `r length(cfs)` predictors.
+Beside the classical MELD variables INR and (direct) bilirubin the lasso
+regression selected cystatin c over creatinine, and
+additionally IL-6, total protein and cholinesterase.
 
@@
         ),
-    ")}}",
-    #"(\\#eq:survcalc)",
-    "\\end{align}$$"
+    ")}}"
 )
 ```
+(\#eq:survcalc)
+\end{align}
+$$
+<!-- using eq:survcalc inside the R code chunk impedes Math ML rendering -->
 
@@
 in \@ref(tab:coeftbl).
+
+## Model comparison
+
+```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE}
+col <- viridisLite::viridis(5)
+m <- list(
+    AMELD = timeROC_RCV,
+    MELD = timeROC_MELD,
+    "MELD-Na" = timeROC_MELDNa,
+    "MELD-Plus7" = timeROC_MELDPlus7
+)
+invisible(roc <- plot_surv_roc(m, timepoint = 90, col = col[-2]))
+roc <- roc[order(-roc)]
+```
+
+```{r roctest}
+tp <- timeROC::compare(
+    timeROC_RCV, timeROC_MELDNa,
+    adjusted = TRUE
+)$p_values_AUC
+rownames(tp) <- paste(
+    c("Non-adjusted", "Adjusted"), "$p$-values"
+)
+knitr::kable(
+    tp,
+    caption = paste0(
+        "Test comparison between AUCROC for MELD-Na and AMELD",
+        "according to @blanche2013 at day ",
+        paste0(timeROC_RCV$times, collapse = ", "), ".",
+        "$p$-values are shown as non-adjusted and adjusted for ",
+        "multiple testing."
+    ),
+    digits = 2
+)
+```
+
+AMELD offers the best AUROC (`r round(roc["AMELD"], 3)`) across all competitors.
+However, the difference against the second best model `r names(roc)[2]`
+(AUROC: `r round(roc[2], 3)`) is not statistical significant (
+`r paste0("$p = ", round(tp[2, "t=90"], 2), "$")`) after correction for multiple
+testing (see Table \@ref(tab:roctest) for complete comparison).
+
+```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = "Calibration of AMELD in comparison to MELD, MELD-Na and MELD-Plus7.", echo = FALSE}
+par(cex = 0.9)
+plot(
+    bootrcv,
+    what = "calibration",
+    main = "90-day survival",
+    legend = FALSE
+)
+
+ps <- lapply(
+    zlog_data[paste0("SurvProbMeld", c("Unos", "NaUnos", "Plus7"))],
+    function(p) {
+        ctpnts <- cutpoints(p, n = ameldcfg$m)
+        f <- cut(p, ctpnts, include.lowest = TRUE)
+        list(
+            predicted = groupmean(p, f = f),
+            observed = observed_survival(
+                amelddata$y, f = f, times = ameldcfg$times
+            )
+        )
+    }
+)
+names(ps) <- c("MELD", "MELD-Na", "MELD-Plus7")
+
+for (i in seq_along(ps)) {
+    lines(
+        ps[[i]]$predicted, ps[[i]]$observed, col = col[i + 2],
+        type = "b", pch = 19
+    )
+}
+legend(
+    "topleft",
+    col = col,
+    legend = c(
+        "AMELD",
+        paste0(
+            "AMELD, optimism error added (based on ",
+            length(bootrcv$models),
+            " bootstrap samples)"
+        ),
+        names(ps)
+    ),
+    pch = c(19, 4, rep(19, 3)),
+    bty = "n"
+)
+```
+
+Fig. \@ref(fig:bootrcv) depicts the calibration curve for AMELD and the
+competing MELD scores.
+While the prediction for low-risk patients is similar to other scores
+AMELD predicts the medium-risk patients more accurate.
+Because of the low number of very ill patients in our cohort the prediction
+performance in high-risk patients is poor but comparable to the other models.
+This would require a recalibration in a larger validation cohort.
 
@@
 the University Hospital of Leipzig we developed a new model to predict the
-90-day survival probability. Our new model outperforms the MELD-Na score and
-the recently published MELD-Plus risk score.
+90-day survival probability. Our new model, AMELD, outperforms the MELD-Na
+score and the recently published MELD-Plus7 risk score.
 
+```{r selvar, echo = FALSE}
+bsrank <- as.vector(rank(-ameld:::.selvars(bootrcv$models, bootrcv$s), ties = "first")["NA_S"])
+
+bsranktxt <-
+    if (!is.na(bsrank)) {
+        paste0("in bootstrap selection rank ", bsrank, " of ", ncol(amelddata$x))
+    } else {
+        paste0("it was never selected in any boostrap sample")
+    }
+
+rfrank <- as.vector(rank(-rngr$variable.importance, ties = "first")["NA_S"])
+rfranktxt <-
+    paste0("in a random forest variable importance rank ", rfrank, " of ", ncol(amelddata$x))
+```
 Although the Organ Procurement and Transplantation Network recommends the use
@@
 random forest model the sodium gains any important weight
-(in bootstrap selection rank
-`r as.vector(rank(-ameld:::.selvars(bootrcv$models, bootrcv$s))["NA_S"])`
-of
-`r ncol(amelddata$x)` and in random forest variable importance
-`r as.vector(rank(-rngr$variable.importance)["NA_S"])` of
-`r ncol(amelddata$x)`,
-see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
+(`r bsranktxt` and `r rfranktxt`, see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
 Already @kartoun2017 have shown that the
@@
 
-**NUMBER_OF_VARIABLES**
+Our model, AMELD, uses `r length(cfs)` predictors.
+Half of them, INR, (direct) bilirubin, and cystatin c are identical
+or similar to the labortory measurements utilized in the classical MELD.
+The direct bilirubin selected by our approach and the total bilirubin in MELD
+are highly correlated (spearman correlation:
+`r round(cor(amelddata$x, method = "spearman")["BILI_S", "BILID_S"])`) and
+therefore interchangeable.
 
-- comparison of different predictors of MELD/MELD-Plus7 (+MELD-Plus9) vs our features
+In our feature selection process cystatin C was chosen instead of creatinine.
+This is inline with recent studies that show a better prediction of mortality
+for cystatin C in unselected and critical ill patients
+[@helmerssonkarlqvist2016; @helmerssonkarlqvist2021]
 
-Results from machine learning can aid in finding clinical meaningful patterns
-and guide further research.
-For example, the systemic inflammatory response syndrome is an
+Furthermore AMELD extends MELD by additional synthesis parameters
+total protein and cholinesterase and
+most importantly the inflammatory marker IL-6.
+
+The latter is an example where results from machine learning can aid in
+finding clinical meaningful patterns and guide further research.
+The systemic inflammatory response syndrome is an
 independent risk factor for poor outcome in patients with cirrhosis
@@
 However neither MELD nor MELD-Na taking inflammation into account.
-In contrast IL6 is the most important variable in our model. Even if IL6 is
+In contrast IL-6 is the most important variable in our model. Even if IL-6 is
 excluded the algorithms select CRP as one of the most important variables which
@@
 We and other researchers have already shown that adding inflammatory markers,
-namely IL6 or CRP, to the classical MELD score increase the accuracy of the
+namely IL-6 or CRP, to the classical MELD score increase the accuracy of the
 90-day mortality prediction [@remmler2018; @cervoni2016].
@@
 clinical data but none of them were chosen in our model selection process.
-@kartoun2017 used a similar penalizing approach and neither select any clinical
+@kartoun2017 also used the lasso approach and neither select any clinical
 data. Especially in retrospective analysis the clinical data are often not
@@
 In our benchmark of `r n_lrns` different machine learning algorithms we have
-seen that the more complex algorithm like boosting and neural networks perform
-worse than simpler ones.
-This may due additive effects that favour regression models or due to our
+seen that the more complex algorithm like boosting, support vector machines
+and neural networks perform worse than simpler ones.
+This may due additive effects that favor regression models or due to our
 relative small sample size.
-In machine learning more than 200 events per predictor variable are necessary
-to achieve good prediction performance [@ploeg2014].
+It has been shown that in machine learning more than 200 events per
+predictor variable are required to achieve good prediction performance
+[@ploeg2014].
 
@@
 times.
-Unfortunately our sample size is very low in the higher MELD categories which
-yield a poor but better than MELD based prediction performance of our model
-for very ill patients.
+Unfortunately our sample size is very low in the high risk patient category
+which yield a poor but better than MELD based prediction performance of our
+AMELD for very ill patients.
+
+## Conclusion
 
@@
 patients during the evaluation process for liver transplantation.
-Our model outperforms the established MELD-Na and the recently published
-MELD-Plus risk scores.
+Our model, AMELD, extends the classical MELD predictors, INR,
+(direct) bilirubin, cystatin c (instead of creatinine) by
+synthesis parameters total protein and cholinesterase and
+most importantly the inflammatory marker IL-6.
+Using these `r length(cfs)` laboratory measurments the AMELD
+outperforms the established MELD-Na and the recently published
+MELD-Plus7 risk scores.
 For a wider adoption a prospective multi center study for model
@@
 -->
+# Data and software availability
+
+**TODO** add zenodo links here.
+The complete code and reproducible analysis could be found at
+https://ampel-leipzig.github.io/ampel-leipzig-meld/.
+Additionally all data are available in the `ameld` R package at
+https://github.com/ampel-leipzig/ameld for further investigations.
+
+# Acknowledgement
+
+The study was supported by the AMPEL project (www.ampel.care).
+This project is co-financed through public funds according to the budget
+decided by the Saxon State Parliament under the
+RL eHealthSax 2017/18 grant number 100331796.
+The funder provided support in the form of
+salaries for author SG
+but did not have any additional role in the study design,
+data collection and analysis, decision to publish, or preparation
+of the manuscript.
+We thank the whole AMPEL team for their support,
+and Stefan Kemnitz for his technical support regarding the use of the
+HPC cluster of the University Greifswald.
 
@@
 A single run of the complete calculation pipeline took roughly
-`r sprintf("%.1f", sum(tm$runtime) / 3600)` hours of computation on an
+`r round(sum(tm$runtime) / 3600)` hours of computation on an
 Intel(R) Xeon(R) E5-2650 v4 and
@@
 (ignoring development, test runs, etc.).
-This is equivalent to `r sprintf("%.1f", km)` km driven by an average car
+This is equivalent to `r round(km)` km driven by an average car
 [@allekotte2021].
 
-# Data and software availability
-
-The complete code and reproducible analysis could be found at
-https://ampel-leipzig.github.io/ampel-leipzig-meld/.
-Additionally all data are available in the `ameld` R package at
-https://github.com/ampel-leipzig/ameld for further investigations.
-
-# Acknowledgement
-
-The study was supported by the AMPEL project (www.ampel.care).
-This project is co-financed through public funds according to the budget
-decided by the Saxon State Parliament under the
-RL eHealthSax 2017/18 grant number 100331796.
-The funder provided support in the form of
-salaries for author SG
-but did not have any additional role in the study design,
-data collection and analysis, decision to publish, or preparation
-of the manuscript.
-We thank the whole AMPEL team for their support,
-and Stefan Kemnitz for his technical support regarding the use of the
-HPC cluster of the University Greifswald.
-
 # References

### 2022-06-16 code/05-ameld-elasticnet.R
refactor: reduce number of repeated cross val

@@ -2,7 +2,7 @@ tg_ameld <- list(
     tar_target(ameldcfg, list(
         alpha = seq(0, 1, len = 11)^3,
         nboot = 200L,
-        nrepcv = 100L,
+        nrepcv = 10L,
         nfolds = 3L,
         m = 50,
         times = 90,

### 2022-06-16 analysis/article.Rmd
fix: don't echo code chunks

@@
 
-```{r survivalcalc, results = "asis"}
+```{r survivalcalc, results = "asis", echo = FALSE}
 cfnmsltx <- paste0("\\text{", gsub(" ", "\\ ", cfnms), "}")
@@
 
-```{r roctest}
+```{r roctest, echo = FALSE}
 tp <- timeROC::compare(

### 2022-06-19 analysis/article.Rmd
fix: cite zlog with zenodo doi

@@
   - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'rpackages', 'article.bib')`"
+  - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'rpackages', 'zlog.bib')`"
 output:
@@
         "gtsummary",
-        "mlr3", "mlr3proba", "zlog",
+        "mlr3", "mlr3proba",
         "survival", "glmnet", "penalized", "ranger", "randomForestSRC",
@@
         "analysis", "bibliography", "rpackages", "article.bib"
+    )
+)
+cat(
+    gsub("@Manual\\{", "@Manual\\{R-zlog", toBibtex(citation("zlog"))),
+    file = rprojroot::find_rstudio_root_file(
+        "analysis", "bibliography", "rpackages", "zlog.bib"
     )

### 2022-06-19 analysis/article.Rmd
feat: cite ameld with zenodo doi

--- 2022-06-16 analysis/article.Rmd
@@
   - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'rpackages', 'article.bib')`"
+  - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'rpackages', 'ameld.bib')`"
+  - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'rpackages', 'zlog.bib')`"
 output:
@@
         "analysis", "bibliography", "rpackages", "article.bib"
+    )
+)
+cat(
+    gsub("@Manual\\{", "@Manual\\{R-ameld", toBibtex(citation("ameld"))),
+    file = rprojroot::find_rstudio_root_file(
+        "analysis", "bibliography", "rpackages", "ameld.bib"
+    )
+)
+cat(
+    gsub("@Manual\\{", "@Manual\\{R-zlog", toBibtex(citation("zlog"))),
+    file = rprojroot::find_rstudio_root_file(
+        "analysis", "bibliography", "rpackages", "zlog.bib"
     )
@@
 -->
+
 # Data and software availability
 
-**TODO** add zenodo links here.
 The complete code and reproducible analysis could be found at
 https://ampel-leipzig.github.io/ampel-leipzig-meld/.
-Additionally all data are available in the `ameld` R package at
-https://github.com/ampel-leipzig/ameld for further investigations.
+Additionally all data are available in the `ameld` R package [@R-ameld] at
+https://github.com/ampel-leipzig/ameld
+[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.6666018.svg)](https://doi.org/10.5281/zenodo.6666018)
+for further investigations.
 

### 2022-06-19 analysis/elasticnet.Rmd
feat: use plot(..., what = "path") introduced in ameld 0.0.25

@@
 plot(bootrcv, what = "selected", cex = 0.5)
-plot(bootrcv$fit$glmnet.fit, xvar = "norm")
-plot(bootrcv$fit$glmnet.fit, xvar = "lambda")
-plot(bootrcv$fit$glmnet.fit, xvar = "dev")
+plot(bootrcv$fit, what = "path", xvar = "norm", nlabel = 14, cex.lab = 0.5)
+plot(bootrcv$fit, what = "path", xvar = "lambda", nlabel = 14, cex.lab = 0.5)
+plot(bootrcv$fit, what = "path", xvar = "dev", nlabel = 14, cex.lab = 0.5)
 ```
@@
 plot(bootrcvcc, what = "selected", cex = 0.5)
-plot(bootrcvcc$fit$glmnet.fit, xvar = "norm")
-plot(bootrcvcc$fit$glmnet.fit, xvar = "lambda")
-plot(bootrcvcc$fit$glmnet.fit, xvar = "dev")
+plot(bootrcvcc$fit, what = "path", xvar = "norm", nlabel = 14, cex.lab = 0.5)
+plot(bootrcvcc$fit, what = "path", xvar = "lambda", nlabel = 14, cex.lab = 0.5)
+plot(bootrcvcc$fit, what = "path", xvar = "dev", nlabel = 14, cex.lab = 0.5)
 ```

### 2022-07-03 analysis/article.Rmd
fix: cystatin c

@@
       Beside the classical MELD variables INR and bilirubin the lasso regression selected
-      cystatin c over creatinine, and additionally IL-6, total protein and cholinesterase.
+      cystatin C over creatinine, and additionally IL-6, total protein and cholinesterase.
       The new model offers a better discrimination and calibration than MELD and MELD with sodium (MELD-Na) or the MELD-Plus7 risk score.
@@
 Beside the classical MELD variables INR and (direct) bilirubin the lasso
-regression selected cystatin c over creatinine, and
+regression selected cystatin C over creatinine, and
 additionally IL-6, total protein and cholinesterase.
@@
 Our model, AMELD, uses `r length(cfs)` predictors.
-Half of them, INR, (direct) bilirubin, and cystatin c are identical
+Half of them, INR, (direct) bilirubin, and cystatin C are identical
 or similar to the labortory measurements utilized in the classical MELD.
@@
 Our model, AMELD, extends the classical MELD predictors, INR,
-(direct) bilirubin, cystatin c (instead of creatinine) by
+(direct) bilirubin, cystatin C (instead of creatinine) by
 synthesis parameters total protein and cholinesterase and

### 2022-07-03 analysis/article.Rmd
refactor: integrate Bernd's suggestions

@@
 [@R-base].
-Prior to the analyses all laboratory values were *zlog* transformed
-as described in @hoffmann2017 and implemented in @R-zlog.
+Prior to the analyses all laboratory values were *zlog* transformed to
+approximate a normal distribution as described in
+@hoffmann2017 and implemented in @R-zlog.
 
@@
 
-To choose the best model we used nested resampling to avoid feature- and
-model-selection-bias.
+We used nested resampling to avoid feature- and
+model-selection-bias to get reliable metrics to choose the best model.
 Therefore, we applied a `r crossval$inner$param_set$values$folds`-fold inner
@@
 `r tuner$param_set$get_values()$resolution`.
-Subsequently we selected the model with the
-highest median concordance index [@harrell1982]. If multiple models were
-comparable we chose the simpler (more penalized) one.
+The models were evaluated using the concordance index, which is a measure of
+rank correlation between predicted and observed events [@harrell1982].
+Subsequently we selected the model with the highest median concordance index.
+If multiple models were comparable we chose the simpler (more penalized) one.
 
@@
 at `r ameldcfg$times` days.
-The predictions were grouped into equal sized intervals so that there
+The predictions were sorted and grouped into equal sized intervals so that there
 are `r ameldcfg$m` patients in each interval.
@@
 )
-s0 <- setNames(bh$hazard, bh$time)
+s0 <- setNames(1 - bh$hazard, bh$time)
 k <- knitr::kable(
@@
     caption = paste0(
-        "Test comparison between AUCROC for MELD-Na and AMELD",
+        "Test comparison between AUCROC for MELD-Na and AMELD ",
         "according to @blanche2013 at day ",
@@
 
-```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = "Calibration of AMELD in comparison to MELD, MELD-Na and MELD-Plus7.", echo = FALSE}
+```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = paste('Calibration of AMELD in comparison to MELD, MELD-Na and MELD-Plus7.\n The points mark the mean predicted survival of', ameldcfg$m, 'patients per interval against their observed mean survival.
+The error bars correspond to the 95 % confidence interval of the survival estimate.'), echo = FALSE}
 par(cex = 0.9)
@@
 
-The study was supported by the AMPEL project (www.ampel.care).
+The study was supported by the AMPEL project ([www.ampel.care](https://www.ampel.care)).
 This project is co-financed through public funds according to the budget
@@
 We thank the whole AMPEL team for their support,
-and Stefan Kemnitz for his technical support regarding the use of the
-HPC cluster of the University Greifswald.
+Stefan Kemnitz for his technical support regarding the use of the
+HPC cluster of the University Greifswald and
+Bernd Klaus for helpful discussions about survival analysis and proof reading the manuscript.
 

### 2022-07-03 analysis/article.Rmd
feat: round spearman correlation to 2 digits

@@
 are highly correlated (spearman correlation:
-`r round(cor(amelddata$x, method = "spearman")["BILI_S", "BILID_S"])`) and
+`r round(cor(amelddata$x, method = "spearman")["BILI_S", "BILID_S"], 2)`) and
 therefore interchangeable.

### 2022-07-08 analysis/article.Rmd
fix: change center to Leipzig

@@
   **Methods**:
-      We retrospectively analysed clinical and laboratory data of 654 patients who were recruited during the evaluation process for liver transplantation at the University Hospital Greifswald.
+      We retrospectively analysed clinical and laboratory data of 654 patients who were recruited during the evaluation process for liver transplantation at the University Hospital Leipzig.
       After comparing 13 different ML algorithms in a nested cross-validation setting and selecting the best performing one we build a new model to predict 90-day mortality in patients with end-stage liver disease.

### 2022-07-17 analysis/article.Rmd
feat: add machine learning to keywords
refactor: integrate Thorsten's review comments

@@
   **Objectives**:
-      The shortage in grafts for liver transplantation requires risk stratification and allocation rules.
-      This study aims to improve the model of end-stage liver disease (MELD) score for 90-day mortality prediction as an risk stratification tool with the help of machine learning (ML) algorithms.
+      The shortage in grafts for liver transplantation requires
+      risk stratification and adequate allocation rules.
+      This study aims to improve the model of end-stage liver disease (MELD)
+      score for 90-day mortality prediction with the help of different
+      machine learning (ML) algorithms.
 
   **Methods**:
-      We retrospectively analysed clinical and laboratory data of 654 patients who were recruited during the evaluation process for liver transplantation at the University Hospital Leipzig.
-      After comparing 13 different ML algorithms in a nested cross-validation setting and selecting the best performing one we build a new model to predict 90-day mortality in patients with end-stage liver disease.
+      We retrospectively analyzed clinical and laboratory data of 654
+      patients who were recruited during the evaluation process for
+      liver transplantation at the University Hospital Leipzig.
+      After comparing 13 different ML algorithms in a nested cross-validation
+      setting and selecting the best performing one we build a new model
+      to predict 90-day mortality in patients with end-stage liver disease.
 
   **Results**:
-      Penalized regression algorithms yield the highest prediction performance in our ML algorithm benchmark.
-      In favour of a simpler model we chose the least absolute shrinkage and selection operator (lasso) regression.
-      Beside the classical MELD variables INR and bilirubin the lasso regression selected
-      cystatin C over creatinine, and additionally IL-6, total protein and cholinesterase.
-      The new model offers a better discrimination and calibration than MELD and MELD with sodium (MELD-Na) or the MELD-Plus7 risk score.
+      Penalized regression algorithms yield the highest prediction performance
+      in our ML algorithm benchmark.
+      In favour of a simpler model we chose the least absolute shrinkage and
+      selection operator (lasso) regression.
+      Beside the classical MELD variables INR and bilirubin the
+      lasso regression selected cystatin C over creatinine, and
+      additionally IL-6, total protein and cholinesterase.
+      The new model offers an improved discrimination and calibration than MELD
+      and MELD with sodium (MELD-Na) or the MELD-Plus7 risk score.
 
   **Conclusion**:
-      We provide a new model of end-stage liver disease that outperforms the classical MELD score for 90-day survival prediction.
+      We provide a new ML based model of end-stage liver disease that
+      outperforms the classical MELD score for 90-day survival prediction.
 
   **Keywords**:
-      end-stage liver disease; MELD; risk stratification; survival estimation; IL-6.
+      end-stage liver disease; MELD; risk stratification; machine learning;
+      survival estimation; IL-6.
 
@@
 Liver cirrhosis is the terminal result of fibrotic remodeling
-of liver tissue due to chronic damage. An organ failure is often
-irreversible and the only available therapy is liver transplantation.
+of liver tissue due to chronic damage.
+This end-stage of liver disease is usually irreversible and
+the only available therapy is liver transplantation.
 However, the shortage in grafts for transplantation from deceased donors
-requires risk stratification and precise allocation rules.
+requires risk stratification and precise and fair allocation rules.
 The allocation of liver transplantation in most countries is based on
@@
 laboratory results, namely bilirubin, creatinine and international normalized
-ratio (INR). In general the MELD score is extended by the sodium level
+ratio (INR).
+The MELD score has been extended by the sodium level
 (MELD-Na score) as this was found to be an important additional risk factor in
@@
 AMELD offers the best AUROC (`r round(roc["AMELD"], 3)`) across all competitors.
-However, the difference against the second best model `r names(roc)[2]`
-(AUROC: `r round(roc[2], 3)`) is not statistical significant (
-`r paste0("$p = ", round(tp[2, "t=90"], 2), "$")`) after correction for multiple
-testing (see Table \@ref(tab:roctest) for complete comparison).
+However, after correction for multiple testing the difference against
+the second best model `r names(roc)[2]`
+(AUROC: `r round(roc[2], 3)`) is not statistical significant
+(`r paste0("$p = ", round(tp[2, "t=90"], 2), "$")`)
+(see Table \@ref(tab:roctest) for complete comparison).
 
@@
 (`r bsranktxt` and `r rfranktxt`, see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
-Already @kartoun2017 have shown that the
-discrimination performance of MELD-Na is insignificantly different from the
-classical MELD.
+In addition sodium levels can be affected by diuretic therapy and already
+@kartoun2017 have shown that the discrimination performance of MELD-Na is
+insignificantly different from the classical MELD.
 Consequentially it is not a component of our model.
 
-It has shown that women get lower scores in MELD-Na and therefore
+It has shown that women receive in median lower scores in MELD-Na and therefore
 are less likely getting an urgent listing for liver transplantation
@@
 To address sexual differences in laboratory measurements we applied the *zlog*
-transformation [@hoffmann2017].
+transformation [@hoffmann2017] in our study.
 Additionally this transformation yield comparable laboratory measurements
@@
 at the University Hospital of Leipzig.
-There was no controlled follow-up process. Instead the follow-up information
-were recored at the next clinical admission.
-Furthermore we excluded all patients who had already a liver transplantation
+There was no controlled follow-up process.
+We excluded all patients who had already a liver transplantation
 which could artificial lower the disease severity of the cohort.
-Next most of our patients suffered from ethyltoxic liver cirrhosis
+Furthermore most of our patients suffered from ethyltoxic liver cirrhosis
 (`r .tbl1("Ethyltoxic", pattern = "{p}%")`) which may
@@
 times.
-Unfortunately our sample size is very low in the high risk patient category
+Unfortunately our sample size is low in the high risk patient category
 which yield a poor but better than MELD based prediction performance of our
@@
 
-The calculations were conducted using the high performance computing (HPC)
+The calculations were conducted using the high-performance computing (HPC)
 cluster of the University Greifswald.

### 2022-07-18 analysis/article.Rmd
refactor: replace gaertner2022 by sealock2022

@@
 creatinine and INR are highly dependent on the utilized laboratory methods
-[@trotter2004; @cholongitas2007]. Patients with identical disease state could
-have very different MELD scores and thus get different priority on the liver
-transplantation waiting list.
+[@trotter2004; @cholongitas2007].
+Additionally women are disadvantage [@sealock2022].
+Patients with identical disease state could have very different MELD scores and
+thus get different priority on the liver transplantation waiting list.
 Furthermore often, e.g. for acute-on-chronic liver failure, the MELD score
@@
 are less likely getting an urgent listing for liver transplantation
-[@gaertner2022].
+[@sealock2022].
 To address sexual differences in laboratory measurements we applied the *zlog*

### 2022-07-18 analysis/article.Rmd
feat: add vancouver csl

@@
       - --lua-filter=pandoc/lua-filters/author-info-blocks.lua
+      - --csl=pandoc/csl/vancouver.csl
 ---

### 2022-07-18 analysis/article.Rmd
refactor: drop IQR for median MELD score in limitations

@@
 
-With a median MELD score of `r .tbl1("ScoreMeldUnos")` our
-cohort was healthier than in previous studies that reported
+With a median MELD score of `r .tbl1("ScoreMeldUnos", pattern = "{median}")`
+our cohort was healthier than in previous studies that reported
 a score of 14-15 [@kim2008; @kartoun2017].

### 2022-07-18 analysis/article.Rmd
chore: add additional authors

@@
         correspondence: true
+    - Thomas Berg:
+        email: thomas.berg@medizin.uni-leipzig.de
+        institute: KIMH
+        correspondence: false
+    - Daniel Seehofer:
+        email: daniel.seehofer@medizin.uni-leipzig.de
+        institute: KVTTG
+        correspondence: false
+    - Berend Isermann:
+        email: berend.isermann@medizin.uni-leipzig.de
+        institute: ILM
+        correspondence: false
     - Thorsten Kaiser:
@@
     - ILM: Institute of Laboratory Medicine, Clinical Chemistry and Molecular Diagnostics, University Hospital Leipzig, Paul-List-Str. 13-15, D-04103 Leipzig, Germany.
-    - KAINS: Anesthesiology and Intensive Care Medicine, University Hospital Greifswald, Ferdinand-Sauerbruch-Straße, D-17475 Greifswald, Germany.
+    - KIMH: Section of Hepatology, Department of Gastroenterology and Rheumatology, University Hospital Leipzig, Liebigstr. 20, D-04103 Leipzig, Germany.
+    - KVTTG: Department of Visceral, Transplantation, Vascular and Thoracic Surgery, University Hospital of Leipzig, Liebigstr. 20, D-04103 Leipzig, Germany.
+    - KAINS: Anesthesiology and Intensive Care Medicine, University Hospital Greifswald, Ferdinand-Sauerbruch-Str., D-17475 Greifswald, Germany.
 date: "`r Sys.Date()`"

### 2022-07-18 analysis/article.Rmd
feat: add author contributions and competing interests section

@@
 
-The study was supported by the AMPEL project ([www.ampel.care](https://www.ampel.care)).
-This project is co-financed through public funds according to the budget
-decided by the Saxon State Parliament under the
-RL eHealthSax 2017/18 grant number 100331796.
-The funder provided support in the form of
-salaries for author SG
-but did not have any additional role in the study design,
-data collection and analysis, decision to publish, or preparation
-of the manuscript.
 We thank the whole AMPEL team for their support,
@@
 
+# Research funding
+
+The study was supported by the AMPEL project ([www.ampel.care](https://www.ampel.care)).
+This project is co-financed through public funds according to the budget
+decided by the Saxon State Parliament under the
+RL eHealthSax 2017/18 grant number 100331796.
+The funder provided support in the form of
+salaries for author SG
+but did not have any additional role in the study design,
+data collection and analysis, decision to publish, or preparation
+of the manuscript.
+
+# Author contributions
+
+Funding was acquired by TK.
+The study was conceptualised by SG and TK.
+The data was acquired by TK.
+The data was analysed and processed by SG.
+Visualisation were produced by SG.
+Original drafts of the manuscript were written by SG and TK.
+Drafts were edited and reviewed by SG, TB, DS, BI, and TK.
+The project was supervised by TK.
+All authors have accepted responsibility for the entire content of this
+manuscript and approved its submission.
+
+# Competing interests
+
+Authors state no conflict of interest.
+
+# Informed consent
+
+Informed consent was obtained from all individuals included in this study.
+
+# Ethical approval
+
+Research involving human subjects complied with all relevant national
+regulations, institutional policies and is in accordance with the tenets of
+the Helsinki Declaration (as revised in 2013), and has been
+approved by the Ethics Committee at the Leipzig University Faculty of
+Medicine (reference number: 039/14ff).
+
 # References

### 2022-07-18 code/01-data.R
refactor: labm requires uppercase litre

@@ -116,9 +116,15 @@ tg_data <- list(
             ]
             abbr <- setNames(character(ncol(x)), colnames(x))
 
+            .rpl_litre <- function(x, upper = TRUE) {
+                if (upper)
+                    sub("([fµmcd]?)[lL]$", "\\1L", x)
+                else
+                    x
+            }
             .lbl <- function(x, unit) {
                 if (nchar(unit))
-                    paste0(x, " [", unit, "]")
+                    paste0(x, " [", .rpl_litre(unit, upper = TRUE), "]")
                 else
                     x
             }

### 2022-07-18 analysis/article.Rmd
feat: use 1200 dpi as required by labm for word output

@@
   bookdown::html_document2:
+    keep_md: true
     pandoc_args: # pandoc_args doesn't support r evaluation
@@
   bookdown::word_document2:
+    keep_md: true
     pandoc_args: # pandoc_args doesn't support r evaluation
@@
 
-```{r knitr_setup, include=FALSE}
+```{r knitr_setup, include = FALSE}
 knitr::write_bib(
@@
 )
+```
+
+```{r knitr_setup_word, eval = !knitr::is_html_output(), include = FALSE}
+knitr::opts_chunk$set(dpi = 1200)
 ```

### 2022-07-18 analysis/supplement.Rmd
refactor: use short journal names
feat: use 1200 dpi as required by labm for word output

@@
     toc: false
+  bookdown::word_document2:
+    keep_md: true
 ---
 
-```{r knitr_setup, include=FALSE}
+```{r knitr_setup, include = FALSE}
 ```
 
-```{r supplement_setup, message = FALSE}
+```{r knitr_setup_word, eval = !knitr::is_html_output(), include = FALSE}
+knitr::opts_chunk$set(dpi = 1200)
+```
+
+```{r supplement_setup, echo = FALSE, message = FALSE}
 library("ameld")
@@
 m <- list(
+    AMELD = timeROC_RCV,
     MELD = timeROC_MELD,
     "MELD-Na" = timeROC_MELDNa,
-    "MELD-Plus7" = timeROC_MELDPlus7,
-    RCV = timeROC_RCV
+    "MELD-Plus7" = timeROC_MELDPlus7
 )

### 2022-07-18 analysis/supplement.Rmd
chore: suppress code output in supplement

@@
 
-```{r benchmark-table}
+```{r benchmark-table, echo = FALSE}
 agg <- bmrk_results$aggregate(msr("surv.cindex", id = "harrell"))
@@
 
-```{r roc-plot, fig.width = 18, fig.height = 12}
+```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12}
 tar_load(timeROC_MELD)

### 2022-07-18 analysis/article.Rmd
feat: generate png and pdfs for each plot

@@
 
-```{r flowchart, results = "asis", eval = knitr::is_html_output(), echo = FALSE, fig.width = 5, fig.height = 5, fig.cap = "Flowchart Inclusion/Exclusion."}
+```{r flowchart, results = "asis", eval = knitr::is_html_output(), echo = FALSE, fig.width = 5, fig.height = 5, fig.cap = "Flowchart Inclusion/Exclusion.", dev = c("png", "pdf")}
 graph <- "
@@
 ```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = paste('Calibration of AMELD in comparison to MELD, MELD-Na and MELD-Plus7.\n The points mark the mean predicted survival of', ameldcfg$m, 'patients per interval against their observed mean survival.
-The error bars correspond to the 95 % confidence interval of the survival estimate.'), echo = FALSE}
+The error bars correspond to the 95 % confidence interval of the survival estimate.'), dev = c('png', 'pdf'), echo = FALSE}
 par(cex = 0.9)

### 2022-07-18 analysis/supplement.Rmd
feat: generate png and pdfs for each plot

@@
 
-```{r benchmark-plot, fig.cap = "Benchmark results of machine learning algorithms.", echo = FALSE, fig.width = 18, fig.height = 12}
+```{r benchmark-plot, fig.cap = "Benchmark results of machine learning algorithms.", echo = FALSE, fig.width = 18, fig.height = 12, dev = c("png", "pdf")}
 tar_load(bmrk_results)
@@
 
-```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12}
+```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, dev = c("png", "pdf")}
 tar_load(timeROC_MELD)
@@
 
-```{r vimpbs, fig.width = 8, fig.height = 10, fig.cap = "Variable importance by frequency of bootstrap selections.", echo = FALSE, message = FALSE}
+```{r vimpbs, fig.width = 8, fig.height = 10, fig.cap = "Variable importance by frequency of bootstrap selections.", dev = c("png", "pdf"), echo = FALSE, message = FALSE}
 tar_load(bootrcv)
@@
 
-```{r vimprf, fig.width = 8, fig.height = 10, fig.cap = "Variable importance by logrank in random forest.", echo = FALSE, message = FALSE}
+```{r vimprf, fig.width = 8, fig.height = 10, fig.cap = "Variable importance by logrank in random forest.", dev = c("png", "pdf"), echo = FALSE, message = FALSE}
 tar_load(rngr)

### 2022-07-18 analysis/article.Rmd
refactor: labm requires down style title

@@
 # https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
-title: "A New Model to Predict Survival in Patients with End-Stage Liver Disease"
+title: "A new model to predict survival in patients with end-stage liver disease"
 subtitle: >

### 2022-07-18 analysis/article.Rmd
refactor: restruct headings
refactor: reorder keywords
refactor: move data and software availability section
refactor: labm requires upper case litre

@@
   **Keywords**:
-      end-stage liver disease; MELD; risk stratification; machine learning;
-      survival estimation; IL-6.
+      end-stage liver disease; IL-6; machine learning; MELD;
+      risk stratification; survival estimation.
 
@@
 @optn_policies2021 using the following formulas:
-$MELD = 10 * (0.957 \ln(creatinine [mg/dl]) + 0.378 \ln(bilirubin [mg/dl]) + 1.120 * \ln(INR) + 0.643)$,
-Creatinine, bilirubin and INR values lower than 1.0 mg/dl were set to 1.0 mg/dl.
-The maximum allowed creatinine was 4.0 mg/dl. If patients received dialysis,
-creatinine was set to 4.0 mg/dl.
+$MELD = 10 * (0.957 \ln(creatinine [mg/dL]) + 0.378 \ln(bilirubin [mg/dL]) + 1.120 * \ln(INR) + 0.643)$,
+Creatinine, bilirubin and INR values lower than 1.0 mg/dL were set to 1.0 mg/dL.
+The maximum allowed creatinine was 4.0 mg/dL. If patients received dialysis,
+creatinine was set to 4.0 mg/dL.
 For MELD-Na MELD was calculated as above and recalculated if greater than 11 using:
-$MELD\text{-}Na = MELD_i + 1.32 * (137 - Na [mmol/l]) - (0.033 * MELD_i * (137 - Na [mmol/l]))$
+$MELD\text{-}Na = MELD_i + 1.32 * (137 - Na [mmol/L]) - (0.033 * MELD_i * (137 - Na [mmol/L]))$
 
-Serum sodium values lower than 125 mmol/l or higher than 137 mmol/l were set
-to 125 mmol/l and 137 mmol/l, respectively.
+Serum sodium values lower than 125 mmol/L or higher than 137 mmol/L were set
+to 125 mmol/L and 137 mmol/L, respectively.
 
@@
 L = 8.53499496 &+ \\
-    2.59679650 &* \log_{10}(1 + creatinine [mg/dl]) + \\
-    2.06503238 &* \log_{10}(1 + bilirubin [mg/dl]) + \\
+    2.59679650 &* \log_{10}(1 + creatinine [mg/dL]) + \\
+    2.06503238 &* \log_{10}(1 + bilirubin [mg/dL]) + \\
     2.99724802 &* \log_{10}(1 + INR) - \\
-    6.47834101 &* \log_{10}(1 + sodium [mmol/l]) - \\
-    6.34990436 &* \log_{10}(1 + albumin [g/l]) + \\
+    6.47834101 &* \log_{10}(1 + sodium [mmol/L]) - \\
+    6.34990436 &* \log_{10}(1 + albumin [g/L]) + \\
     1.92811726 &* \log_{10}(1 + wbc [th/cumm]) + \\
@@
 
-## Conclusion
+# Conclusion
 
@@
 
-# Data and software availability
-
-The complete code and reproducible analysis could be found at
-https://ampel-leipzig.github.io/ampel-leipzig-meld/.
-Additionally all data are available in the `ameld` R package [@R-ameld] at
-https://github.com/ampel-leipzig/ameld
-[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.6666018.svg)](https://doi.org/10.5281/zenodo.6666018)
-for further investigations.
-
-# Acknowledgement
+## Acknowledgement
 
@@
 
-# Research funding
+## Research funding
 
@@
 
-# Author contributions
+## Author contributions
 
@@
 
-# Competing interests
+## Competing interests
 
@@
 
-# Informed consent
+## Informed consent
 
@@
 
-# Ethical approval
+## Ethical approval
 
@@
 
+## Data and software availability
+
+The complete code and reproducible analysis could be found at
+https://ampel-leipzig.github.io/ampel-leipzig-meld/.
+Additionally all data are available in the `ameld` R package [@R-ameld] at
+https://github.com/ampel-leipzig/ameld
+[![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.6666018.svg)](https://doi.org/10.5281/zenodo.6666018)
+for further investigations.
+
 # References

### 2022-07-19 analysis/article.Rmd
fix: typo
refactor: replace DiagrammeR CONSORT diagram by consort

@@
 
-```{r flowchart, results = "asis", eval = knitr::is_html_output(), echo = FALSE, fig.width = 5, fig.height = 5, fig.cap = "Flowchart Inclusion/Exclusion.", dev = c("png", "pdf")}
-graph <- "
-digraph flowchart {
-    graph [splines = false];
+```{r flowchart, results = "asis", echo = FALSE, fig.width = 4.5, fig.height = 2, fig.cap = "Flowchart Inclusion/Exclusion."}
+library("consort")
+library("grid")
 
-    // invisible nodes
-    node [shape = point, width = 0, height = 0];
-    iAR [style = 'invis'];
+options(txt_gp = gpar(cex = 0.8))
 
-    // visible nodes
-    node [shape = box];
-    A [label = '@@1'];
-    E [label = '@@3'];
-    R [label = '@@2'];
-
-    // edge statements
-    A -> iAR [arrowhead = none];
-    {rank = same; iAR -> E [minlen = 5]};
-    iAR -> R;
-}
-
-[1]: paste0('Total Records\\n(n = ', n_total, ')')
-[2]: paste0('Analysed Records\\n(n = ', n_analysed, ')')
-[3]: str_excluded
-"
-str_excluded <- paste0(
-    'Excluded (n = ', sum(n_excluded), ')\\\\l',
+total_records <- paste0("Total records (n = ", n_total, ")")
+analysed_records <- paste0("Analysed records (n = ", n_analysed, ")")
+excluded_records <- paste0(
+    "Excluded (n = ", sum(n_excluded), ")\n",
     paste0(
-        '- ',
-        paste(
-            c(
-                'Younger than 18 years:',
-                'Liver Transplantation before visit:',
-                'Follow-up missing/invalid:'
+        "- ",
+        c("Younger than 18 years",
+          "Liver transplantation before visit",
+          "Follow-up missing/invalid"
         ),
-            n_excluded
-        ),
-        collapse = '\\\\l'
-    ),
-    '\\\\l'
+        " (n = ", n_excluded, ")",
+        collapse = "\n"
+    )
 )
 
-DiagrammeR::grViz(graph)
+totalbox <- textbox(
+    total_records,
+    x = unit(0.25, "npc"), y = unit(0.9, "npc"),
+    box_fn = rectGrob, name = "vertbox"
+)
+graph <- gList(gList(), totalbox)
+class(graph) <- union("consort", class(graph))
+graph <- structure(graph, split_layout = NULL)
+graph <- add_side_box(graph, txt = excluded_records, dist = 0.1)
+graph <- add_box(graph, txt = analysed_records, dist = 0.1)
+
+#graph <-
+#    add_box(txt = total_records) |>
+#    add_side_box(txt = excluded_records, dist = 0.05) |>
+#    add_box(txt = analysed_records, dist = 0.05) |>
+plot(graph)
 ```
@@
 
-In this study we analysed a cohort of `r n_analysed` patients evaluated for
+In this study we analyzed a cohort of `r n_analysed` patients evaluated for
 liver transplantation (Tab. \@ref(tab:tbl1)).
@@
 The data was acquired by TK.
-The data was analysed and processed by SG.
+The data was analyzed and processed by SG.
 Visualisation were produced by SG.

### 2022-07-19 analysis/article.Rmd
refactor: create png and pdf just for non-html output

@@
 ```{r knitr_setup_word, eval = !knitr::is_html_output(), include = FALSE}
-knitr::opts_chunk$set(dpi = 1200)
+knitr::opts_chunk$set(dpi = 1200, dev = c("png", "pdf"))
 ```
@@
 ```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = paste('Calibration of AMELD in comparison to MELD, MELD-Na and MELD-Plus7.\n The points mark the mean predicted survival of', ameldcfg$m, 'patients per interval against their observed mean survival.
-The error bars correspond to the 95 % confidence interval of the survival estimate.'), dev = c('png', 'pdf'), echo = FALSE}
+The error bars correspond to the 95 % confidence interval of the survival estimate.'), echo = FALSE}
 par(cex = 0.9)

### 2022-07-19 analysis/article.Rmd
fix: labm requires vs.

@@
     caption = paste0(
-        "Observed vs MELD-expected 90-day mortality. ",
+        "Observed vs. MELD-expected 90-day mortality. ",
         "MELD mortality values are taken from @wiesner2003. ",

### 2022-07-19 analysis/article.Rmd
feat: extend figure captions

@@
 
-```{r flowchart, results = "asis", echo = FALSE, fig.width = 4.5, fig.height = 2, fig.cap = "Flowchart Inclusion/Exclusion."}
+```{r flowchart, results = "asis", echo = FALSE, fig.width = 4.5, fig.height = 2, fig.cap = "Flowchart. Inclusion and exclusion of records."}
 library("consort")
@@
 
-```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE}
+The final model was compared against the MELD, MELD-Na and MELD-Plus7 score
+using the area under the time-dependent receiver operating characteristic curve
+(AUROC) based on the nonparametric inverse probability of censoring weighting
+estimator (IPCW) [@blanche2013, @R-timeROC].
+Testing was applied as described in @blanche2013.
+
+```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE, fig.cap = "Receiver operating characteristic (ROC) curve.
+Area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na and MELD-Plus7 as described in [@blanche2013]."}
 col <- viridisLite::viridis(5)

### 2022-07-19 analysis/supplement.Rmd
refactor: create png and pdf just for non-html output
feat: extend supplement with survival plot

--- 2022-07-18 analysis/supplement.Rmd
@@
 ---
-title: "Supplement"
+title: "A new model to predict survival in patients with end-stage liver disease"
+subtitle: >
+    Supplement
+author:
+    - Sebastian Gibb^[mail@sebastiangibb.de]
+    - Thomas Berg
+    - Daniel Seehofer
+    - Berend Isermann
+    - Thorsten Kaiser
+bibliography:
+  - "`r rprojroot::find_rstudio_root_file('analysis', 'bibliography', 'bibliography.bib')`"
 output:
@@
     keep_md: true
+    pandoc_args: # pandoc_args doesn't support r evaluation
+      - --csl=pandoc/csl/vancouver.csl
 ---
@@
 ```{r knitr_setup_word, eval = !knitr::is_html_output(), include = FALSE}
-knitr::opts_chunk$set(dpi = 1200)
+knitr::opts_chunk$set(dpi = 1200, echo = FALSE)
 ```
@@
 library("ggplot2")
+```
+
+# Survival plot
+
+```{r survival.plot, fig.cap = "Survival plot. Kaplan-Meier survival estimate for the first 90 days of the analysed population.", fig.width = 8, fig.height = 5}
+tar_load(zlog_data)
+srv <- Surv(zlog_data$DaysAtRisk, zlog_data$Deceased)
+col <- palette.colors(nlevels(zlog_data$MeldCategory))
+
+srvfit <- survfit(srv ~ 1)
+times <- c(2, 7, 14, 30, 90)
+
+## calculate risk tables
+sm <- summary(srvfit, times = times)
+nrisk <- as.matrix(sm$n.risk)
+ncumevents <- as.matrix(cumsum(sm$n.event))
+rownames(nrisk) <- rownames(ncumevents) <- times
+
+## keep old graphic parameters and restore them afterwards
+old.par <- par(no.readonly = TRUE)
+
+layout(matrix(1:3, nrow = 3), height = c(5, 1, 1))
+plot_surv(
+    srvfit,
+    main = "Kaplan-Meier survival estimate to day 90",
+    times = times,
+    xmax = 90,
+    col = col[2],
+    cex = 1.2
+)
+par(mar = c(1.1, 5.1, 1.1, 2.1))
+plot_table(
+    nrisk, at = times, main = "Number at risk",
+    xaxis = FALSE, cex.text = 1, ylabels = FALSE
+)
+par(mar = c(1.1, 5.1, 1.1, 2.1))
+plot_table(
+    ncumevents, at = times, main = "Cumulative number of events",
+    xaxis = FALSE, cex.text = 1, ylabels = FALSE
+)
+par(old.par)
 ```
@@
 setorder(agg, -harrell)
-knitr::kable(agg, col.names = c("Learner ID", "Harrell C"))
+knitr::kable(
+    agg,
+    col.names = c("Learner ID", "Harrell C"),
+    caption = paste0("Benchmark results. Ranking of the ", nrow(agg), " tested machine learning algorithms.")
+)
 ```
@@
 
-```{r roc-plot, fig.width = 18, fig.height = 12}
+```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in AUROC.
+Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na and MELD-Plus7 as described in [@blanche2013]."}
 tar_load(timeROC_MELD)
@@
 )
-plot_surv_roc_trend(m)
+plot_surv_roc_trend(m, xlab = "time [days]", col = viridisLite::viridis(5)[-2])
 ```
@@
 ```
+
+# References

### 2022-07-22 analysis/article.Rmd
refactor: apply proofreading suggestions

--- 2022-07-19 analysis/article.Rmd
@@
   **Objectives**:
-      The shortage in grafts for liver transplantation requires
+      The shortage of grafts for liver transplantation requires
       risk stratification and adequate allocation rules.
@@
       score for 90-day mortality prediction with the help of different
-      machine learning (ML) algorithms.
+      machine-learning algorithms.
 
   **Methods**:
-      We retrospectively analyzed clinical and laboratory data of 654
+      We retrospectively analyzed the clinical and laboratory data of 654
       patients who were recruited during the evaluation process for
-      liver transplantation at the University Hospital Leipzig.
-      After comparing 13 different ML algorithms in a nested cross-validation
-      setting and selecting the best performing one we build a new model
-      to predict 90-day mortality in patients with end-stage liver disease.
+      liver transplantation at University Hospital Leipzig.
+      After comparing 13 different machine-learning algorithms in a nested
+      cross-validation setting and selecting the best performing one we build
+      a new model to predict 90-day mortality in patients with end-stage liver
+      disease.
 
   **Results**:
-      Penalized regression algorithms yield the highest prediction performance
-      in our ML algorithm benchmark.
-      In favour of a simpler model we chose the least absolute shrinkage and
+      Penalized regression algorithms yielded the highest prediction
+      performance in our machine-learning algorithm benchmark.
+      In favor of a simpler model we chose the least absolute shrinkage and
       selection operator (lasso) regression.
-      Beside the classical MELD variables INR and bilirubin the
-      lasso regression selected cystatin C over creatinine, and
-      additionally IL-6, total protein and cholinesterase.
-      The new model offers an improved discrimination and calibration than MELD
+      Beside the classical MELD variables international normalized ratio (INR)
+      and bilirubin, the lasso regression selected cystatin C over creatinine,
+      as well IL-6, total protein, and cholinesterase.
+      The new model offers improved discrimination and calibration over MELD
       and MELD with sodium (MELD-Na) or the MELD-Plus7 risk score.
@@
   **Conclusion**:
-      We provide a new ML based model of end-stage liver disease that
-      outperforms the classical MELD score for 90-day survival prediction.
+      We provide a new machine-learning-based model of end-stage liver disease
+      that outperforms the classical MELD score for 90-day survival prediction.
 
@@
       end-stage liver disease; IL-6; machine learning; MELD;
-      risk stratification; survival estimation.
+      risk stratification; survival estimation
 
@@
 
-Liver cirrhosis is the terminal result of fibrotic remodeling
+Liver cirrhosis is the terminal result of the fibrotic remodeling
 of liver tissue due to chronic damage.
-This end-stage of liver disease is usually irreversible and
+This end-stage of liver disease is usually irreversible, and
 the only available therapy is liver transplantation.
-However, the shortage in grafts for transplantation from deceased donors
+However, the shortage of grafts for transplantation from deceased donors
 requires risk stratification and precise and fair allocation rules.
@@
 [@malinchoc2000; @wiesner2003; @optn_policies2021].
-The MELD score estimates the patients' 3-month mortality risk based on
-laboratory results, namely bilirubin, creatinine and international normalized
-ratio (INR).
+The MELD score estimates patients' 3-month mortality risk based on
+laboratory results, namely bilirubin, creatinine, and
+the international normalized ratio (INR).
 The MELD score has been extended by the sodium level
-(MELD-Na score) as this was found to be an important additional risk factor in
-liver cirrhosis [@kim2008; @optn_policies2021].
-The MELD was initially developed for predicting the survival of patients
+(MELD-Na score) because this was found to be an important additional
+risk factor in liver cirrhosis [@kim2008; @optn_policies2021].
+The MELD was initially developed to predict the survival of patients
 undergoing transjugular intrahepatic portosystemic shunts.
-Afterwards it was revalidated for predicting mortality risk in patients awaiting
+It was subsequently revalidated to predict mortality risk in patients awaiting
 a liver transplantation.
-Although the MELD score should be an objective allocation score especially the
-creatinine and INR are highly dependent on the utilized laboratory methods
+Although the MELD score should be an objective allocation score,
+creatinine and INR are highly dependent on the laboratory methods used
 [@trotter2004; @cholongitas2007].
-Additionally women are disadvantage [@sealock2022].
-Patients with identical disease state could have very different MELD scores and
-thus get different priority on the liver transplantation waiting list.
-Furthermore often, e.g. for acute-on-chronic liver failure, the MELD score
-underestimates the mortality risk [@hernaez2020].
+In addition women are disadvantaged [@sealock2022].
+Patients with identical disease states can have very different MELD scores and
+thus receive different priority levels on the liver transplantation waiting
+list.
+Furthermore, for acute-on-chronic liver failure, for example, the MELD score
+often underestimates the mortality risk [@hernaez2020].
 
-There have been some attempts to use the data extracted from more than 300.000
+There have been some attempts to use the data extracted from more than 300,000
 electronic medical records from two hospitals in the United States to improve
-the MELD score [@kartoun2017]. The derived MELD-Plus7 and MELD-Plus9 risk score
-add albumin, white blood cell count, total cholesterol, age and length of
-stay to the MELD-Na variables. Despite its published prediction improvement the
-MELD-Plus scores are not used for transplant allocation yet.
+the MELD score [@kartoun2017]. The derived MELD-Plus7 and MELD-Plus9
+risk scores add albumin, white blood cell count, total cholesterol, age, and
+length of stay to the MELD-Na variables.
+Despite their published prediction improvement, the MELD-Plus scores are not
+yet used for transplant allocation.
 
-As depicted by MELD-Plus risk scores better predictive scores often need more
+As shown by the MELD-Plus risk scores, better predictive scores often need more
 variables and are more complicated. To reduce the risk of overlooking or
@@
 clinical decision support system based on laboratory diagnostics that should
-support clinical practitioners in interpreting the laboratory results and taking
+support clinical practitioners in interpreting laboratory results and taking
 the necessary medical interventions [@eckelt2020].
 
-This study aims to find clinical and laboratory values that improve the risk
-stratification for liver transplantation over classical MELD, MELD-Na and
-MELD-Plus scores and could be implemented as part of the AMPEL clinical decision
-support system.
+This study aims to find clinical and laboratory values that improve on the risk
+stratification for liver transplantation of the classical MELD, MELD-Na, and
+MELD-Plus scores and that might be implemented as part of the
+AMPEL clinical decision support system.
 
@@
 patients, who were recruited during the evaluation process for
-liver transplantation at the University Hospital of Leipzig from
+liver transplantation at University Hospital Leipzig from
 `r paste(format(attr(raw_data, "period"), "%B %Y"), collapse = " to ")`.
-For each patient we recorded `r ncol(raw_data)` variables.
-Among them are age, sex, etiology of liver disease, complications as listed in
-Table \@ref(tab:tbl1) and `r length(cn_labs)` laboratory measurements.
+For each patient, we recorded `r ncol(raw_data)` variables, including
+age, sex, etiology of liver disease, complications as listed in
+Table \@ref(tab:tbl1), and `r length(cn_labs)` laboratory measurements.
 
-```{r flowchart, results = "asis", echo = FALSE, fig.width = 4.5, fig.height = 2, fig.cap = "Flowchart Inclusion/Exclusion."}
+```{r flowchart, results = "asis", echo = FALSE, fig.width = 4.5, fig.height = 2, fig.cap = "Flowchart showing inclusion and exclusion of records."}
 library("consort")
@@
 total_records <- paste0("Total records (n = ", n_total, ")")
-analysed_records <- paste0("Analysed records (n = ", n_analysed, ")")
+analysed_records <- paste0("Analyzed records (n = ", n_analysed, ")")
 excluded_records <- paste0(
@@
 (Figure \@ref(fig:flowchart)).
-`r if (n_excluded["Children"] == 1) "One was" else paste(n_excluded["Children"], "were")` younger than 18 years.
-`r n_excluded["LiverTransplantation"]` had a liver transplantation before and
-in `r n_excluded["LostFollowUp"]` cases the follow up data were missing or
+`r if (n_excluded["Children"] == 1) "One was" else paste(n_excluded["Children"], "were")` younger than 18 years,
+`r n_excluded["LiverTransplantation"]` had had a liver transplant before and
+in `r n_excluded["LostFollowUp"]` cases the follow-up data were missing or
 invalid.
@@
 Creatinine, bilirubin and INR values lower than 1.0 mg/dL were set to 1.0 mg/dL.
-The maximum allowed creatinine was 4.0 mg/dL. If patients received dialysis,
+The maximum allowed creatinine was 4.0 mg/dL. If the patients received dialysis,
 creatinine was set to 4.0 mg/dL.
-For MELD-Na MELD was calculated as above and recalculated if greater than 11 using:
+For MELD-Na, MELD was calculated as above and recalculated if greater than
+11 using:
 $MELD\text{-}Na = MELD_i + 1.32 * (137 - Na [mmol/L]) - (0.033 * MELD_i * (137 - Na [mmol/L]))$
@@
 The MELD-Plus9 score was ignored because the length of stay information was not
-available for our cohort and the model itself wasn't superior to MELD-Plus7 in
-its original publication [@kartoun2017].
+available for our cohort, and the model itself was not superior to MELD-Plus7
+in its original publication [@kartoun2017].
 
@@
 
-All data processing, statistical and machine learning analyses were performed
+All data processing, statistical, and machine-learning analyses were performed
 using R version
@@
 [@R-base].
-Prior to the analyses all laboratory values were *zlog* transformed to
+Prior to the analyses, all laboratory values were *zlog* transformed to
 approximate a normal distribution as described in
@@
 
-Where $LL$ and $UL$ are the lower and upper reference limit of
-the corresponding laboratory diagnostic, respectively.
+where $LL$ and $UL$ are the lower and upper age-, and sex-related
+reference limits of the corresponding laboratory diagnostic, respectively.
 Missing laboratory measurements were replaced with a *zlog* value of zero.
-In similar manner if information about complications or
-comorbidities were missing we treat them as not present.
+In a similar manner, where information about complications or
+comorbidities was missing, we treated them as not present.
 
-We compared `r n_lrns` different statistical and machine learning
+We compared `r n_lrns` different statistical and machine-learning
 algorithms using the *mlr3* framework
 [@mlr3; @R-mlr3; @mlr3proba2021; @R-mlr3proba].
-The used algorithms are designed for analysis of survival data:
+The algorithms used were designed for analysis of survival data:
 Cox proportional hazards regression model
-[@cox1972; @survival-book; @R-survival],
+[@cox1972; @survival-book; @R-survival];
 penalized regressions,
-namely two different implementations of the
+namely, two different implementations of the
 lasso (least absolute shrinkage and selection operator) regression
@@
 ridge regression [@glmnet2011; @R-glmnet; @penalized2010; @R-penalized], and
-the elastic net regression [@glmnet2011; @R-glmnet],
+the elastic net regression [@glmnet2011; @R-glmnet];
 two different random forest implementations
-[@ranger2017; @R-ranger; @randomForestSRC2008; @R-randomForestSRC],
-extreme gradient boosting [@chen2016; @R-xgboost],
-two different support vector machines [@vanbelle2011; @R-survivalsvm]
-and two neural networks [@katzman2018; @kvamme2019; @R-survivalmodels].
+[@ranger2017; @R-ranger; @randomForestSRC2008; @R-randomForestSRC];
+extreme gradient boosting [@chen2016; @R-xgboost];
+two different support vector machines [@vanbelle2011; @R-survivalsvm];
+;nd two neural networks [@katzman2018; @kvamme2019; @R-survivalmodels].
 
 We used nested resampling to avoid feature- and
-model-selection-bias to get reliable metrics to choose the best model.
-Therefore, we applied a `r crossval$inner$param_set$values$folds`-fold inner
-and a `r crossval$outer$param_set$values$folds`-fold outer cross-validation
+model-selection-bias to obtain reliable metrics to choose the best model.
+We thus applied a `r crossval$inner$param_set$values$folds`-fold inner
+and `r crossval$outer$param_set$values$folds`-fold outer cross-validation
 that was repeated `r crossval$outer$param_set$values$repeats` times.
@@
 rank correlation between predicted and observed events [@harrell1982].
-Subsequently we selected the model with the highest median concordance index.
-If multiple models were comparable we chose the simpler (more penalized) one.
+Subsequently, we selected the model with the highest median concordance index.
+If multiple models were comparable, we chose the simpler (more penalized) one.
 
-After selecting the final model we trained and validated it as described in
+After selecting the final model, we trained and validated it as described in
 @harrell1996.
-Briefly we developed our model using a `r ameldcfg$nfold` cross-validation
-that was repeated `r ameldcfg$nrepcv` times and finally predicted survival
-at `r ameldcfg$times` days.
-The predictions were sorted and grouped into equal sized intervals so that there
-are `r ameldcfg$m` patients in each interval.
-For each group the mean survival probability and the difference to the
+Briefly, we developed our model using a `r ameldcfg$nfold`-fold
+cross-validation that was repeated `r ameldcfg$nrepcv` times and finally
+predicted survival at `r ameldcfg$times` days.
+The predictions were sorted and grouped into equal-sized intervals so that there
+were `r ameldcfg$m` patients in each interval.
+For each group, the mean survival probability and the difference to the
 Kaplan-Meier estimate was calculated.
 The same procedure was repeated for `r ameldcfg$nboot` bootstrap samples.
-Subsequently the differences between predicted survival and Kaplan-Meier
+Subsequently, the differences between predicted survival and Kaplan-Meier
 estimates of the bootstrap samples were averaged and added to the original
-model difference to obtain a bias-corrected estimated between predicted and
+model difference to obtain a bias-corrected estimate between predicted and
 observed survival.
 
-The final model was compared against the MELD, MELD-Na and MELD-Plus7 score
+The final model was compared against the MELD, MELD-Na, and MELD-Plus7 scores
 using the area under the time-dependent receiver operating characteristic curve
@@
 Testing was applied as described in @blanche2013.
-A $p < 0.05$ was considered as statistical significant difference.
+A $p < 0.05$ was considered a statistically significant difference.
 
@@
 
-In this study we analyzed a cohort of `r n_analysed` patients evaluated for
+In this study, we analyzed a cohort of `r n_analysed` patients evaluated for
 liver transplantation (Tab. \@ref(tab:tbl1)).
+Of these,
 `r sprintf("%i (%s)", sum(raw_data$Sex == "male"), gtsummary::style_percent(mean(raw_data$Sex == "male"), symbol = TRUE))`
-of them were men.
+were men.
 The median (IQR) follow-up times was `r .tbl1("DaysAtRisk")` days.
-Within the follow-up time `r .tbl1("LTx")` patients received a liver transplant
+Within the follow-up time, `r .tbl1("LTx")` patients received a liver transplant
 and were censored for the analysis beyond this time point.
@@
 (`r .tbl1("Cirrhosis")`).
-The remaining patients had an acute liver failure
+The remaining patients had acute liver failure
 (`r .tbl1("ALF")`) or
-a chronic liver insufficiency
+chronic liver insufficiency
 (`r .tbl1("CLI")`).
-At evaluation time the median (IQR) MELD and MELD-Na score were
+At evaluation time, the median (IQR) MELD and MELD-Na scores were
 `r .tbl1("ScoreMeldUnos")` and
 `r .tbl1("ScoreMeldNaUnos")`, respectively.
-Within 90 days `r .tbl1("M90")` patients died.
-In nearly two third of all patients alcohol was the main cause
-(`r .tbl1("Ethyltoxic")`) of their end-stage liver disease.
-The remaining patients had viral infections
-(hepatitis B and C viruse `r .tbl1("HBV")` and `r .tbl1("HCV")`, respectively),
-autoimmune processes (autoimmune hepatitis `r .tbl1("AIH")`,
-primary sclerosing cirrhosis `r .tbl1("PSC")`),
-primary biliary cirrhosis `r .tbl1("PBC")` or unknown causes
-(non-alcoholic steatohepatitis `r .tbl1("NASH")` and
-cryptogenic hepatitis `r .tbl1("Cryptogenic")`).
+Within 90 days, `r .tbl1("M90")` patients had died.
+In nearly two thirds of all patients
+(`r .tbl1("Ethyltoxic")`), alcohol was the main cause
+of their end-stage liver disease.
+The remaining patients had viral infections -
+hepatitis B (`r .tbl1("HBV")`) or C (`r .tbl1("HCV")`) -
+autoimmune autoimmune (`r .tbl1("AIH")`),
+primary sclerosing cirrhosis (`r .tbl1("PSC")`),
+primary biliary cirrhosis (`r .tbl1("PBC")`), unknown causes
+(non-alcoholic steatohepatitis (`r .tbl1("NASH")`), or
+cryptogenic hepatitis (`r .tbl1("Cryptogenic")`)).
 The most common complications among all patients were gastrointestinal
@@
 spontaneous bacterial peritonitis (SBP) in `r .tbl1("SBP")` patients.
-Just `r .tbl1("Dialysis")` patients required a renal dialysis.
+Just `r .tbl1("Dialysis")` patients required renal dialysis.
 
@@
 
-According to @wiesner2003 we divided our cohort into
+According to @wiesner2003, we divided our cohort into
 `r nlevels(labelled_meld_data$MeldCategory)` MELD-score risk
@@
 respectively.
-As shown in Table \@ref(tab:tbl-observed-vs-expected-mortality)
+As shown in Table \@ref(tab:tbl-observed-vs-expected-mortality),
 our observed 90-day mortality was
 `r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
-times higher than the predicted one. Just in the group with the lowest MELD
-scores the mortality was much lower.
+times higher than the predicted one.
+Only in the group with the lowest MELD scores was the mortality much lower.
 
@@
 The penalized regression and random forest models
-result in a similar high median performance in the benchmark process.
-In favour of simpler models we chose a lasso based approach for the final model
+resulted in a similar high median performance in the benchmark process.
+In favour of simpler models, we chose a lasso-based approach for the final model
 development.
-The classical cox model, the data and computational intense algorithms
-extreme gradient boosting, support vector machines and neural networks yield a
-worse performance.
+The classical Cox model, the data and computational intense algorithms
+extreme gradient boosting, the support vector machines and the neural networks
+yielded a worse performance.
 
@@
 
-The resulting model AMPEL model of end-stage liver disease
+The resulting AMPEL model of end-stage liver disease
 (AMELD) uses `r length(cfs)` predictors.
-Beside the classical MELD variables INR and (direct) bilirubin the lasso
-regression selected cystatin C over creatinine, and
-additionally IL-6, total protein and cholinesterase.
+Besides the classical MELD variables INR and (direct) bilirubin,
+the lasso regression selected cystatin C over creatinine, as well as
+IL-6, total protein, and cholinesterase.
 
@@
 
-The 90-day survival estimate of a given patient could be calculated as shown in
+The 90-day survival estimate of a given patient can be calculated as shown in
 \@ref(eq:survcalc).
@@
 
-Where $S_{0}$ is the baseline survival estimate from
+where $S_{0}$ is the baseline survival estimate from
 \@ref(tab:basehaz) and $LP$ is the linear predictor of the coefficients
@@
 
-```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE}
+```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE, fig.cap = "Receiver operating characteristic (ROC) curve.
+Area under the time-dependent ROC curve (AUC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, and MELD-Plus7, as described in [@blanche2013]."}
 col <- viridisLite::viridis(5)
@@
 
-AMELD offers the best AUROC (`r round(roc["AMELD"], 3)`) across all competitors.
-However, after correction for multiple testing the difference against
-the second best model `r names(roc)[2]`
-(AUROC: `r round(roc[2], 3)`) is not statistical significant
+AMELD offers the best AUROC (`r round(roc["AMELD"], 3)`) of all competitors.
+However, after correction for multiple testing, the difference against
+the second-best model, `r names(roc)[2]`
+(AUROC: `r round(roc[2], 3)`), is not statistical significant
 (`r paste0("$p = ", round(tp[2, "t=90"], 2), "$")`)
@@
 
-```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = paste('Calibration of AMELD in comparison to MELD, MELD-Na and MELD-Plus7.\n The points mark the mean predicted survival of', ameldcfg$m, 'patients per interval against their observed mean survival.
-The error bars correspond to the 95 % confidence interval of the survival estimate.'), echo = FALSE}
+```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = paste('Calibration of AMELD in comparison to MELD, MELD-Na, and MELD-Plus7.\n The points mark the mean predicted survival of', ameldcfg$m, 'patients per interval against their observed mean survival.
+The error bars correspond to the 95% confidence interval of the survival estimate.'), echo = FALSE}
 par(cex = 0.9)
@@
 competing MELD scores.
-While the prediction for low-risk patients is similar to other scores
-AMELD predicts the medium-risk patients more accurate.
-Because of the low number of very ill patients in our cohort the prediction
+While the prediction for low-risk patients is similar to other scores,
+AMELD predicts the medium-risk patients more accurately.
+Because of the low number of very ill patients in our cohort, the prediction
 performance in high-risk patients is poor but comparable to the other models.
-This would require a recalibration in a larger validation cohort.
+This would require recalibration in a larger validation cohort.
 
@@
 
-In this retrospective analysis of `r n_analysed` consecutive patients,
+In this retrospective analysis of `r n_analysed` consecutive patients
 who were recruited during the evaluation process for liver transplantation at
-the University Hospital of Leipzig we developed a new model to predict the
+University Hospital Leipzig, we developed a new model to predict
 90-day survival probability. Our new model, AMELD, outperforms the MELD-Na
@@
 ```
+
 Although the Organ Procurement and Transplantation Network recommends the use
-of MELD-Na over MELD the sodium does not add any information to the score.
+of MELD-Na over MELD, the sodium does not add any information to the score.
 Neither in the bootstrap selection nor in the variable importance of a
-random forest model the sodium gains any important weight
+random forest model does the sodium gain any important weight
 (`r bsranktxt` and `r rfranktxt`, see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
-In addition sodium levels can be affected by diuretic therapy and already
-@kartoun2017 have shown that the discrimination performance of MELD-Na is
-insignificantly different from the classical MELD.
-Consequentially it is not a component of our model.
+In addition, sodium levels can be affected by diuretic therapy, and
+the discrimination performance of MELD-Na has already been shown by
+@kartoun2017 to be insignificantly different from the classical MELD.
+Consequently it is not a component of our model.
 
-It has shown that women receive in median lower scores in MELD-Na and therefore
-are less likely getting an urgent listing for liver transplantation
+It has been shown that women receive median lower scores in MELD-Na and
+therefore are less likely to receive an urgent listing for liver transplantation
 [@sealock2022].
-To address sexual differences in laboratory measurements we applied the *zlog*
+To address sexual differences, in laboratory measurements we applied the *zlog*
 transformation [@hoffmann2017] in our study.
-Additionally this transformation yield comparable laboratory measurements
-across different clinical laboratories and methods.
-It also reduce the influence of age- and sex-related differences,
+This transformation also yields comparable laboratory measurements
+across different clinical laboratories and methods, as well as reducing
+the influence of age- and sex-related differences,
 which could improve the objectivity of the allocation process.
 The already nearly normal scaled *zlog*-transformed laboratory measurements
-allow us to avoid the sample depending scaling prior to model development.
+allow us to avoid the sample-dependent scaling prior to model development.
 Both should increase the external applicability of our approach.
@@
 Our model, AMELD, uses `r length(cfs)` predictors.
-Half of them, INR, (direct) bilirubin, and cystatin C are identical
-or similar to the labortory measurements utilized in the classical MELD.
+Three of these - INR, (direct) bilirubin, and cystatin C - are identical
+or similar to the laboratory measurements utilized in the classical MELD.
 The direct bilirubin selected by our approach and the total bilirubin in MELD
-are highly correlated (spearman correlation:
+are highly correlated (Spearman correlation:
 `r round(cor(amelddata$x, method = "spearman")["BILI_S", "BILID_S"], 2)`) and
@@
 
-In our feature selection process cystatin C was chosen instead of creatinine.
-This is inline with recent studies that show a better prediction of mortality
-for cystatin C in unselected and critical ill patients
-[@helmerssonkarlqvist2016; @helmerssonkarlqvist2021]
+In our feature selection process, cystatin C was chosen instead of creatinine.
+This is in line with recent studies that show a better prediction of mortality
+for cystatin C in unselected and critically ill patients
+[@helmerssonkarlqvist2016; @helmerssonkarlqvist2021].
 
-Furthermore AMELD extends MELD by additional synthesis parameters
-total protein and cholinesterase and
+Furthermore AMELD, extends MELD by the additional synthesis parameters
+total protein and cholinesterase, and
 most importantly the inflammatory marker IL-6.
 
-The latter is an example where results from machine learning can aid in
-finding clinical meaningful patterns and guide further research.
-The systemic inflammatory response syndrome is an
-independent risk factor for poor outcome in patients with cirrhosis
+The latter is an example where results from machine-learning can aid in
+finding clinically meaningful patterns and guide further research.
+Systemic inflammatory response syndrome is an
+independent risk factor for poor outcomes in patients with cirrhosis
 [@thabut2007].
-However neither MELD nor MELD-Na taking inflammation into account.
-In contrast IL-6 is the most important variable in our model. Even if IL-6 is
-excluded the algorithms select CRP as one of the most important variables which
-highlights the important role of inflammation in the prognosis of patients
-with end-stage liver disease.
+However, neither MELD nor MELD-Na took inflammation into account.
+By contrast, IL-6 is the most important variable in our model. Even if IL-6 is
+excluded, the algorithms select CRP as one of the most important variables,
+which highlights the important role of inflammation in the prognosis of
+patients with end-stage liver disease.
 We and other researchers have already shown that adding inflammatory markers,
-namely IL-6 or CRP, to the classical MELD score increase the accuracy of the
+namely IL-6 or CRP, to the classical MELD score increases the accuracy of the
 90-day mortality prediction [@remmler2018; @cervoni2016].
 
-Beside laboratory measurements we investigated
-`r (ncol(raw_data) - length(cn_labs) - 2)`
-clinical data but none of them were chosen in our model selection process.
-@kartoun2017 also used the lasso approach and neither select any clinical
-data. Especially in retrospective analysis the clinical data are often not
-detailed enough to keep valuable information.
+Besides laboratory measurements, we investigated
+`r (ncol(raw_data) - length(cn_labs) - 2)` other items of
+clinical data, but none were chosen in our model-selection process.
+@kartoun2017 also used the lasso approach and did not select any clinical
+data.
+In retrospective analyses in particular, the clinical data are often not
+detailed enough to provide valuable information.
 
-In our benchmark of `r n_lrns` different machine learning algorithms we have
-seen that the more complex algorithm like boosting, support vector machines
-and neural networks perform worse than simpler ones.
-This may due additive effects that favor regression models or due to our
-relative small sample size.
-It has been shown that in machine learning more than 200 events per
+In our benchmark of `r n_lrns` different machine-learning algorithms we have
+seen that the more complex algorithms, such as boosting,
+support vector machines, and neural networks, perform worse than simpler ones.
+This may be due to additive effects that favor regression models or due to our
+relatively small sample size.
+It has been shown that in machine-learning more than 200 events per
 predictor variable are required to achieve good prediction performance
 [@ploeg2014].
-
-<!--
-As we shown above machine learning can aid in finding clinical meaningful
-patterns.
--->
-
-<!--
-
-- higher mortality than predicted, compare @hernaez2020
-- adding sodium doesn't add much/any discrimination information
-    - as already seen in @kartoun2017
-- @kartoun2017 uses > 60 variables mostly clinical stuff, just few laboratory values, none of the clinical stuff remains
-- general low sample size thus simpler models/fewer predictors are preferred.
-- comparison of different predictors of MELD-Plus7 (+MELD-Plus9) vs our features
-    - 90 day mortality *post* discharge
-    - poorer performance in own validation set (lower and more uniform disease severity, also lower MELD)
-- machine learning algorithms worse than statistical learning (lower c-index, higher variance/instability)
-    - additive effects (favour statistical models)
-    - low sample size, for ML more than 200 events per predictor variable necessary [@ploeg2014]
-- sex-differences and sample-dependencies (by scaling) addressed by *zlog*.
-- more variables, harder to cheat
-- national register required
--->
 
@@
 Our study has several limitations.
-Principally it is a single center, retrospective analysis and without an
+Principally, it is a single-center, retrospective analysis lacking an
 external validation cohort.
 The data were collected during the evaluation process for liver transplantation
-at the University Hospital of Leipzig.
+at University Hospital Leipzig.
 There was no controlled follow-up process.
-We excluded all patients who had already a liver transplantation
-which could artificial lower the disease severity of the cohort.
-Furthermore most of our patients suffered from ethyltoxic liver cirrhosis
-(`r .tbl1("Ethyltoxic", pattern = "{p}%")`) which may
+We excluded all patients who had already undergone liver transplantation,
+which could artificially lower the disease severity of the cohort.
+Furthermore, most of our patients suffered from ethyltoxic liver cirrhosis
+(`r .tbl1("Ethyltoxic", pattern = "{p}%")`), which may
 limited the transfer to other etiologies.
 
-With a median MELD score of `r .tbl1("ScoreMeldUnos", pattern = "{median}")`
-our cohort was healthier than in previous studies that reported
+With a median MELD score of `r .tbl1("ScoreMeldUnos", pattern = "{median}")`,
+our cohort was healthier than those in previous studies that reported
 a score of 14-15 [@kim2008; @kartoun2017].
-However the 90-day mortality rate of `r .tbl1("M90")` was comparable to
-the mortality rates seen in earlier studies that ranges from 6 to 30%
+However, the 90-day mortality rate of `r .tbl1("M90")` was comparable to
+the mortality rates seen in earlier studies ranging from 6% to 30%
 [@malinchoc2000; @kim2008; @kartoun2017].
-Although our cohort was healthier with a medium mortality rate the classical
-MELD score underestimates the observed mortality
+Although our cohort was healthier, with a medium mortality rate, the classical
+MELD score underestimates the observed mortality by
 `r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
 times.
-Unfortunately our sample size is low in the high risk patient category
-which yield a poor but better than MELD based prediction performance of our
+Unfortunately, our sample size was low in the high-risk patient category,
+resulting in a poor but better than MELD-based prediction performance by our
 AMELD for very ill patients.
@@
 
-In conclusion we developed a new model to predict the 90-day survival of
+In conclusion, we have developed a new model to predict the 90-day survival of
 patients during the evaluation process for liver transplantation.
-Our model, AMELD, extends the classical MELD predictors, INR,
-(direct) bilirubin, cystatin C (instead of creatinine) by
-synthesis parameters total protein and cholinesterase and
-most importantly the inflammatory marker IL-6.
-Using these `r length(cfs)` laboratory measurments the AMELD
+Our model, AMELD, extends the classical MELD predictors INR,
+(direct) bilirubin, and cystatin C (instead of creatinine) by using the
+synthesis parameters total protein and cholinesterase and,
+most importantly, the inflammatory marker IL-6.
+Using these six laboratory measurements, AMELD
 outperforms the established MELD-Na and the recently published
 MELD-Plus7 risk scores.
-For a wider adoption a prospective multi center study for model
+For wider adoption, a prospective multi-center study for model
 recalibration and validation is needed.
 
-<!--
-Mortality
-[@malinchoc2000] ~ 30 %
-[@kim2008] ~ 6-7 %
-[@kartoun2017] ~ 16.3 % (90 d, post-discharge-mortality)
-
-random time point for blood sampling
-
-It was not possible to ensure that patients died of end-stage liver disease.
-
-The variables we found and the model itself have to be validated and
-recalibrated in a multi center, prospective trial.
-
-Few patients with high MELD scores, poor calibration for high mortality
-
-- comparable mortality?
-- retrospective analysis
-- single center
-- unknown cause of death (death with vs of end-stage liver disease)
-- bias caused by LTx censorship?
--->
-
-## Acknowledgement
+## Acknowledgments
 
@@
 Stefan Kemnitz for his technical support regarding the use of the
-HPC cluster of the University Greifswald and
-Bernd Klaus for helpful discussions about survival analysis and proof reading the manuscript.
+high-performance computing (HPC) cluster at the University of Greifswald and
+Bernd Klaus for helpful discussions about survival analysis and
+proofreading the manuscript.
 
@@
 
-The calculations were conducted using the high-performance computing (HPC)
-cluster of the University Greifswald.
-In 2018 the average carbon efficiency of the German power grid was
+The calculations were conducted using the HPC cluster at
+University of Greifswald.
+In 2018, the average carbon efficiency of the German power grid was
 `r kgco2kwh` kgCO~2~eq/kWh [@icha2021].
@@
 This project is co-financed through public funds according to the budget
-decided by the Saxon State Parliament under the
+decided by the Saxon State Parliament under
 RL eHealthSax 2017/18 grant number 100331796.
@@
 Funding was acquired by TK.
-The study was conceptualised by SG and TK.
-The data was acquired by TK.
-The data was analyzed and processed by SG.
-Visualisation were produced by SG.
-Original drafts of the manuscript were written by SG and TK.
+The study was conceptualized by SG and TK.
+The data were acquired by TK.
+The data were analyzed and processed by SG.
+Visualization was produced by SG.
+The original drafts of the manuscript were written by SG and TK.
 Drafts were edited and reviewed by SG, TB, DS, BI, and TK.
@@
 All authors have accepted responsibility for the entire content of this
-manuscript and approved its submission.
+manuscript and have approved its submission.
 
@@
 
-Authors state no conflict of interest.
+The authors state no conflict of interest.
 
@@
 Research involving human subjects complied with all relevant national
-regulations, institutional policies and is in accordance with the tenets of
-the Helsinki Declaration (as revised in 2013), and has been
+regulations and institutional policies, as well as the tenets of
+the Helsinki Declaration (as revised in 2013), and was
 approved by the Ethics Committee at the Leipzig University Faculty of
@@
 
-The complete code and reproducible analysis could be found at
+The complete code and reproducible analysis can be found at
 https://ampel-leipzig.github.io/ampel-leipzig-meld/.
-Additionally all data are available in the `ameld` R package [@R-ameld] at
+In addition, all data are available in the `ameld` R package [@R-ameld] at
 https://github.com/ampel-leipzig/ameld
 [![DOI](https://zenodo.org/badge/DOI/10.5281/zenodo.6666018.svg)](https://doi.org/10.5281/zenodo.6666018)
-for further investigations.
+for further investigation.
 

### 2022-07-22 analysis/supplement.Rmd
refactor: apply proofreading suggestions

@@
 
-# Benchmark results of machine learning algorithms
+# Benchmark results of machine-learning algorithms
 
-```{r benchmark-plot, fig.cap = "Benchmark results of machine learning algorithms.", echo = FALSE, fig.width = 18, fig.height = 12}
+```{r benchmark-plot, fig.cap = "Benchmark results of machine-learning algorithms.", echo = FALSE, fig.width = 18, fig.height = 12}
 tar_load(bmrk_results)
@@
     col.names = c("Learner ID", "Harrell C"),
-    caption = paste0("Benchmark results. Ranking of the ", nrow(agg), " tested machine learning algorithms.")
+    caption = paste0("Benchmark results. Ranking of the ", nrow(agg), " tested machine-learning algorithms.")
 )
@@
 
-```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in AUROC.
-Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na and MELD-Plus7 as described in [@blanche2013]."}
+```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, and MELD-Plus7, as described in [@blanche2013]."}
 tar_load(timeROC_MELD)

### 2022-07-23 analysis/supplement.Rmd
fix: typo

@@
 
-```{r survival.plot, fig.cap = "Survival plot. Kaplan-Meier survival estimate for the first 90 days of the analysed population.", fig.width = 8, fig.height = 5}
+```{r survival.plot, fig.cap = "Survival plot. Kaplan-Meier survival estimate for the first 90 days of the analyzed population.", fig.width = 8, fig.height = 5}
 tar_load(zlog_data)

### 2022-08-09 analysis/article.Rmd
refactor: change title

@@
 # https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
-title: "A new model to predict survival in patients with end-stage liver disease"
+title: "A new machine-learning based prediction of survival in patients with end-stage liver disease"
 subtitle: >

### 2022-08-09 analysis/article.Rmd
refactor: integrate Berend's suggestions
refactor: add additional co-author

@@
         email: thomas.berg@medizin.uni-leipzig.de
+        institute: KIMH
+        correspondence: false
+    - Adam Herber:
+        email: adam.herber@medizin.uni-leipzig.de
         institute: KIMH
@@
 AMELD predicts the medium-risk patients more accurately.
-Because of the low number of very ill patients in our cohort, the prediction
-performance in high-risk patients is poor but comparable to the other models.
-This would require recalibration in a larger validation cohort.
 
@@
 The original drafts of the manuscript were written by SG and TK.
-Drafts were edited and reviewed by SG, TB, DS, BI, and TK.
+Drafts were edited and reviewed by SG, TB, AH, DS, BI, and TK.
 The project was supervised by TK.

### 2022-08-09 analysis/supplement.Rmd
refactor: add additional co-author

@@
     - Thomas Berg
+    - Adam Herber
     - Daniel Seehofer

### 2022-08-09 analysis/article.Rmd
fix: add Thorsten's new affiliation

@@
         email: thorsten.kaiser@medizin.uni-leipzig.de
-        institute: ILM
+        institute: [ILM, ILMOWL]
         correspondence: false
@@
     - KAINS: Anesthesiology and Intensive Care Medicine, University Hospital Greifswald, Ferdinand-Sauerbruch-Str., D-17475 Greifswald, Germany.
+    - ILMOWL: Institute of Laboratory Medicine, Microbiology and Clinical Pathobiochemistry, University Hospital OWL, Hospital Lippe, Röntgenstr. 18, D-32756 Detmold, Germany.
 date: "`r Sys.Date()`"

### 2022-08-09 analysis/article.Rmd
refactor: add TB's COI

@@
 
-The authors state no conflict of interest.
+TB recieved grants or research support from
+Abbvie, BMS, Gilead, MSD/Merck, Humedics, Intercept, Merz, Norgine, Novartis, Orphalan, Sequana Medical.
+TB received honoraria or consultation fees/advisory board from
+Abbvie, Alexion, Bayer, Gilead, GSK, Eisai, Enyo Pharma, Falk Foundation, HepaRegeniX GmbH, Humedics,  Intercept, Ipsen, Janssen, MSD/Merck, Novartis, Orphalan, Roche,
+Sequana Medical, SIRTEX, SOBI, and Shionogi
+TB participated in a company sponsored speaker’s bureau for
+Abbvie, Alexion, Bayer, Gilead, Eisai, Intercept, Ipsen, Janssen, MedUpdate GmbH, MSD/Merck, Novartis, Orphalan, Sequana Medica, SIRTEX, and SOBI.
+All other authors state no conflict of interest.
 

### 2022-08-09 analysis/article.Rmd
fix: add dash between machine-learning

@@
 # https://labrtorian.com/2019/08/26/rmarkdown-template-that-manages-academic-affiliations/
-title: "A new machine-learning based prediction of survival in patients with end-stage liver disease"
+title: "A new machine-learning-based prediction of survival in patients with end-stage liver disease"
 subtitle: >
@@
   **Keywords**:
-      end-stage liver disease; IL-6; machine learning; MELD;
+      end-stage liver disease; IL-6; machine-learning; MELD;
       risk stratification; survival estimation

### 2022-08-09 analysis/supplement.Rmd
fix: add dash between machine-learning

--- 2022-07-23 analysis/supplement.Rmd
@@
 ---
-title: "A new model to predict survival in patients with end-stage liver disease"
+title: "A new machine-learning-based prediction of survival in patients with end-stage liver disease"
 subtitle: >

### 2022-08-10 analysis/article.Rmd
refactor: drop DS from author list

@@
         correspondence: false
-    - Daniel Seehofer:
-        email: daniel.seehofer@medizin.uni-leipzig.de
-        institute: KVTTG
-        correspondence: false
     - Berend Isermann:
@@
     - KIMH: Section of Hepatology, Department of Gastroenterology and Rheumatology, University Hospital Leipzig, Liebigstr. 20, D-04103 Leipzig, Germany.
-    - KVTTG: Department of Visceral, Transplantation, Vascular and Thoracic Surgery, University Hospital of Leipzig, Liebigstr. 20, D-04103 Leipzig, Germany.
     - KAINS: Anesthesiology and Intensive Care Medicine, University Hospital Greifswald, Ferdinand-Sauerbruch-Str., D-17475 Greifswald, Germany.
@@
 The original drafts of the manuscript were written by SG and TK.
-Drafts were edited and reviewed by SG, TB, AH, DS, BI, and TK.
+Drafts were edited and reviewed by SG, TB, AH, BI, and TK.
 The project was supervised by TK.

### 2022-08-10 analysis/supplement.Rmd
refactor: drop DS from author list

--- 2022-07-23 analysis/supplement.Rmd
@@
     - Thomas Berg
-    - Daniel Seehofer
+    - Adam Herber
     - Berend Isermann

### 2022-11-27 scripts/Rscript.sh
chore: adapt to new singularity version on brain

@@ -6,7 +6,7 @@ BIND_PATH="/etc/passwd,/etc/group,/var/run"
 MODULESPATH="/usr/share/[Mm]odules/init/bash"
 if [ -f ${MODULESPATH} ] ; then
     . ${MODULESPATH}
-    SINGULARITYVER=3.8.2
+    SINGULARITYVER=3.10.0
     if [ ! -d "/opt/singularity-${SINGULARITYVER}" ] ; then
         SINGULARITYVER=3.4.2
     fi

### 2022-11-27 analysis/article.Rmd
docs: typo

@@
 two different support vector machines [@vanbelle2011; @R-survivalsvm];
-;nd two neural networks [@katzman2018; @kvamme2019; @R-survivalmodels].
+and two neural networks [@katzman2018; @kvamme2019; @R-survivalmodels].
 

### 2022-11-27 analysis/article.Rmd
feat: use mortality estimates from @vanderwerken2021

@@
 `r paste(round(range(tbl_observed_vs_expected_mortality$StandardizedMortalityRatio[-1]), 1), collapse = " to ")`
-times higher than the predicted one.
+times higher than the predicted one [@vanderwerken2021].
 Only in the group with the lowest MELD scores was the mortality much lower.
@@
         "Observed vs. MELD-expected 90-day mortality. ",
-        "MELD mortality values are taken from @wiesner2003. ",
+        "MELD mortality values are taken from @vanderwerken2021 and ",
+        "the mean value was calculated for each MELD category introduced ",
+        "in @wiesner2003.",
         "All patients censored before day 90 are ignored for ",

### 2022-11-27 code/08-tables.R

@@ -185,14 +185,34 @@ tg_tables <- list(
     packages = c("ameld", "gtsummary", "survival"),
     deployment = "main"
     ),
-    tar_target(tbl_observed_vs_expected_mortality,
+    tar_target(tbl_observed_vs_expected_mortality, {
+
+        ## MELD mortality values are taken from @vanderwerken2021
+        vanderwerken <- data.frame(
+            MELD = 6:40,
+            mortality = 1 -
+            c(0.9700, 0.9720, 0.9715, 0.9744, 0.9802, 0.9759, 0.9747, 0.9791,
+              0.9752, 0.9651, 0.9606, 0.9514, 0.9380, 0.9309, 0.9128, 0.9048,
+              0.8721, 0.8509, 0.8098, 0.7869, 0.7406, 0.6787, 0.6619, 0.5791,
+              0.5244, 0.4906, 0.4831, 0.4069, 0.3860, 0.3629, 0.2944, 0.2429,
+              0.2469, 0.1003, 0.1562)
+        )
+        vanderwerken$MeldCategory <- cut(
+            vanderwerken$MELD,
+            breaks = c(-Inf, seq(10, 40, by=10), Inf),
+            labels = c("[6,9]", "[10,20)", "[20,30)", "[30,40)", "[40,52)"),
+            right = FALSE
+        )
+        expected <- tapply(
+            vanderwerken$mortality, vanderwerken$MeldCategory, mean
+        )
+
         observed_vs_expected_mortality(
             Surv(labelled_meld_data$DaysAtRisk, labelled_meld_data$Deceased),
             f = labelled_meld_data$MeldCategory,
-            time = 90,
-            ## MELD mortality values are taken from @wiesner2003.
-            c(0.019, 0.06, 0.196, 0.526, 0.713)
-        ),
+            time = 90, expected
+        )
+    },
     packages = "ameld",
     deployment = "main"
     )

### 2022-11-27 analysis/supplement.Rmd
refactor: change supplemental figure 2.1 according to reviewer's suggestions

@@
 r$filter(task_id = "zlog_eldd", learner_id = id)
-autoplot(r) +
-    geom_boxplot(aes(fill = learner_id)) +
+
+relabel_learners <- function(x) {
+    x <- gsub("\\.tuned", "", x)
+    x <- gsub("1$", "", x)
+    x <- gsub("^surv\\.", "", x)
+    x
+}
+
+## taken from mlr3viz::autoplot.BenchmarkResults
+task <- r$tasks$task[[1L]]
+measure <- mlr3::assert_measure(mlr3::as_measure(msr("surv.cindex"), task_type = task$task_type), task = task)
+measure_id <- measure$id
+tab <- fortify(r, measure = measure)
+tab$nr <- sprintf("%09d", tab$nr)
+tab$learner_id <- relabel_learners(tab$learner_id)
+tab$group <- tab$learner_id
+tab$group[tab$group == "xgboost"] <- "boosting"
+tab$group[grepl("lasso|ridge|net", tab$group)] <- "penalized regression"
+tab$group[grepl("svm", tab$group)] <- "svm"
+tab$group[tab$group %in% c("ranger", "rfsrc")] <- "random forests"
+tab$group[tab$group %in% c("coxtime", "deepsurv")] <- "neural networks"
+tab$group <- factor(
+    tab$group,
+    levels = c(
+        "cox", "penalized regression",
+        "random forests", "boosting", "svm", "neural networks"
+    )
+)
+learner_label_map <- tab[!duplicated(tab$nr), c("nr", "learner_id")]
+learner_labels <- learner_label_map$learner_id
+names(learner_labels) <- learner_label_map$nr
+ggplot(tab, mapping = aes(x = .data$nr, y = .data[[measure_id]])) +
+    geom_boxplot(aes(fill = .data$group)) +
+    labs(fill = "Algorithm category") +
+    labs(y = "Harrell's concordance index") +
+    labs(x = "Model") +
+    scale_x_discrete(labels = learner_labels) +
+    scale_fill_viridis(labels = learner_labels, discrete = TRUE, option = "cividis") +
+    facet_grid(cols = vars(group), scales = "free_x", space = "free") +
+    mlr3viz:::apply_theme(list(theme_mlr3())) +
     geom_jitter(position = position_jitter(0.2)) +
-    scale_fill_viridis(discrete = TRUE) +
-    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5))
+    theme(
+        legend.position = "none",
+        strip.text.x = element_text(size = 20),
+        text = element_text(size = 24),
+        axis.text.x = element_text(angle = 45, hjust = 1)
+    )
 ```
@@
 setorder(agg, -harrell)
+agg$name <- relabel_learners(agg$learner_id)
 knitr::kable(
-    agg,
-    col.names = c("Learner ID", "Harrell C"),
+    agg[, list(name, learner_id, harrell)],
+    col.names = c("Model", "Learner ID", "Harrell C"),
     caption = paste0("Benchmark results. Ranking of the ", nrow(agg), " tested machine-learning algorithms.")

### 2022-11-27 analysis/elasticnet.Rmd
feat: add MELD 3.0 score

@@
 ps <- lapply(
-    zlog_data[paste0("SurvProbMeld", c("Unos", "NaUnos", "Plus7"))],
+    zlog_data[paste0("SurvProbMeld", c("Unos", "NaUnos", "3", "Plus7"))],
     function(p) {
@@
 )
-names(ps) <- c("MELD", "MELD-Na", "MELD-Plus7")
-col <- viridisLite::viridis(6)[4:6]
+names(ps) <- c("MELD", "MELD-Na", "MELD 3.0", "MELD-Plus7")
+col <- viridisLite::viridis(7)[4:7]
 
@@
 ps <- lapply(
-    zlog_data_complete_cases[paste0("SurvProbMeld", c("Unos", "NaUnos", "Plus7"))],
+    zlog_data_complete_cases[paste0("SurvProbMeld", c("Unos", "NaUnos", "3", "Plus7"))],
     function(p) {

### 2022-11-27 analysis/histograms.Rmd
feat: add MELD 3.0 score

@@
 meldv <-
-    c(paste0("ScoreMeld", c("", "Unos", "Na", "NaUnos")),
-      paste0("SurvProbMeld", c("Unos", "NaUnos", "Plus7")))
+    c(paste0("ScoreMeld", c("", "Unos", "Na", "NaUnos", "3")),
+      paste0("SurvProbMeld", c("Unos", "NaUnos", "3", "Plus7")))
 for (v in meldv) {

### 2022-11-27 analysis/supplement.Rmd
feat: add MELD 3.0 score

@@
 tar_load(timeROC_MELDNa)
+tar_load(timeROC_MELD3)
 tar_load(timeROC_MELDPlus7)
@@
     "MELD-Na" = timeROC_MELDNa,
+    "MELD 3.0" = timeROC_MELD3,
     "MELD-Plus7" = timeROC_MELDPlus7
 )
-plot_surv_roc_trend(m, xlab = "time [days]", col = viridisLite::viridis(5)[-2])
+plot_surv_roc_trend(m, xlab = "time [days]", col = viridisLite::viridis(6)[-2])
 ```

### 2022-11-27 code/01-data.R

@@ -29,8 +29,17 @@
         dialysis = x$Dialysis,
         type = "unos"
     )
+    x$ScoreMeld3 <- meld3(
+        creatinine = as_metric(x$CRE_S, "creatinine"),
+        bilirubin = as_metric(x$BILI_S, "bilirubin"),
+        inr = x$INR_C,
+        sodium = x$NA_S,
+        albumin = x$ALB_S / 10,
+        female = x$Sex == "female"
+    )
     x$SurvProbMeldUnos <- pmeld(x$ScoreMeldUnos)
     x$SurvProbMeldNaUnos <- pmeld(x$ScoreMeldNaUnos)
+    x$SurvProbMeld3 <- pmeld3(x$ScoreMeld3)
     x$SurvProbMeldPlus7 <- pmeld_plus7(
         creatinine = as_metric(x$CRE_S, "creatinine"),
         bilirubin = as_metric(x$BILI_S, "bilirubin"),
@@ -161,6 +170,8 @@ tg_data <- list(
                 attr(x$ScoreMeldUnos, "label") <- "MELD score"
             attr(x$ScoreMeldNa, "label") <-
                 attr(x$ScoreMeldNaUnos, "label") <- "MELD-Na score"
+            attr(x$ScoreMeld3, "label") <-
+                attr(x$ScoreMeld3, "label") <- "MELD 3.0 score"
             attr(x$MeldCategory, "label") <- "MELD Category"
 
             ## labels where the column names are identical to the abbreviation

### 2022-11-27 code/06-timeroc.R

@@ -1,5 +1,5 @@
-cols <- c("SurvProbMeldUnos", "SurvProbMeldNaUnos", "SurvProbMeldPlus7")
-nms <- c("MELD", "MELDNa", "MELDPlus7")
+cols <- c("SurvProbMeldUnos", "SurvProbMeldNaUnos", "SurvProbMeldPlus7", "SurvProbMeld3")
+nms <- c("MELD", "MELDNa", "MELDPlus7", "MELD3")
 
 smbs <- rlang::syms(paste0("timeROC_", nms))
 

### 2022-11-27 code/08-tables.R

@@ -31,6 +31,7 @@ tg_tables <- list(
                ScoreMeldUnos,
                MeldCategory,
                ScoreMeldNaUnos,
+               ScoreMeld3,
             ) |>
             tbl_summary(
                 by = Sex,

### 2022-11-30 analysis/article.Rmd
feat: add MELD 3.0 and adapt to reviewer's suggestions

@@
 subtitle: >
-    Comparision against MELD, MELD-Na, MELD-Plus7.
+    Comparision against MELD, MELD-Na, MELD 3.0, MELD-Plus7.
 abstract: >
@@
       The new model offers improved discrimination and calibration over MELD
-      and MELD with sodium (MELD-Na) or the MELD-Plus7 risk score.
+      and MELD with sodium (MELD-Na), MELD 3.0, or the MELD-Plus7 risk score.
 
@@
       We provide a new machine-learning-based model of end-stage liver disease
-      that outperforms the classical MELD score for 90-day survival prediction.
+      that incorporates synthesis and inflammatory markers and
+      could improve the classical MELD score for 90-day survival prediction.
 
@@
 tar_load(timeROC_MELDPlus7)
+tar_load(timeROC_MELD3)
 tar_load(timeROC_RCV)
@@
 risk factor in liver cirrhosis [@kim2008; @optn_policies2021].
+Recently MELD 3.0 was introduced adding a factor for female sex, albumin and
+interactions between bilirubin and sodium and between albumin and creatinine.
 The MELD was initially developed to predict the survival of patients
@@
 [@trotter2004; @cholongitas2007].
-In addition women are disadvantaged [@sealock2022].
+In addition women are disadvantaged in MELD-Na [@sealock2022].
 Patients with identical disease states can have very different MELD scores and
@@
 
-As shown by the MELD-Plus risk scores, better predictive scores often need more
-variables and are more complicated. To reduce the risk of overlooking or
+As shown by the MELD-Plus risk and MELD 3.0 scores, better predictive scores
+often need more variables and are more complicated.
+To reduce the risk of overlooking or
 incorrectly calculating and interpreting the results,
@@
 This study aims to find clinical and laboratory values that improve on the risk
-stratification for liver transplantation of the classical MELD, MELD-Na, and
-MELD-Plus scores and that might be implemented as part of the
-AMPEL clinical decision support system.
+stratification for liver transplantation of the classical
+MELD, MELD-Na, MELD 3.0 and MELD-Plus scores and that might be implemented as
+part of the AMPEL clinical decision support system.
 
@@
 
+MELD 3.0 was calculated as described in kim2022:
+
+$$
+\begin{aligned}
+MELD3 = 1.33 (if female) &+
+        4.56 &* \ln(bilirubin [mg/dL]) +  \\
+        0.82 &* (137 - Na [mmol/L]) - 0.24 * (137 - Na [mmol/L]) * \ln(bilirubin [mg/dL]) + \\
+        9.09 &* \ln(INR) + \\
+        11.14 &* \ln(creatinine [mg/dL]) + 1.85 * (3.5 - albumin [g/dL]) - 1.83 * (3.5 - albumin [g/dL]) * \ln(creatinine [mg/dL]) + 6
+\end{aligned}
+$$
+In contrast to MELD-Na the maximum allowed creatinine was 3.0 mg/dL.
+
 The MELD-Plus7 risk score was calculated as described in @kartoun2017:
@@
     6.47834101 &* \log_{10}(1 + sodium [mmol/L]) - \\
-    6.34990436 &* \log_{10}(1 + albumin [g/L]) + \\
+    6.34990436 &* \log_{10}(1 + albumin [g/dL]) + \\
     1.92811726 &* \log_{10}(1 + wbc [th/cumm]) + \\
@@
 \end{aligned}
-$$
+$
 
@@
 
-## Data processing and machine learning
+## Data processing and machine-learning
 
@@
 
-The final model was compared against the MELD, MELD-Na, and MELD-Plus7 scores
-using the area under the time-dependent receiver operating characteristic curve
-(AUROC) based on the nonparametric inverse probability of censoring weighting
+The final model was compared against the MELD, MELD-Na, MELD 3.0 and MELD-Plus7
+scores using the area under the time-dependent receiver operating
+characteristic curve (AUROC) based on the
+nonparametric inverse probability of censoring weighting
 estimator (IPCW) [@blanche2013, @R-timeROC].
@@
 (`r .tbl1("CLI")`).
-At evaluation time, the median (IQR) MELD and MELD-Na scores were
-`r .tbl1("ScoreMeldUnos")` and
-`r .tbl1("ScoreMeldNaUnos")`, respectively.
+At evaluation time, the median (IQR) MELD, MELD-Na and MELD 3.0 scores were
+`r .tbl1("ScoreMeldUnos")`,
+`r .tbl1("ScoreMeldNaUnos")`, and
+`r .tbl1("ScoreMeld3")`, respectively.
 Within 90 days, `r .tbl1("M90")` patients had died.
@@
 
-According to @wiesner2003, we divided our cohort into
-`r nlevels(labelled_meld_data$MeldCategory)` MELD-score risk
+We divided our cohort into
+`r nlevels(labelled_meld_data$MeldCategory)` MELD-Na-score risk
 categories, with `r enum(table(labelled_meld_data$MeldCategory))` patients,
@@
 times higher than the predicted one [@vanderwerken2021].
-Only in the group with the lowest MELD scores was the mortality much lower.
+Only in the group with the lowest MELD-Na scores was the mortality much lower.
 
@@
         c(
-            "MELD category",
+            "MELD-Na category",
             "Observed deaths (n)",
@@
     caption = paste0(
-        "Observed vs. MELD-expected 90-day mortality. ",
-        "MELD mortality values are taken from @vanderwerken2021 and ",
-        "the mean value was calculated for each MELD category introduced ",
-        "in @wiesner2003.",
+        "Observed vs. MELD-Na-expected 90-day mortality. ",
+        "MELD-Na mortality values are taken from @vanderwerken2021 and ",
+        "the mean value was calculated for each MELD-Na category.",
         "All patients censored before day 90 are ignored for ",
-        " the calculation of the MELD-expected deaths. ",
+        "the calculation of the MELD-Na-expected deaths. ",
         "SMR, Standardized mortality ratio = observed deaths/expected deaths."
@@
 ```{r roc, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE, fig.cap = "Receiver operating characteristic (ROC) curve.
-Area under the time-dependent ROC curve (AUC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, and MELD-Plus7, as described in [@blanche2013]."}
-col <- viridisLite::viridis(5)
+Area under the time-dependent ROC curve (AUC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0, MELD-Plus7, as described in [@blanche2013]."}
+col <- viridisLite::viridis(6)
 m <- list(
@@
     "MELD-Na" = timeROC_MELDNa,
+    "MELD 3.0" = timeROC_MELD3,
     "MELD-Plus7" = timeROC_MELDPlus7
@@
 invisible(roc <- plot_surv_roc(m, timepoint = 90, col = col[-2]))
-roc <- roc[order(-roc)]
+o <- order(-roc)
+ci <- attr(roc, "CI")[o]
+roc <- roc[o]
+
+pAUC <- function(auc, ci) {
+    sprintf("AUROC %0.3f %s", auc, ci)
+}
 ```
@@
 
-AMELD offers the best AUROC (`r round(roc["AMELD"], 3)`) of all competitors.
+AMELD offers the best `r pAUC(roc["AMELD"], ci["AMELD"])` of all competitors.
 However, after correction for multiple testing, the difference against
 the second-best model, `r names(roc)[2]`
-(AUROC: `r round(roc[2], 3)`), is not statistical significant
+(`r pAUC(roc[2], ci[2])`), is not statistical significant
 (`r paste0("$p = ", round(tp[2, "t=90"], 2), "$")`)
@@
 
-```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = paste('Calibration of AMELD in comparison to MELD, MELD-Na, and MELD-Plus7.\n The points mark the mean predicted survival of', ameldcfg$m, 'patients per interval against their observed mean survival.
+```{r bootrcv, echo = FALSE, fig.width = 8, fig.height = 6, fig.cap = paste('Calibration of AMELD in comparison to MELD, MELD-Na, MELD 3.0 and MELD-Plus7.\n The points mark the mean predicted survival of', ameldcfg$m, 'patients per interval against their observed mean survival.
 The error bars correspond to the 95% confidence interval of the survival estimate.'), echo = FALSE}
@@
 )
-names(ps) <- c("MELD", "MELD-Na", "MELD-Plus7")
+names(ps) <- c("MELD", "MELD-Na", "MELD 3.0", "MELD-Plus7")
 
@@
 University Hospital Leipzig, we developed a new model to predict
-90-day survival probability. Our new model, AMELD, outperforms the MELD-Na
-score and the recently published MELD-Plus7 risk score.
+90-day survival probability. Our new model, AMELD,
+incorporates synthesis and inflammatory markers and
+could improve the prediction over previous published
+MELD scores.
 
@@
 [@sealock2022].
-To address sexual differences, in laboratory measurements we applied the *zlog*
+MELD 3.0 tries to circumvent this by adding 1.33 to the score for femal sex.
+To address sexual differences in laboratory measurements we applied the *zlog*
 transformation [@hoffmann2017] in our study.
@@
 [@thabut2007].
-However, neither MELD nor MELD-Na took inflammation into account.
+However, neither MELD, MELD-Na nor MELD 3.0 took inflammation into account.
 By contrast, IL-6 is the most important variable in our model. Even if IL-6 is
@@
 
-With a median MELD score of `r .tbl1("ScoreMeldUnos", pattern = "{median}")`,
+With a median MELD-Na score of `r .tbl1("ScoreMeldNaUnos", pattern = "{median}")`,
 our cohort was healthier than those in previous studies that reported
-a score of 14-15 [@kim2008; @kartoun2017].
+a score of 14-18.5 [@kim2008; @kartoun2017; @vanderwerken2021].
 However, the 90-day mortality rate of `r .tbl1("M90")` was comparable to
@@
 Using these six laboratory measurements, AMELD
-outperforms the established MELD-Na and the recently published
-MELD-Plus7 risk scores.
+could improve the prediction over previous published
+MELD scores.
 For wider adoption, a prospective multi-center study for model

### 2022-11-30 analysis/supplement.Rmd
feat: add MELD 3.0 and adapt to reviewer's suggestions

@@
 knitr::kable(
-    agg[, list(name, learner_id, harrell)],
-    col.names = c("Model", "Learner ID", "Harrell C"),
-    caption = paste0("Benchmark results. Ranking of the ", nrow(agg), " tested machine-learning algorithms.")
+    agg[, list(name, harrell)],
+    col.names = c("Model", "Harrell's concordance index"),
+    caption = paste0("Benchmark results. Ranking of the ", nrow(agg), " tested machine-learning algorithms."),
+    digits = 3
+)
+```
+
+# Observed versus MELD-Na-expected 90-day mortality
+
+```{r tbl-observed-vs-expected-mortality5, results = "asis", echo = FALSE, message = FALSE}
+## identical to table in article.Rmd but categories spanning just 5 points
+tar_load(labelled_meld_data)
+
+## MELD mortality values are taken from @vanderwerken2021
+vanderwerken <- data.frame(
+    MELD = 6:40,
+    mortality = 1 -
+    c(0.9700, 0.9720, 0.9715, 0.9744, 0.9802, 0.9759, 0.9747, 0.9791,
+      0.9752, 0.9651, 0.9606, 0.9514, 0.9380, 0.9309, 0.9128, 0.9048,
+      0.8721, 0.8509, 0.8098, 0.7869, 0.7406, 0.6787, 0.6619, 0.5791,
+      0.5244, 0.4906, 0.4831, 0.4069, 0.3860, 0.3629, 0.2944, 0.2429,
+      0.2469, 0.1003, 0.1562)
+)
+
+breaks <- c(-Inf, seq(10, 40, by=5), Inf)
+labels <- c(
+    paste0(
+        "[",
+        floor(min(labelled_meld_data$ScoreMeldNaUnos, na.rm = TRUE)),
+        ",9]"
+    ),
+    paste0("[", seq(10, 35, by=5), ",", seq(15, 40, by=5), ")"),
+    paste0(
+        "[40,",
+        ceiling(max(labelled_meld_data$ScoreMeldNaUnos, na.rm = TRUE)),
+        ")"
+    )
+)
+
+vanderwerken$MeldCategory <- cut(
+    vanderwerken$MELD,
+    breaks = breaks,
+    labels = labels,
+    right = FALSE
+)
+expected <- tapply(
+    vanderwerken$mortality, vanderwerken$MeldCategory, mean
+)
+
+tbl_observed_vs_expected_mortality5 <- observed_vs_expected_mortality(
+    Surv(labelled_meld_data$DaysAtRisk, labelled_meld_data$Deceased),
+    f = cut(
+        labelled_meld_data$ScoreMeldNaUnos,
+        breaks = breaks, labels = labels,
+        right = FALSE
+    ),
+    time = 90, expected
+)
+
+tbl_observed_vs_expected_mortality5$ObservedMortality <-
+    tbl_observed_vs_expected_mortality5$ObservedMortality * 100
+tbl_observed_vs_expected_mortality5$ExpectedMortality <-
+    tbl_observed_vs_expected_mortality5$ExpectedMortality * 100
+knitr::kable(
+    cbind(
+        row.names(tbl_observed_vs_expected_mortality5),
+        tbl_observed_vs_expected_mortality5
+    ),
+    row.names = FALSE,
+    col.names =
+        c(
+            "MELD category",
+            "Observed deaths (n)",
+            "Expected deaths (n)",
+            "Standardized mortality ratio (SMR)",
+            "Observed mortality (%)",
+            "Expected mortality (%)"
+        ),
+    caption = paste0(
+        "Observed vs. MELD-Na-expected 90-day mortality. ",
+        "MELD-Na mortality values are taken from @vanderwerken2021 and ",
+        "the mean value was calculated for each MELD-Na category.",
+        "All patients censored before day 90 are ignored for ",
+        "the calculation of the MELD-Na-expected deaths. ",
+        "SMR, Standardized mortality ratio = observed deaths/expected deaths."
+    ),
+    digits = 1
 )
@@
 
-```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, and MELD-Plus7, as described in [@blanche2013]."}
+```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0 and MELD-Plus7, as described in [@blanche2013]."}
 tar_load(timeROC_MELD)
@@
 )
+plot_surv_roc_trend(m, xlab = "time [days]", col = viridisLite::viridis(6)[-2])
+```
+
+```{r roc-plot-wo-cb, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0 and MELD-Plus7, as described in [@blanche2013].
+Identical to the figure above but without confidence bands."}
+m <- lapply(m, function(x) { x$confint <- NULL;x })
 plot_surv_roc_trend(m, xlab = "time [days]", col = viridisLite::viridis(6)[-2])

### 2022-11-30 code/01-data.R

@@ -50,15 +50,15 @@
         age = x$Age
     )
     x$MeldCategory <- cut(
-        x$ScoreMeldUnos,
+        x$ScoreMeldNaUnos,
         breaks = c(-Inf, seq(10, 40, by=10), Inf),
         labels = c(
             paste0(
-                "[", floor(min(x$ScoreMeldUnos, na.rm = TRUE)), ",9]"
+                "[", floor(min(x$ScoreMeldNaUnos, na.rm = TRUE)), ",9]"
             ),
             "[10,20)", "[20,30)", "[30,40)",
             paste0(
-                "[40,", ceiling(max(x$ScoreMeldUnos, na.rm = TRUE)), ")"
+                "[40,", ceiling(max(x$ScoreMeldNaUnos, na.rm = TRUE)), ")"
             )
         ),
         right = FALSE

### 2022-11-30 code/08-tables.R

@@ -187,7 +187,6 @@ tg_tables <- list(
     deployment = "main"
     ),
     tar_target(tbl_observed_vs_expected_mortality, {
-
         ## MELD mortality values are taken from @vanderwerken2021
         vanderwerken <- data.frame(
             MELD = 6:40,

### 2022-12-01 analysis/article.Rmd
fix: typos

@@
       that incorporates synthesis and inflammatory markers and
-      could improve the classical MELD score for 90-day survival prediction.
+      may improve the classical MELD score for 90-day survival prediction.
 
@@
 Recently MELD 3.0 was introduced adding a factor for female sex, albumin and
-interactions between bilirubin and sodium and between albumin and creatinine.
+interactions between bilirubin and sodium and between albumin and creatinine
+[@kim2021].
 The MELD was initially developed to predict the survival of patients
@@
 stratification for liver transplantation of the classical
-MELD, MELD-Na, MELD 3.0 and MELD-Plus scores and that might be implemented as
+MELD, MELD-Na, MELD 3.0, and MELD-Plus scores and that might be implemented as
 part of the AMPEL clinical decision support system.
@@
 
-MELD 3.0 was calculated as described in kim2022:
+MELD 3.0 was calculated as described in kim2021:
 
@@
 $$
-In contrast to MELD-Na the maximum allowed creatinine was 3.0 mg/dL.
+In contrast to MELD-Na the maximum allowed creatinine was set to 3.0 mg/dL.
 
@@
         "MELD-Na mortality values are taken from @vanderwerken2021 and ",
-        "the mean value was calculated for each MELD-Na category.",
+        "the mean value was calculated for each MELD-Na category. ",
         "All patients censored before day 90 are ignored for ",
@@
 incorporates synthesis and inflammatory markers and
-could improve the prediction over previous published
+may improve the prediction over previous published
 MELD scores.
@@
 [@thabut2007].
-However, neither MELD, MELD-Na nor MELD 3.0 took inflammation into account.
+However, neither MELD, MELD-Na, nor MELD 3.0 took inflammation into account.
 By contrast, IL-6 is the most important variable in our model. Even if IL-6 is
@@
 Using these six laboratory measurements, AMELD
-could improve the prediction over previous published
+may improve the prediction over previous published
 MELD scores.

### 2022-12-01 analysis/supplement.Rmd
fix: typos

@@
 
-```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0 and MELD-Plus7, as described in [@blanche2013]."}
+```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0, and MELD-Plus7, as described in [@blanche2013]."}
 tar_load(timeROC_MELD)
@@
 
-```{r roc-plot-wo-cb, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0 and MELD-Plus7, as described in [@blanche2013].
+```{r roc-plot-wo-cb, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0, and MELD-Plus7, as described in [@blanche2013].
 Identical to the figure above but without confidence bands."}

### 2022-12-01 code/08-tables.R

@@ -29,8 +29,8 @@ tg_tables <- list(
             select(
                Sex,
                ScoreMeldUnos,
-               MeldCategory,
                ScoreMeldNaUnos,
+               MeldCategory,
                ScoreMeld3,
             ) |>
             tbl_summary(

### 2022-12-01 analysis/article.Rmd
fix: missing linebreak

@@
 \begin{aligned}
-MELD3 = 1.33 (if female) &+
+MELD3 = 1.33 (if female) &+ \\
         4.56 &* \ln(bilirubin [mg/dL]) +  \\
@@
         9.09 &* \ln(INR) + \\
-        11.14 &* \ln(creatinine [mg/dL]) + 1.85 * (3.5 - albumin [g/dL]) - 1.83 * (3.5 - albumin [g/dL]) * \ln(creatinine [mg/dL]) + 6
+        11.14 &* \ln(creatinine [mg/dL]) + \\
+        1.85 &* (3.5 - albumin [g/dL]) - 1.83 * (3.5 - albumin [g/dL]) * \ln(creatinine [mg/dL]) + 6
 \end{aligned}

### 2022-12-05 analysis/supplement.Rmd
feat: add confidence bands to roc plots

@@
 
-# AUROC trend
+# ROC curves
 
-```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0, and MELD-Plus7, as described in [@blanche2013]."}
+```{r roc-cb-all, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE, fig.cap = "Receiver operating characteristic (ROC) curve.
+Area under the time-dependent ROC curve (AUC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0, MELD-Plus7, as described in [@blanche2013].
+The dashed lines depict the corresponding 95% confidence bands calculated by threshold averaging as described in [@fawcett2004]."}
 tar_load(timeROC_MELD)
@@
 tar_load(timeROC_RCV)
-
+col <- viridisLite::viridis(6)[-2]
 m <- list(
@@
 )
+invisible(roc <- plot_surv_roc(m, timepoint = 90, col = col, legend = FALSE))
+o <- order(-roc)
+ci <- attr(roc, "CI")[o]
+roc <- roc[o]
+
+rjlegend(
+    legend = c(
+        sprintf("AUC %s: %0.3f %s", names(roc), roc, ci),
+        "95% confidence bands"
+    ),
+    col = c(col[o], "#808080"),
+    lty = c(rep(1, length(m)), 2)
+)
+
+tar_load(confidencebands_MELD)
+tar_load(confidencebands_MELDNa)
+tar_load(confidencebands_MELD3)
+tar_load(confidencebands_MELDPlus7)
+tar_load(confidencebands_RCV)
+cb <- list(
+    AMELD = confidencebands_RCV,
+    MELD = confidencebands_MELD,
+    "MELD-Na" = confidencebands_MELDNa,
+    "MELD 3.0" = confidencebands_MELD3,
+    "MELD-Plus7" = confidencebands_MELDPlus7
+)
+
+for (i in seq_along(cb)) {
+    lines(cb[[i]]$FP["2.5%",], cb[[i]]$TP["97.5%",], col = col[o][i], lty = 2)
+    lines(cb[[i]]$FP["97.5%",], cb[[i]]$TP["2.5%",], col = col[o][i], lty = 2)
+}
+```
+
+```{r roc-cb-ameld, fig.width = 7, fig.height = 7, echo = FALSE, message = FALSE, fig.cap = "Receiver operating characteristic (ROC) curve.
+Area under the time-dependent ROC curve (AUC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0, MELD-Plus7, as described in [@blanche2013].
+The dashed lines depict the corresponding 95% confidence bands for AMELD calculated by threshold averaging as described in [@fawcett2004].
+The other confidence bands are hidden for easier readability."}
+invisible(plot_surv_roc(m, timepoint = 90, col = col, legend = FALSE))
+j <- which(m[["AMELD"]]$times == 90)
+#lines(m[["AMELD"]]$FP[, j], m[["AMELD"]]$TP[, j], col = col[1], lty = 1, lwd = 2)
+rjlegend(
+    legend = c(
+        sprintf("AUC %s: %0.3f %s", names(roc), roc, ci),
+        "95% confidence bands"
+    ),
+    col = c(col[o], "#808080"),
+    lty = c(rep(1, length(m)), 2)
+)
+lines(cb[["AMELD"]]$FP["2.5%",], cb[["AMELD"]]$TP["97.5%",], col = col[1], lty = 2)
+lines(cb[["AMELD"]]$FP["97.5%",], cb[["AMELD"]]$TP["2.5%",], col = col[1], lty = 2)
+polygon(
+    c(cb[["AMELD"]]$FP["2.5%",], rev(cb[["AMELD"]]$FP["97.5%",])),
+    c(cb[["AMELD"]]$TP["97.5%",], rev(cb[["AMELD"]]$TP["2.5%",])),
+    col = sub("FF$", "11", col[1]), lty = 0
+)
+```
+
+# AUROC trend
+
+```{r roc-plot, echo = FALSE, fig.width = 18, fig.height = 12, fig.cap = "Trend in the area under the time-dependent receiver operating characteristic curve (AUROC) based on the nonparametric inverse probability of censoring weighting estimate (IPCW) for AMELD, MELD, MELD-Na, MELD 3.0, and MELD-Plus7, as described in [@blanche2013]."}
 plot_surv_roc_trend(m, xlab = "time [days]", col = viridisLite::viridis(6)[-2])

### 2022-12-05 _targets.R

@@ -76,6 +76,8 @@ list(
    tg_ameld,
    tg_timeroc,
    tg_timeroc_psurv,
+   tg_confidencebands,
+   tg_confidencebands_psurv,
    tg_variable_importance,
    tg_reports,
    tg_site

### 2022-12-05 code/06-timeroc.R

@@ -24,6 +24,25 @@ tg_timeroc <- tar_eval(
     )
 )
 
+smbs <- rlang::syms(paste0("confidencebands_", nms))
+
+tg_confidencebands <- tar_eval(
+    values = list(s = smbs, col = cols),
+    tar_target(
+        s,
+        confidencebands(
+            T = zlog_data$DaysAtRisk,
+            delta = zlog_data$Deceased,
+            marker = 1 - zlog_data[[col]],
+            cause = 1,
+            timepoint = 90,
+            nboot = 200
+        ),
+        packages = c("survival", "timeROC", "ameld"),
+        deployment = "main"
+    )
+)
+
 tg_timeroc_psurv <- list(
     tar_target(timeROC_RCV, {
         tr <- timeROC::timeROC(
@@ -83,6 +102,21 @@ tg_timeroc_psurv <- list(
         tr
     },
     packages = c("survival", "timeROC"),
+    deployment = "main"
+    )
+)
+
+tg_confidencebands_psurv <- list(
+    tar_target(confidencebands_RCV,
+    confidencebands(
+            T = zlog_data$DaysAtRisk,
+            delta = zlog_data$Deceased,
+            marker = 1 - psurvbootrcv,
+            cause = 1,
+            timepoint = 90,
+            nboot = 200
+    ),
+    packages = c("survival", "timeROC", "ameld"),
     deployment = "main"
     )
 )

### 2022-12-19 analysis/article.Rmd
feat: add some paragraphs for variable importance

@@
 
+Further we determined the variable importance by counting how often the
+variable was chosen during the feature selection in the bootstrap process
+described above and by the logrank value returned by one random forest
+implementation [@R-ranger].
+
 Summary tables were built using the *gtsummary* package
@@
 
+## Variable importance
+
+```{r selvar, echo = FALSE}
+bsrank <- rank(-ameld:::.selvars(bootrcv$models, bootrcv$s), ties = "first")
+
+bsrankSodium <- as.vector(bsrank["NA_S"])
+
+bsranktxtSodium <-
+    if (!is.na(bsrankSodium)) {
+        paste0(
+            "in bootstrap selection rank ", bsrankSodium,
+            " of ", ncol(amelddata$x)
+        )
+    } else {
+        paste0("it was never selected in any boostrap sample")
+    }
+
+rfrank <- rank(-rngr$variable.importance, ties = "first")
+rfrankSodium <- as.vector(rfrank["NA_S"])
+rfranktxtSodium <- paste0(
+    "in a random forest variable importance rank ", rfrankSodium, " of ",
+    ncol(amelddata$x)
+)
+```
+
+The AMELD selected features
+`r enum(gsub("^([A-Z])([a-z])", "\\L\\1\\2", cfnms, perl = TRUE)) `
+are the `r max(bsrank[names(cfnms)])` highest ranked features in the bootstrap
+process. In the random forest logrank they are amoung the
+`r max(rfrank[names(cfnms)])` highest ranked features out of `r length(rfrank)`
+features.
+Sodium and albumin introduced by MELD-Na and MELD 3.0 are on position
+`r paste(bsrank[c("NA_S", "ALB_S")], collapse = " and ")` for the bootstrap and
+`r paste(rfrank[c("NA_S", "ALB_S")], collapse = " and ")` for the random forest
+variable importance ranking, respectively
+(see \@ref(fig:vimpbs), \@ref(fig:vimprf) for the complete rankings).
+
 # Discussion
@@
 
-```{r selvar, echo = FALSE}
-bsrank <- as.vector(rank(-ameld:::.selvars(bootrcv$models, bootrcv$s), ties = "first")["NA_S"])
-
-bsranktxt <-
-    if (!is.na(bsrank)) {
-        paste0("in bootstrap selection rank ", bsrank, " of ", ncol(amelddata$x))
-    } else {
-        paste0("it was never selected in any boostrap sample")
-    }
-
-rfrank <- as.vector(rank(-rngr$variable.importance, ties = "first")["NA_S"])
-rfranktxt <-
-    paste0("in a random forest variable importance rank ", rfrank, " of ", ncol(amelddata$x))
-```
-
 Although the Organ Procurement and Transplantation Network recommends the use
@@
 random forest model does the sodium gain any important weight
-(`r bsranktxt` and `r rfranktxt`, see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
+(`r bsranktxtSodium` and `r rfranktxtSodium`, see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
 In addition, sodium levels can be affected by diuretic therapy, and
@@
 Consequently it is not a component of our model.
+
+The albumin included in MELD 3.0 is part of the total protein that has
+throughout higher rankings in both variable importance measurements
+(`r bsrank["ALB_S"]` vs. `r bsrank["PROT_S"]` in the bootstrap selection
+and `r rfrank["ALB_S"]` vs. `r rfrank["PROT_S"]` in the random forest variable
+importance, see \@ref(fig:vimpbs), \@ref(fig:vimprf)).
+Using total protein instead of albumin as synthesis marker
+offers more information about additional globulins and especially
+coagulation factors that are incompletely measured by INR.
 
